### 2020-06-04 outline.Rmd
adding outline of paper

---
title: "Outline"
output: github_document
---

# Introduction
  - MEME Suite is an excellent collection of tools for motif analysis
  - MEME Suite Webserver serves thousands of users/year
  - Problem: Manual uploads & configuration of each job makes iteration difficult
  - Problem: occasionally long queue times
  - Problem: Although data are returned in machine-readable format (.xml), there are few tools which can parse these data, requiring researchers to write their own solutions.
	- Genomics requires high throughput
  - Together, this can increase friction in integrating the MEME suite into an analysis workflow and may result in these powerful tools being underutilized during an analysis.
  - We made {dremeR}, which integrates the most popular MEME suite tools with the bioconductor stack to allow simple, rapid development of motif analysis steps.
		 - enables more iterations to identify best parameters
		 - describe uses of new functionality & improvied efficiency (briefly)
	- Also novel data visualization
		 - There are currently limited ways to visualize MEME data, but by bringing data into R, opens up the strengths of R data viz.
	** Couch some problems & soutiuons as novel develppments by this package**
	- explicitly say not changing core functions of MEME

# Main Text
  - A key design principle was to create tools that can be integrated in existing analysis pipelines, in addition to allowing users to create new ones.
## Figure: flowchart of each step, getsequence-> run-> import -> visualize -> goto run
## Overarching design principles/{dremeR} framework
  - MEME utilities work with files on disk & outputs data to files on disk
  - {dremeR} works by handling writing files from R object inputs, constructing the commandline calls, then reading in the data from the output files.
      - by default, this is all done using temporary files, so the data is not permanently stored on disk.
      - This is to shift the perspective from needing to save the files, to focusing on writing reproducible code which can regenerate the data when run.
      - However, users can set the `outdir` argument to purposefully save the outputs to disk if needed.
        - Also, if users use a 'fasta' file as input, the output data will be stored alongside the fasta file location
  - For functions that use MEME commandline tools, they need the location of 'meme/bin/' on disk, we provide utilities for making this detection simple to reduce friction of using the tool.
    - `check_meme_install` does user-friendly check that meme can be detected from R environment
    - Environment variable `MEME_PATH` (set & forget), `options(meme_path)` (configure for each script), `meme_path` argument, for each function call
  - Whenever possible, {dremeR} uses standard R data types for better integration with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this way, users don't need to learn new ways of working with data.

## Core Utilities
  - We built a common interface to each MEME utility. 
    - Tools taking sequence inputs can use fasta files as input, or a Biostrings representation of sequences.
    - Tools taking motif input can use .meme file path, universalmotif list, or the results object of functions returning motifs as output (allows seamless chaining of tools together)
  - All commandline arguments can be passed to each function, if users mistype the function error handling will suggest typo fixes.
  - `runDreme()`, `runMeme()` = Denovo motif discovery
     - Can be slow
  - `runAme()` = motif enrichment testing
  - `runTomTom()` = motif comparison (often used for matching denovo motifs to known database)
     - both ame & tomtom can use the `meme_db` setting which takes .meme path input, or universalmotif list.
     - can set & forget default in .Renviron
  - `runFimo()` if used with built-in helpers returns genomic coordinates of each motif in input sequences.
     - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
  - for dreme/ame/meme, we've paid special attention to enabling discriminative analysis
     - users can easily perform multiple comparisons of condtion1 vs 2, etc.
     - No special programming is required, no new patterns of data manipulation, etc. Just set input = list_of_sequences, control = "name(s) of entry in list" to compare all entries in list against that set of data (or pooled set if multiple entries).

## Data Visualization
  - `view_tomtom_hits` allows simple comparison of motifs w/ their matches, allowing users to quickly identify & use their domain-specific knowledge to decide if top hit is not "best" hit in context of experiment.
    - Will display "No Match" if there's no hit
    - can use the `force_best_match()` function to alter the best match info
     - see motif data manipulation section for more details
  - `ame_plot_heatmap` allows users to visualize patters of motif enrichment across multiple groups
     - `scale_max` & method = "normalize" provide additional ways to view the data
     - ame_compare_heatmap_methods helper gives users a way to select cutoffs & see tradeoffs of different visualizations, our package documentation (at <website> and in vignettes) go into great detail of how to reason about these differences.
     - function is built in `ggplot2` which can be used to further customize the heatmap
  - These plots can be exported as high-res SVGs for figures in papers

## Motif Metadata manipulation
  - Motif metadata is stored as a data.frame w/ special "motif" column holding the universalmotif object.
  - the `update_motifs()` function uses the data.frame values to update the motif metadata
    - allows users to manipulate motif metadata in bulk, then commit changes to motifs, which can be written to file using universalmotif `write_` functions, or used in plots, more {dremeR} functions, etc.
    - facilitates manipulation of `MotifDb` metadata, or integrating database information with gene expression, to, for example, filter the search space to only expressed TFs. Code examples of this are described in package documentation.

## Documentation & Tutorials
 - {dremeR} website & vignettes contain detailed practical tutorials on how to perform comprehensive analysis of ChIP-seq data, query MotifDb to build a database of expressed TFs, and more.

# Conclusion
 - {dremeR} fits into a broader framework to enable deep analysis of motif content of sequence data.

# Consider adding??
## Most MEME Suite Tools can run fine on a laptop
  - The webserver may give the appearance that lots of computational resources are needed for these tools,
    this isn't really true for TomTom, AME, and Fimo. DREME & MEME Have reasonable runtimes under certain conditions, but can be resource heavy depending on the task.
  - ? Provide data ?
    - Idea would be to run a set of tests on a laptop w/ i5 processor and ~4-8GB RAM (Like a Macbook), 
      and compare to a compute node w/ 16GB RAM, faster CPU. Just to show that you don't need a workhorse for it.

### 2020-06-05 outline.Rmd
rename package

@@
   - Together, this can increase friction in integrating the MEME suite into an analysis workflow and may result in these powerful tools being underutilized during an analysis.
-  - We made {dremeR}, which integrates the most popular MEME suite tools with the bioconductor stack to allow simple, rapid development of motif analysis steps.
+  - We made {memer}, which integrates the most popular MEME suite tools with the bioconductor stack to allow simple, rapid development of motif analysis steps.
 		 - enables more iterations to identify best parameters
@@
 ## Figure: flowchart of each step, getsequence-> run-> import -> visualize -> goto run
-## Overarching design principles/{dremeR} framework
+## Overarching design principles/{memer} framework
   - MEME utilities work with files on disk & outputs data to files on disk
-  - {dremeR} works by handling writing files from R object inputs, constructing the commandline calls, then reading in the data from the output files.
+  - {memer} works by handling writing files from R object inputs, constructing the commandline calls, then reading in the data from the output files.
       - by default, this is all done using temporary files, so the data is not permanently stored on disk.
@@
     - Environment variable `MEME_PATH` (set & forget), `options(meme_path)` (configure for each script), `meme_path` argument, for each function call
-  - Whenever possible, {dremeR} uses standard R data types for better integration with common analysis tools, like `dplyr`, `GenomicRanges`, or base R.
+  - Whenever possible, {memer} uses standard R data types for better integration with common analysis tools, like `dplyr`, `GenomicRanges`, or base R.
 In this way, users don't need to learn new ways of working with data.
@@
   - the `update_motifs()` function uses the data.frame values to update the motif metadata
-    - allows users to manipulate motif metadata in bulk, then commit changes to motifs, which can be written to file using universalmotif `write_` functions, or used in plots, more {dremeR} functions, etc.
+    - allows users to manipulate motif metadata in bulk, then commit changes to motifs, which can be written to file using universalmotif `write_` functions, or used in plots, more {memer} functions, etc.
     - facilitates manipulation of `MotifDb` metadata, or integrating database information with gene expression, to, for example, filter the search space to only expressed TFs.
@@
 ## Documentation & Tutorials
- - {dremeR} website & vignettes contain detailed practical tutorials on how to perform comprehensive analysis of ChIP-seq data, query MotifDb to build a database of expressed TFs, and more.
+ - {memer} website & vignettes contain detailed practical tutorials on how to perform comprehensive analysis of ChIP-seq data, query MotifDb to build a database of expressed TFs, and more.
 
 # Conclusion
- - {dremeR} fits into a broader framework to enable deep analysis of motif content of sequence data.
+ - {memer} fits into a broader framework to enable deep analysis of motif content of sequence data.
 

### 2020-06-10 outline.Rmd
adding links to similar papers
updates to outline & refs

@@
 
-# Main Text
+# Results 
+## Design & Implementation
   - A key design principle was to create tools that can be integrated in existing analysis pipelines, in addition to allowing users to create new ones.
@@
   - Whenever possible, {memer} uses standard R data types for better integration with common analysis tools, like `dplyr`, `GenomicRanges`, or base R.
-In this way, users don't need to learn new ways of working with data.
+In this way, users don't need to learn new ways of working with data, and outputs can be integrated with other tools with limited effort.
 
@@
 
+## Availability & Future Directions
+ - github (bioconductor?)
+ - Support additional MEME Suite tools & visualizations
+
 # Conclusion
@@
       and compare to a compute node w/ 16GB RAM, faster CPU. Just to show that you don't need a workhorse for it.
+
+# Other issues:
+ - PWMEnrich (similar to AME) is now deprecated in bioconductor 3.12 (author request)
+ - MotIV (has TOMTom like function) is now deprecated in 3.12
+
+# Similar software papers:
+https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1007654
+
+https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1007664

### 2020-06-10 manuscript.Rmd
add manuscript

---
title: "{memer}: an R interface to the MEME Suite"
author: 
 - name: Spencer L. Nystrom
   ORCID: 
 - name: Daniel J. McKay
output: github_document
---

```{r setup}
library(dremeR)
```

# Introduction

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, motif enrichment
testing, motif comparison, and motif matching, among others \cite{meme-suite}.
MEME Suite tools take as input fasta files of sequences which will be searched
for motifs, and if necessary, .meme format motif files containing motifs to use
in the search.

# Results 
## Design & Implementation

The key underlying design principle of `memer` was to create tools that easily
integrate into the existing suite of tools available in `Bioconductor` and the
broader `R` package landscape. In this way, `memer` tools can be easily integrated
into existing analysis pipelines, or used to create novel analyses.

MEME Suite tools are run on the commandline and use files stored on-disk as
input while returning a series of output files which together compose the full
results of the tool. While all tools produce machine-readable data formats, few
options exist for parsing the full complement of these data into human readable
format for use outside of the MEME Suite environment.

MEME Suite are run on the commandline and take 
as input fasta files of sequences
which will be searched for motifs, and if necessary, .meme format motif files
containing motifs to use in the search (REF: table of inputs).

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, motif enrichment
testing, motif comparison, and motif matching, among others \cite{meme-suite}.
MEME Suite tools take as input fasta files of sequences which will be searched
for motifs, and if necessary, .meme format motif files containing motifs to use
in the search.

At its most basic form, `memer` acts as a wrapper of these functions: users can
point `memer` tools to files on disk, `memer` will construct the commandline
call, output files will be saved on disk and imported into R for further analysis.
However, `memer` is more than just a simple wrapper of the MEME Suite, it is an
analysis framework allowing seamless integration of MEME Suite tools by
providing an abstraction layer which handles file input/output (I/O) to allow
inputs and outputs in the form of in-memory R objects.

When running a MEME Suite tool using `memer`, inputs are allowed in multiple
formats providing the user additional flexibility when performing motif analysis
(see "Core Utilities" for details). Of note, most outputs of `memer` tools can
be used as inputs to other `memer` tools, allowing easy construction of
"pipelines" (for example, discovering *de-novo* motifs with DREME, then finding
the genomic positions of those motifs in target regions using FIMO).

By shifting the focus from managing files on disk which may change infrequently to R data
objects, this framework reinforces the requirement of writing reproducible
analysis code to generate reproducible results.

Whenever possible, `memer` uses standard R data types for better integration
with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
way, users don't need to learn new ways of working with data, and outputs can
be integrated with other tools with limited effort.

### Detecting a local MEME Suite Install (remove this section?)

`memer` requires the user to install the MEME Suite to the analysis environment
in which `memer` will be used. The `check_meme_install()` function allows users
to determine whether their R session is able to detect a local install of the
MEME Suite.
  - For functions that use MEME commandline tools, they need the location of 'meme/bin/' on disk, we provide utilities for making this detection simple to reduce friction of using the tool.
    - `check_meme_install` does user-friendly check that meme can be detected from R environment
    - Environment variable `MEME_PATH` (set & forget), `options(meme_path)` (configure for each script), `meme_path` argument, for each function call

## Core Utilities
We built a common interface to each MEME Suite tool. Tools requiring sequences
as input can use either a path to a .fasta format file, a
`Biostrings::XStringSet`, or `Biostrings::BStringSetList` as input. These data
structures ensure support for all biological sequence types: DNA, RNA, and
amino-acid. For the remainder of the text, we will focus our examples on DNA
sequences, but `memer` tools support all sequence types. For DNA and RNA
sequences, the `get_sequence()` utility acts as the entrypoint to analysis with
`memer`. `get_sequence()` returns `Biostrings::XStringSet` or
`Biostrings::BStringSetList` types from range-based data types such as
`GenomicRanges` or `GenomicRangesList` which serve as sequence input to the
`memer` core functions.

Some `memer` tools require motifs as input. In these instances, `memer` will
always accept a path to a .meme format file, or a list of motifs in
`universalmotif` format \cite{universalmotif}. The `universalmotif` package
provides simple import and motif conversion functions for all common motif data
structures and file types. Using `universalmotif` objects ensures
cross-compatability with all commonly used motif data formats, providing the
user with maximum flexibility in their analysis. When `memer` functions return
motifs, they are returned to the user in `universalmotif` format. This ensures
that all `memer` motif outputs can be used as inputs to other `memer` functions
that accept motif inputs. Furthermore, `universalmotif` conversion tools allow
`memer` detected motifs to be used in many other popular Bioconductor motif
packages, such as `TFBSTools` and `motifStack`.

The core functionality of `memer` revolves around the interface to MEME Suite
tools. Each supported MEME Suite tool is assigned a `runToolname()` function,
which accepts inputs as described above and any valid commandline flags as
function arguments.

```{bash, eval=F}
# Shell command setting random seed to 100
dreme -s 100 input.fasta
```

```{r, eval=F}
# memer command setting random seed to 100
runDreme(input_sequences, control = "shuffle", s = 100)
```

### Discriminative analysis
```{r}
sequences_list <- list("Static" = "seq",
                       "Increasing" = "seq",
                       "Decreasing" = "seq")

# run Dreme on Static, Increasing, Decreasing sites using shuffled input as background
runDreme(sequences_list, "shuffle")

# run Dreme on Increasing, Decreasing sites using Static sites as background
runDreme(sequences_list, "Static")

# run Dreme on Decreasing sites using a combined set of Static + Increasing sites as background
runDreme(sequences_list, c("Static", "Increasing"))
```

1. MEME & DREME
 - get_sequences allows parsing genomic coordinates of match from MEME
1. Tools like dreme/meme/ame that can do discriminative testing
2. Tools like tomtom/fimo

### AME/MEME?
 - `get_sequence` solves the problem of having different requirements for fasta format to use certain features
  - score
  - parse_genomic_coord

 - and in some cases extends MEME Suite function to add additional features, such
 as parsing the genomic coordinates of meme matches.
 - also data visualization

\begin{figure}
\includegraphics[width=\textwidth]{figures/memer_model.png}
\caption{A diagram of memer tools & workflows?
flowchart of each step, getsequence-> run-> import -> visualize -> goto run

Diagram how meme-suite tools work vs memer?
}
\label{fig:model} 
\end{figure}

## Core Utilities
  - All commandline arguments can be passed to each function, if users mistype the function error handling will suggest typo fixes.
  - `runDreme()`, `runMeme()` = Denovo motif discovery
     - Can be slow
  - `runAme()` = motif enrichment testing
  - `runTomTom()` = motif comparison (often used for matching denovo motifs to known database)
     - both ame & tomtom can use the `meme_db` setting which takes .meme path input, or universalmotif list.
     - can set & forget default in .Renviron
  - `runFimo()` if used with built-in helpers returns genomic coordinates of each motif in input sequences.
     - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
  - for dreme/ame/meme, we've paid special attention to enabling discriminative analysis
     - users can easily perform multiple comparisons of condtion1 vs 2, etc.
     - No special programming is required, no new patterns of data manipulation, etc. Just set input = list_of_sequences, control = "name(s) of entry in list" to compare all entries in list against that set of data (or pooled set if multiple entries).

# Future developments:
 - Additional meme suite tools
 - better support for PSP generation

### 2020-06-10 manuscript.Rmd
edit

@@
 
+Example of commandline flag as function input:
 ```{bash, eval=F}
@@
 # memer command setting random seed to 100
+# user explicitly states they want to shuffle input sequences as control
 runDreme(input_sequences, control = "shuffle", s = 100)
@@
 ### Discriminative analysis
+Care taken to enable a simple interface to create complex discriminative analysis designs
 ```{r}

### 2020-06-10 manuscript.Rmd
adding more sections, begin rework to use ChIP-seq example to frame discussion.

--- 2020-06-10 manuscript.Rmd
@@
 ## Core Utilities
-We built a common interface to each MEME Suite tool. Tools requiring sequences
-as input can use either a path to a .fasta format file, a
-`Biostrings::XStringSet`, or `Biostrings::BStringSetList` as input. These data
-structures ensure support for all biological sequence types: DNA, RNA, and
-amino-acid. For the remainder of the text, we will focus our examples on DNA
-sequences, but `memer` tools support all sequence types. For DNA and RNA
-sequences, the `get_sequence()` utility acts as the entrypoint to analysis with
-`memer`. `get_sequence()` returns `Biostrings::XStringSet` or
-`Biostrings::BStringSetList` types from range-based data types such as
-`GenomicRanges` or `GenomicRangesList` which serve as sequence input to the
-`memer` core functions.
+
+The core functionality of `memer` revolves around the interface to MEME Suite
+tools. Each supported MEME Suite tool is assigned a `run` function
+(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
+accept sequence and motif inputs as needed and any valid commandline flags as
+additional function parameters.
+
+Tools requiring sequences as input can use a path to a .fasta format
+file, a `Biostrings::XStringSet`, or a `Biostrings::BStringSetList` as input.
+These data structures support all biological sequence types: DNA,
+RNA, and amino-acid. For the remainder of the text, we will 
+**use chip-seq data analysis as a framework to highlight memer features**
+\strikethrough{focus our examples on DNA sequences.}
+
+For DNA and RNA sequences, the `get_sequence()` utility acts
+as the entrypoint to analysis with `memer`. `get_sequence()` returns
+`Biostrings::XStringSet` or `Biostrings::BStringSetList` types from range-based
+data types such as `GenomicRanges` or `GenomicRangesList` to serve as
+sequence input to the `memer` core functions.
+
+When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
+Peaks) are commonly represented as `GenomicRanges` objects, which contain
+genomic positions of each peak, along with an arbitrary number of metadata
+columns describing features of each peak.
 
@@
 provides simple import and motif conversion functions for all common motif data
-structures and file types. Using `universalmotif` objects ensures
-cross-compatability with all commonly used motif data formats, providing the
-user with maximum flexibility in their analysis. When `memer` functions return
-motifs, they are returned to the user in `universalmotif` format. This ensures
-that all `memer` motif outputs can be used as inputs to other `memer` functions
-that accept motif inputs. Furthermore, `universalmotif` conversion tools allow
-`memer` detected motifs to be used in many other popular Bioconductor motif
-packages, such as `TFBSTools` and `motifStack`.
+structures and file types, ensuring cross-compatability with all commonly used
+motif data formats. When `memer` functions return motifs, they are returned to
+the user in `universalmotif` format. This ensures that all `memer` motif outputs
+can be used as inputs to other `memer` functions that accept motif inputs.
+Furthermore, `universalmotif` conversion tools allow `memer` detected motifs to
+be used in many other popular Bioconductor motif packages, such as `TFBSTools`
+and `motifStack`, providing the user with maximum flexibility in their analysis.
 
-The core functionality of `memer` revolves around the interface to MEME Suite
-tools. Each supported MEME Suite tool is assigned a `runToolname()` function,
-which accepts inputs as described above and any valid commandline flags as
-function arguments.
-
+**Example of commandline vs `memer`?**
+For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
 ```{bash, eval=F}
@@
 
+The `memer` implementation of the above command would be written as follows:
 ```{r, eval=F}
 # memer command setting random seed to 100
-runDreme(input_sequences, control = "shuffle", s = 100)
+# user explicitly states they want to shuffle input sequences as control
+runDreme("input.fasta", control = "shuffle", s = 100)
 ```
 
+**Somewhere say "memer can be used as a simple wrapper, but here we highlight its utility in chip-seq data analysis".**
+
+### Analysis using multiple groups (split-apply-combine approach)
+Users can adopt a "split-apply-combine" strategy for using `memer`. `GRanges`
+object acts as the centerpoint, can be split by data, fed to `get_sequence()`,
+then the list will run the tools for each group automatically.
+
 ### Discriminative analysis
+A key feature of several MEME Suite Tools is the ability to perform
+discriminative analysis, where motifs are identified or tested for enrichement
+in one group of sequences relative to another. In many instances, running tools
+in discriminative mode on a set of control sequences outperforms simply using
+shuffled input sequences as the control (back this up, change wording?). Therefore, 
+enabling complex discriminative analysis designs was central to the design of `memer`.
+
+When using list input to `memer` tools, names of list object members can be
+passed to the `control` argument. This will result in running the tool in
+discriminative mode using those sequences as the control sequences for all other
+list members. By passing multiple names to `control`, the sequences will be
+combined together to create a background set.
+
 ```{r}
@@
 
-1. MEME & DREME
- - get_sequences allows parsing genomic coordinates of match from MEME
-1. Tools like dreme/meme/ame that can do discriminative testing
-2. Tools like tomtom/fimo
+### Motif Comparison (TomTom)
+ - takes runDreme/meme output as input -> pipes
 
-### AME/MEME?
- - `get_sequence` solves the problem of having different requirements for fasta format to use certain features
-  - score
-  - parse_genomic_coord
-
- - and in some cases extends MEME Suite function to add additional features, such
- as parsing the genomic coordinates of meme matches.
- - also data visualization
+### Motif Matching (FIMO)
+ - returns locations as GRanges
+ - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
 
@@
 
-## Core Utilities
-  - All commandline arguments can be passed to each function, if users mistype the function error handling will suggest typo fixes.
-  - `runDreme()`, `runMeme()` = Denovo motif discovery
-     - Can be slow
-  - `runAme()` = motif enrichment testing
-  - `runTomTom()` = motif comparison (often used for matching denovo motifs to known database)
-     - both ame & tomtom can use the `meme_db` setting which takes .meme path input, or universalmotif list.
-     - can set & forget default in .Renviron
-  - `runFimo()` if used with built-in helpers returns genomic coordinates of each motif in input sequences.
-     - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
-  - for dreme/ame/meme, we've paid special attention to enabling discriminative analysis
-     - users can easily perform multiple comparisons of condtion1 vs 2, etc.
-     - No special programming is required, no new patterns of data manipulation, etc.
-Just set input = list_of_sequences, control = "name(s) of entry in list" to compare all entries in list against that set of data (or pooled set if multiple entries).
-
 # Future developments:

### 2020-06-11 manuscript.Rmd
small updates

@@
 in the search.
+
+Data from MEME Suite tools can be underutilized because HTML interface?
+ - HTML is great human-readable reporting interface
+ - `memer` allows customized reporting.
+
+Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memer` can fill the gap and add additional value.
 
@@
  - better support for PSP generation
+ - Better support for amino-acid sequence usage (please contact us with feature requests)

### 2020-06-24 manuscript.Rmd
cleaned up split-apply-combine section

@@
 integrate into the existing suite of tools available in `Bioconductor` and the
-broader `R` package landscape. In this way, `memer` tools can be easily integrated
-into existing analysis pipelines, or used to create novel analyses.
+broader `R` package landscape. In this way, `memer` tools can be used to extend
+existing analysis pipelines, or to create novel analyses.
 
@@
 the genomic positions of those motifs in target regions using FIMO).
+ - unlike current MEME tools (ex tomtom -> FIMO is hard to do)
 
@@
 
+**Describe each tool briefly**
+
+**table of comparison to commandline software?**
+  - input type:
+  - output type:
+
 Tools requiring sequences as input can use a path to a .fasta format
@@
 \strikethrough{focus our examples on DNA sequences.}
-
-For DNA and RNA sequences, the `get_sequence()` utility acts
-as the entrypoint to analysis with `memer`. `get_sequence()` returns
-`Biostrings::XStringSet` or `Biostrings::BStringSetList` types from range-based
-data types such as `GenomicRanges` or `GenomicRangesList` to serve as
-sequence input to the `memer` core functions.
-
-When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
-Peaks) are commonly represented as `GenomicRanges` objects, which contain
-genomic positions of each peak, along with an arbitrary number of metadata
-columns describing features of each peak.
 
@@
 and `motifStack`, providing the user with maximum flexibility in their analysis.
+
+When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
+Peaks) are commonly represented as `GenomicRanges` objects, which contain
+genomic positions of each peak, along with an arbitrary number of metadata
+columns describing features of each peak.
+
+A key difference between `memer` and the MEME Suite tools is that `memer`
+provides tools allowing the peak-level metadata to frame subsequent analysis.
+For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
+entrypoint to analysis with `memer`. `get_sequence()` returns a `Biostrings`
+object containing the DNA sequence of each feature in the input ranges. If
+ranges are split into groups by a metadata value, `get_sequence()` will return a
+list of sequences for each group. When a sequence list is used as input into
+`memer` core functions, the function will run once for each group, enabling a
+native "split-apply-combine" workflow without any additional user overhead
+\cite{splitapplycombine, whickham}. Together, these features empower users to
+use range-based metadata as the centerpoint of their analyses.
+
+### Discriminative analysis
+A powerful feature of several MEME Suite Tools is the ability to perform
+discriminative analysis, where motifs are identified or tested for enrichement
+in one group of sequences relative to another. In many instances, running tools
+in discriminative mode on a set of control sequences outperforms simply using
+shuffled input sequences as the control (back this up, change wording?). Therefore, 
+enabling complex discriminative analysis designs was central to the design of `memer`.
+
+When using list input to `memer` tools, names of list object members can be
+passed to the `control` argument. This results in running the tool in
+discriminative mode using those sequences as the control sequences for all other
+list members. By passing multiple names to `control`, the sequences are
+combined together to create a merged background set.
+
+```{r}
+sequences_list <- list("Static" = "seq",
+                       "Increasing" = "seq",
+                       "Decreasing" = "seq")
+
+# run Dreme on Static, Increasing, Decreasing sites using shuffled input as background
+runDreme(sequences_list, "shuffle")
+
+# run Dreme on Increasing, Decreasing sites using Static sites as background
+runDreme(sequences_list, "Static")
+
+# run Dreme on Decreasing sites using a combined set of Static + Increasing sites as background
+runDreme(sequences_list, c("Static", "Increasing"))
+```
 
@@
 
-### Analysis using multiple groups (split-apply-combine approach)
-Users can adopt a "split-apply-combine" strategy for using `memer`. `GRanges`
-object acts as the centerpoint, can be split by data, fed to `get_sequence()`,
-then the list will run the tools for each group automatically.
-
-### Discriminative analysis
-A key feature of several MEME Suite Tools is the ability to perform
-discriminative analysis, where motifs are identified or tested for enrichement
-in one group of sequences relative to another. In many instances, running tools
-in discriminative mode on a set of control sequences outperforms simply using
-shuffled input sequences as the control (back this up, change wording?). Therefore, 
-enabling complex discriminative analysis designs was central to the design of `memer`.
-
-When using list input to `memer` tools, names of list object members can be
-passed to the `control` argument. This will result in running the tool in
-discriminative mode using those sequences as the control sequences for all other
-list members. By passing multiple names to `control`, the sequences will be
-combined together to create a background set.
-
-```{r}
-sequences_list <- list("Static" = "seq",
-                       "Increasing" = "seq",
-                       "Decreasing" = "seq")
-
-# run Dreme on Static, Increasing, Decreasing sites using shuffled input as background
-runDreme(sequences_list, "shuffle")
-
-# run Dreme on Increasing, Decreasing sites using Static sites as background
-runDreme(sequences_list, "Static")
-
-# run Dreme on Decreasing sites using a combined set of Static + Increasing sites as background
-runDreme(sequences_list, c("Static", "Increasing"))
-```
-
 ### Motif Comparison (TomTom)
  - takes runDreme/meme output as input -> pipes
+ - enables using best matched motif as input to other tools (unlike TomTom on commandline)
 

### 2020-06-25 manuscript.Rmd
small edits

@@
 native "split-apply-combine" workflow without any additional user overhead
-\cite{splitapplycombine, whickham}. Together, these features empower users to
+\cite{splitapplycombine, wickham}. Together, these features empower users to
 use range-based metadata as the centerpoint of their analyses.
@@
 A powerful feature of several MEME Suite Tools is the ability to perform
-discriminative analysis, where motifs are identified or tested for enrichement
+discriminative analysis, where motifs are identified or tested for enrichment
 in one group of sequences relative to another. In many instances, running tools
 in discriminative mode on a set of control sequences outperforms simply using
-shuffled input sequences as the control (back this up, change wording?). Therefore, 
-enabling complex discriminative analysis designs was central to the design of `memer`.
+shuffled input sequences as the control (back this up, change wording? maybe:
+using biologically relevant control sequences can improve quality of results?).
+Therefore, enabling complex discriminative analysis designs was central to the
+design of `memer`.
 
@@
 ```
+
+## Metadata manipulation
+ - `update_motifs()`
+ - `force_best_match()`
+
+## Data Visualization
+ - tomtom best match
+ - ame heatmaps by group
 

### 2020-06-26 manuscript.Rmd
results draft to dan

@@
 ---
-title: "{memer}: an R interface to the MEME Suite"
+title: '{memer}: an R interface to the MEME Suite'
 author:
- - name: Spencer L. Nystrom
-   ORCID: 
+- ORCID: null
+  name: Spencer L. Nystrom
 - name: Daniel J. McKay
-output: github_document
+output: word_document
 ---
@@
 library(dremeR)
+library(magrittr)
 ```
@@
 be used as inputs to other `memer` tools, allowing easy construction of
-"pipelines" (for example, discovering *de-novo* motifs with DREME, then finding
+pipelines (for example, discovering *de-novo* motifs with DREME, then finding
 the genomic positions of those motifs in target regions using FIMO).
@@
 
-### Detecting a local MEME Suite Install (remove this section?)
-
-`memer` requires the user to install the MEME Suite to the analysis environment
-in which `memer` will be used. The `check_meme_install()` function allows users
-to determine whether their R session is able to detect a local install of the
-MEME Suite.
-  - For functions that use MEME commandline tools, they need the location of 'meme/bin/' on disk, we provide utilities for making this detection simple to reduce friction of using the tool.
-    - `check_meme_install` does user-friendly check that meme can be detected from R environment
-    - Environment variable `MEME_PATH` (set & forget), `options(meme_path)` (configure for each script), `meme_path` argument, for each function call
-
 ## Core Utilities
@@
 
-**Describe each tool briefly**
+**Describe each tool briefly?**
+ - This could also be relegated to a table
 
@@
 discriminative mode using those sequences as the control sequences for all other
-list members. By passing multiple names to `control`, the sequences are
-combined together to create a merged background set.
+list members. By passing multiple names to `control`, the sequences are combined
+together to create a merged background set. In this way, performing data-driven
+discriminative analysis is just as simple as one using shuffled sequences as
+background.
 
-```{r}
+```{r, eval=F}
+############
+# Replace this for a real set of peaks 
 sequences_list <- list("Static" = "seq",
@@
                        "Decreasing" = "seq")
+############
 
@@
 ## Metadata manipulation
- - `update_motifs()`
- - `force_best_match()`
+`memer` is built on top of the `universalmotif` framework, which provides
+methods for importing motifs from various file formats into a unified R data
+representation complete with metadata information \cite{universalmotif}.
+However, a major limitation of the `universalmotif` framework is that 
+manipulating or filtering multiple `universalmotif` entries requires potentially
+unfamiliar overhead for less experienced users, acting as a barrier to
+performing potentially useful operations.
 
-## Data Visualization
- - tomtom best match
- - ame heatmaps by group
+The `as_universalmotif_dataframe()` function from `memer` converts
+`universalmotif` objects into a more familiar `data.frame` format which can be
+manipulated using `tidyverse` or base R tools. These data.frames also contain a
+special `motif` column which stores the `universalmotif` object of each entry.
+Users can update columns in the data.frame which will be reflected in the
+corresponding metadata entries of the `motif` by running `update_motifs()` after
+manipulating the columns. These data.frames can be converted back into
+`universalmotif` format using `as_universalmotif`, which also updates the
+metadata using column values. The results from running the core MEME Suite
+functions are stored in this format whenever appropriate, creating a consistent
+framework for users to explore and manipulate motifs and their associated data values.
+
+**This section is maybe too listy and could be fleshed out more / integrated above better**
+Examples of common manipulations to motif databases which are aided by this
+format include bulk renaming of entries, identification and removal of redundant
+entries, and exploratory analysis of motif databases.
+
+```{r}
+# Example of renaming motifs w/ universalmotif syntax
+library(universalmotif)
+m1 <- create_motif()
+m2 <- create_motif()
+my_motifs <- list(m1,m2)
+
+my_names <- c("motif one", "motif two")
+
+for (i in 1:length(my_motifs)) {
+  my_motifs[[i]]@name <- my_names[i]
+}
+
+# Example of renaming motifs w/ memer syntax
+my_motifs <- list(m1,m2)
+my_motifs %<>% 
+  as_universalmotif_dataframe() %>% 
+  dplyr::mutate(name = my_names) %>% 
+  as_universalmotif()
+```
+
+This data format also allows users to integrate external data sources into their
+analyses, for example, to filter a motif database to include only motifs for
+genes which are expressed in their tissue of interest. Using motif databases
+filtered in this way has been demonstrated in multiple contexts to provide
+higher quality results when searching for novel regulators based on motif
+content \cite{hector franco, eric olson, john stamm?, shep & greenleaf, klu worm
+paper}.
+
+## Data Visualization (section unfinished)
+
+The MEME Suite provides a limited set of data visualizations which have limited
+customizability. `memer` leverages the advantages of the R graphics environment
+to provide novel data visualizations which are infinitely customizable by the user.
+
+  - `view_tomtom_hits()` allows examination of tomtom results, allowing the user
+  to evaluate the quality of the assigned match.
+
+```{r}
+# Example comparison of view_tomtom_hits results
+plot_topHits <- example_tomtom %>% 
+  dplyr::mutate(name = "Example Motif (Has Matches)") %>% 
+  update_motifs() %>% 
+  view_tomtom_hits(2)
+
+plot_noMatch <- create_motif() %>% 
+  as_universalmotif_dataframe() %>% 
+  dplyr::mutate(name = "Example Motif (No Matches)") %>% 
+  update_motifs() %>% 
+  {
+    x <- .
+    x$tomtom <- list(NULL)
+    return(x)
+  } %>% 
+  view_tomtom_hits()
+
+cowplot::plot_grid(plotlist = list(plot_topHits[[1]], plot_noMatch[[1]]), ncol = 2, labels = "AUTO")
+```
+
+  - `ame_plot_heatmap()` allows comparison of AME results between multiple
+  groups. Heirarchical ordering of hits shows commonalities and differences in enrichment between groups.
+```{r}
+# AME Heatmap example plots
+example_ame$Decreasing %>% 
+  ame_plot_heatmap()
+
+example_ame %>% 
+  dplyr::bind_rows(.id = "id") %>% 
+  ame_plot_heatmap(group = id)
+```
+
+# Conclusion
+ - `memer` is cool
+ - use `memer`
+ - it's on github (provide link)
+ - manual is here:
+
+# Notes
 
@@
 ```
-
-**Somewhere say "memer can be used as a simple wrapper, but here we highlight its utility in chip-seq data analysis".**
 

### 2020-07-23 manuscript.Rmd
added abstract for bioc2020

@@
 ```
+
+# Bioconductor Abstract (300 words)
+
+Biopolymers, such as DNA and protein, perform varying functions based on their
+primary sequence. Short, repeated sequences, or "motifs" represent functional
+units within biopolymers which can act as interaction surfaces, create
+structure, or have enzymatic activity. Identification of shared motifs across
+multiple sequences can provide evidence for shared function, thus the ability to
+identify, classify, and compare motifs represents a key step in the analysis of
+biological sequences. Although tools implementing these features exist within
+Bioconductor, they often rely on proprietary data types with poor
+interoperability with other tools. Furthermore, several of these packages will
+be deprecated in the next Bioconductor release, leaving a gap in functionality
+with no clear alternatives.
+
+The MEME Suite is a series of commandline tools which provide a variety of
+features for motif analysis including *de-novo* motif discovery, enrichment
+testing, comparison, and matching. Here we present `memer`, a motif analysis
+package which provides a seamless interface to MEME Suite tools within R.
+
+`memer` uses standard R and Bioconductor data types for better integration with
+common analysis tools like the `tidyverse` and `GRanges`. `memer` outputs also
+function as inputs to other `memer` commands, allowing simple construction of motif
+analysis pipelines.
+
+In addition to wrapping MEME Suite utilities, `memer` extends the functionality
+of `universalmotif` to provide base R and `tidyverse` compatible syntax for
+manipulating and analyzing motif metadata. These features allow motif databases
+to integrate external data to aid analysis, such as filtering a set of
+transcription factor motifs based on gene expression. `memer` also provides
+simple utilities for data visualization.
+
+`memer` is designed for maximum flexibility and ease of use to allow rapid
+iteration during analysis. Here we present several examples of how `memer`
+allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
+interoperability between MEME Suite tools and the broader R landscape.
+
+# Abstract Draft
+Biopolymers, such as DNA and protein, perform varying functions based on their
+primary sequence. Short, repeated sequences, or "motifs" represent functional
+units within biopolymers which can act as interaction surfaces, create
+structure, or have enzymatic activity. Identification of shared motifs across
+multiple sequences can provide evidence for shared function, thus the ability to
+identify, classify, and compare motifs represents a key step in the analysis of
+biological sequences. Although tools that implement these features have
+historically existed in the Bioconductor landscape, they often rely on
+proprietary data types with poor interoperability with other tools. Furthermore,
+several of these packages will be deprecated in the next Bioconductor release,
+leaving a gap in functionality with no clear alternatives.
+
+The MEME Suite is a series of commandline tools which provide a variety of
+features for motif analysis such as *de-novo* motif discovery, enrichment
+testing,comparison, and matching. Here we present `memer`, a motif analysis
+package which provides a seamless interface to MEME Suite tools within R.
+
+`memer` uses base R and Bioconductor data types for better integration with
+common analysis tools like the `tidyverse` and other Bioconductor packages.
+Unlike the commandline implementation, `memer` outputs also function as inputs
+to other MEME Suite tools, allowing simple construction of motif analysis
+pipelines.
+
+In addition to wrapping MEME Suite utilities, `memer` extends the functionality
+of `universalmotif` to provide base R and `tidyverse` compatible syntax for
+manipulating and analyzing motif metadata. These features allow motif databases
+to integrate external data to aid analysis, such as filtering a set of
+transcription factor motifs based on gene expression in the tissue
+of interest. `memer` also provides simple utilities for data visualization where
+appropriate.
+
+`memer` is designed for maximum flexibility and ease of use to allow users to
+iterate rapidly during analysis. Here we present several examples of how `memer`
+allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
+interoperability between MEME Suite tools and the broader R package landscape.
 

### 2020-07-23 manuscript.Rmd
change pkg name to {memes}

@@
 ---
-title: '{memer}: an R interface to the MEME Suite'
+title: '{memes}: an R interface to the MEME Suite'
 author:
@@
 features for motif analysis including *de-novo* motif discovery, enrichment
-testing, comparison, and matching. Here we present `memer`, a motif analysis
+testing, comparison, and matching. Here we present `memes`, a motif analysis
 package which provides a seamless interface to MEME Suite tools within R.
 
-`memer` uses standard R and Bioconductor data types for better integration with
-common analysis tools like the `tidyverse` and `GRanges`. `memer` outputs also
-function as inputs to other `memer` commands, allowing simple construction of motif
+`memes` uses standard R and Bioconductor data types for better integration with
+common analysis tools like the `tidyverse` and `GRanges`. `memes` outputs also
+function as inputs to other `memes` commands, allowing simple construction of motif
 analysis pipelines.
 
-In addition to wrapping MEME Suite utilities, `memer` extends the functionality
+In addition to wrapping MEME Suite utilities, `memes` extends the functionality
 of `universalmotif` to provide base R and `tidyverse` compatible syntax for
@@
 to integrate external data to aid analysis, such as filtering a set of
-transcription factor motifs based on gene expression. `memer` also provides
+transcription factor motifs based on gene expression. `memes` also provides
 simple utilities for data visualization.
 
-`memer` is designed for maximum flexibility and ease of use to allow rapid
-iteration during analysis. Here we present several examples of how `memer`
+`memes` is designed for maximum flexibility and ease of use to allow rapid
+iteration during analysis. Here we present several examples of how `memes`
 allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
@@
 features for motif analysis such as *de-novo* motif discovery, enrichment
-testing,comparison, and matching. Here we present `memer`, a motif analysis
+testing,comparison, and matching. Here we present `memes`, a motif analysis
 package which provides a seamless interface to MEME Suite tools within R.
 
-`memer` uses base R and Bioconductor data types for better integration with
+`memes` uses base R and Bioconductor data types for better integration with
 common analysis tools like the `tidyverse` and other Bioconductor packages.
-Unlike the commandline implementation, `memer` outputs also function as inputs
+Unlike the commandline implementation, `memes` outputs also function as inputs
 to other MEME Suite tools, allowing simple construction of motif analysis
@@
 
-In addition to wrapping MEME Suite utilities, `memer` extends the functionality
+In addition to wrapping MEME Suite utilities, `memes` extends the functionality
 of `universalmotif` to provide base R and `tidyverse` compatible syntax for
@@
 transcription factor motifs based on gene expression in the tissue
-of interest. `memer` also provides simple utilities for data visualization where
+of interest. `memes` also provides simple utilities for data visualization where
 appropriate.
 
-`memer` is designed for maximum flexibility and ease of use to allow users to
-iterate rapidly during analysis. Here we present several examples of how `memer`
+`memes` is designed for maximum flexibility and ease of use to allow users to
+iterate rapidly during analysis. Here we present several examples of how `memes`
 allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
@@
  - HTML is great human-readable reporting interface
- - `memer` allows customized reporting.
+ - `memes` allows customized reporting.
 
-Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memer` can fill the gap and add additional value.
+Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memes` can fill the gap and add additional value.
 
@@
 
-The key underlying design principle of `memer` was to create tools that easily
+The key underlying design principle of `memes` was to create tools that easily
 integrate into the existing suite of tools available in `Bioconductor` and the
-broader `R` package landscape. In this way, `memer` tools can be used to extend
+broader `R` package landscape. In this way, `memes` tools can be used to extend
 existing analysis pipelines, or to create novel analyses.
@@
 
-At its most basic form, `memer` acts as a wrapper of these functions: users can
-point `memer` tools to files on disk, `memer` will construct the commandline
+At its most basic form, `memes` acts as a wrapper of these functions: users can
+point `memes` tools to files on disk, `memes` will construct the commandline
 call, output files will be saved on disk and imported into R for further analysis.
-However, `memer` is more than just a simple wrapper of the MEME Suite, it is an
+However, `memes` is more than just a simple wrapper of the MEME Suite, it is an
 analysis framework allowing seamless integration of MEME Suite tools by
@@
 
-When running a MEME Suite tool using `memer`, inputs are allowed in multiple
+When running a MEME Suite tool using `memes`, inputs are allowed in multiple
 formats providing the user additional flexibility when performing motif analysis
-(see "Core Utilities" for details). Of note, most outputs of `memer` tools can
-be used as inputs to other `memer` tools, allowing easy construction of
+(see "Core Utilities" for details). Of note, most outputs of `memes` tools can
+be used as inputs to other `memes` tools, allowing easy construction of
 pipelines (for example, discovering *de-novo* motifs with DREME, then finding
@@
 
-Whenever possible, `memer` uses standard R data types for better integration
+Whenever possible, `memes` uses standard R data types for better integration
 with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
@@
 
-The core functionality of `memer` revolves around the interface to MEME Suite
+The core functionality of `memes` revolves around the interface to MEME Suite
 tools. Each supported MEME Suite tool is assigned a `run` function
@@
 RNA, and amino-acid. For the remainder of the text, we will 
-**use chip-seq data analysis as a framework to highlight memer features**
+**use chip-seq data analysis as a framework to highlight memes features**
 \strikethrough{focus our examples on DNA sequences.}
 
-Some `memer` tools require motifs as input. In these instances, `memer` will
+Some `memes` tools require motifs as input. In these instances, `memes` will
 always accept a path to a .meme format file, or a list of motifs in
@@
 structures and file types, ensuring cross-compatability with all commonly used
-motif data formats. When `memer` functions return motifs, they are returned to
-the user in `universalmotif` format. This ensures that all `memer` motif outputs
-can be used as inputs to other `memer` functions that accept motif inputs.
-Furthermore, `universalmotif` conversion tools allow `memer` detected motifs to
+motif data formats. When `memes` functions return motifs, they are returned to
+the user in `universalmotif` format. This ensures that all `memes` motif outputs
+can be used as inputs to other `memes` functions that accept motif inputs.
+Furthermore, `universalmotif` conversion tools allow `memes` detected motifs to
 be used in many other popular Bioconductor motif packages, such as `TFBSTools`
@@
 
-A key difference between `memer` and the MEME Suite tools is that `memer`
+A key difference between `memes` and the MEME Suite tools is that `memes`
 provides tools allowing the peak-level metadata to frame subsequent analysis.
 For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
-entrypoint to analysis with `memer`. `get_sequence()` returns a `Biostrings`
+entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
 object containing the DNA sequence of each feature in the input ranges. If
@@
 list of sequences for each group. When a sequence list is used as input into
-`memer` core functions, the function will run once for each group, enabling a
+`memes` core functions, the function will run once for each group, enabling a
 native "split-apply-combine" workflow without any additional user overhead
@@
 Therefore, enabling complex discriminative analysis designs was central to the
-design of `memer`.
+design of `memes`.
 
-When using list input to `memer` tools, names of list object members can be
+When using list input to `memes` tools, names of list object members can be
 passed to the `control` argument. This results in running the tool in
@@
 ## Metadata manipulation
-`memer` is built on top of the `universalmotif` framework, which provides
+`memes` is built on top of the `universalmotif` framework, which provides
 methods for importing motifs from various file formats into a unified R data
@@
 
-The `as_universalmotif_dataframe()` function from `memer` converts
+The `as_universalmotif_dataframe()` function from `memes` converts
 `universalmotif` objects into a more familiar `data.frame` format which can be
@@
 
-# Example of renaming motifs w/ memer syntax
+# Example of renaming motifs w/ memes syntax
 my_motifs <- list(m1,m2)
@@
 The MEME Suite provides a limited set of data visualizations which have limited
-customizability. `memer` leverages the advantages of the R graphics environment
+customizability. `memes` leverages the advantages of the R graphics environment
 to provide novel data visualizations which are infinitely customizable by the user.
@@
 # Conclusion
- - `memer` is cool
- - use `memer`
+ - `memes` is cool
+ - use `memes`
  - it's on github (provide link)
@@
 
-**Example of commandline vs `memer`?**
+**Example of commandline vs `memes`?**
 For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
@@
 
-The `memer` implementation of the above command would be written as follows:
+The `memes` implementation of the above command would be written as follows:
 ```{r, eval=F}
-# memer command setting random seed to 100
+# memes command setting random seed to 100
 # user explicitly states they want to shuffle input sequences as control
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{figures/memer_model.png}
-\caption{A diagram of memer tools & workflows?
+\includegraphics[width=\textwidth]{figures/memes_model.png}
+\caption{A diagram of memes tools & workflows?
 flowchart of each step, getsequence-> run-> import -> visualize -> goto run
 
-Diagram how meme-suite tools work vs memer?
+Diagram how meme-suite tools work vs memes?
 }

### 2020-07-28 manuscript.Rmd
updating design & implementation

@@
 
+Biopolymers, such as DNA and protein, perform varying functions based on their
+primary sequence. Short, repeated sequences, or "motifs" represent functional
+units within biopolymers which can act as interaction surfaces, create
+structure, or have enzymatic activity. Identification of shared motifs across
+multiple sequences can provide evidence for shared function, thus the ability to
+identify, classify, and compare motifs represents a key step in the analysis of
+biological sequences. Although tools that implement these features have
+historically existed in the Bioconductor landscape, they often rely on
+proprietary data types with poor interoperability with other tools. Furthermore,
+several of these packages will be deprecated in the next Bioconductor release,
+leaving a gap in functionality with no clear alternatives.
+
+The MEME Suite is a series of commandline and web-based tools which provide a variety
+of features for motif analysis such as *de-novo* motif discovery, enrichment
+testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
+accepted set of tools with over 22,000 citations to date, and over 30,000 unique
+users of the webserver implementation annually \cite{meme-usage-report-page,
+google-scholar-meme}.
+
+The MEME Suite family of tools provides a
+comprehensive set of utilities to interrogate motif content across a broad range
+of biological contexts. However, several factors limit the full potential of
+these tools for use in data analysis. 
+  1) must do significant preprocessing to get data in correct input format
+  2) Easily parseable output data often doesn't contain all information
+    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
+      - matrix is present in XML file, but no tools exist to parse this
+  3) limited data visualization capabilities
+
+Here we present `memes`, a motif analysis package which provides a seamless
+interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
+data types for data input and output, facilitating better integration with
+common analysis tools like the `tidyverse` and other Bioconductor packages.
+Unlike the commandline implementation, `memes` outputs also function as inputs
+to other MEME Suite tools, allowing simple construction of motif analysis
+pipelines without additional data processing steps. Additionally, R/Bioconductor data
+structures provide a full-featured representaiton of MEME Suite output data, providing
+users quick access to all relevant data structures with simple syntax.
+
+In addition to wrapping MEME Suite utilities, `memes` extends the functionality
+of `universalmotif` to provide base R and `tidyverse` compatible syntax for
+manipulating and analyzing motif metadata. These features allow motif databases
+to integrate external data to aid analysis, such as filtering a set of
+transcription factor motifs based on gene expression in the tissue of interest.
+
+`memes` also provides simple utilities for data visualization where appropriate.
+Providing R representations of MEME Suite data objects enables generation of
+novel data visualizations using the well established tools available in the R
+landscape. Therefore, `memes` empowers users to deeply analyze and visualize
+motif analysis data.
+
+`memes` is designed for maximum flexibility and ease of use to allow users to
+iterate rapidly during analysis. Here we present several examples of how `memes`
+allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
+interoperability between MEME Suite tools and the broader R package landscape.
+
 # Results 
@@
 
+### Datatypes
+
+MEME Suite tools are run on the commandline and use files stored on-disk as
+input while returning a series of output files containing varying data types.
+`memes` reduces each MEME Suite tool to an R function which internally handles
+file I/O such that all MEME Suite calling functions accept R objects as input,
+and return R objects as output. Sequence inputs are stored in `BioStrings`
+format, an R/Bioconductor package for storing biopolymer sequence data. Motif
+inputs are passed as `universalmotif` objects, another R/Bioconductor package
+for representing motif matrices and their associated metadata. `memes` will also
+always accept paths to files on disk, such as fasta files for sequence inputs,
+and meme format files for motif inputs.
+
+The MEME Suite tools do not return consistent data structures across tools. For
+example, the tool TomTom, which compares query motifs to a database of known
+motifs, returns tab-separated results which do not contain the frequency matrix
+representations of the matched motifs. Instead, users must parse these matrices
+back out from the input databases, creating additional friction for creating
+novel data visualizations and exploratory analyses. Often, these data are
+contained in XML formatted outputs, which although they hold all relevant
+information, are difficult to parse. To our knowledge, no publicly available
+tools exist to parse MEME Suite XML formatted data, therefore users must write
+custom code to extract these data. `memes` exports data import functions for
+each supported MEME Suite tool, which import these data as R data.frames. These
+functions work internally to create the seamless file I/O when calling MEME
+Suite tools, but the import functions can be called directly by users to import
+previously generated data for use in R.
+
+Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
+describe properties of motifs. For example, MEME and DREME return de-novo
+discovered motifs from query regions along with statistical information about
+their enrichment. The `universalmotif` data structure is a powerful
+R/Bioconductor representation for motif matrices and their associated metadata. 
+However, `universalmotif` objects are not ideal for performing bulk manipulation
+or queries of metadata, nor do they support complex data structures such as
+nested elements. Many MEME Suite analysis pipelines involve identifying motifs
+which are further analyzed by multiple tools, creating additional metadata. Thus
+an ideal representation for these data is one which can hold an unlimited amount
+of metadata, and contain complex heirarchies. `memes` therefore extends the
+`universalmotif` format by providing data.frames holding motif metadata which
+retain links to their `universalmotif` representation. These universalmotif
+data.frames combine the advantages of the `universalmotif` framework for motif
+manipulation with the extensibility and flexibility of R data.frames, and are
+the basis for a majority of `memes` data outputs. These structures are also
+valid input types to `memes` functions, and when used as such, output data are
+appended as new columns to the input data, ensuring data provenance.
+
+**Figure?: showing how the data.frame connects to universalmotif**
+
+`universalmotif` data.frames can additionally be used to examine or manipulate
+the properties of any motif database. Using `universalmotif` functions to import
+motif data, `as_universalmotif_dataframe()` converts any `universalmotif` object
+into a `universalmotif` data.frame which can be manipulated using base R or
+`tidyverse` functions, just like regular R data.frames. Manipulation of linked
+metadata columns will result in updating the linked `universalmotif` object when
+calling `update_motifs()` to retain the data.frame structure, or when calling
+`as_universalmotif()` to convert the data.frame into `universalmotif` format. In
+this way, `memes` enables large-scale motif database manipulation & analysis
+within the preexisting R framework.
+
+### Genomic range-based data as entrypoint
+
+When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
+Peaks) are commonly represented as `GenomicRanges` objects, which contain
+genomic positions of each peak, along with an arbitrary number of metadata
+columns describing features of each peak.
+
+A key difference between `memes` and the MEME Suite tools is that `memes`
+provides tools allowing the peak-level metadata to frame subsequent analysis.
+For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
+entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
+object containing the DNA sequence of each feature in the input ranges. If
+ranges are split into groups by a metadata value, `get_sequence()` will return a
+list of sequences for each group. When a sequence list is used as input into
+`memes` core functions, the function will run once for each group, enabling a
+native "split-apply-combine" workflow without any additional user overhead
+\cite{splitapplycombine, wickham}. Together, these features empower users to
+use range-based metadata as the centerpoint of their analyses.
+
+### Old Design & Implementation
 The key underlying design principle of `memes` was to create tools that easily

### 2020-07-28 manuscript.Rmd
more reworking design section

@@
 
+### Core Utilities
+
+The core functionality of `memes` revolves around the interface to MEME Suite
+tools. Each supported MEME Suite tool is assigned a `run` function
+(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
+accept sequence and motif inputs as needed and any valid commandline flags as
+additional function parameters.
+
+**Describe each tool briefly?**
+ - This could also be relegated to a table
+
+**table of comparison to commandline software?**
+  - input type:
+  - output type:
+
 ### Datatypes
@@
 
-When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
-Peaks) are commonly represented as `GenomicRanges` objects, which contain
-genomic positions of each peak, along with an arbitrary number of metadata
-columns describing features of each peak.
+**CONTINUE**
+
+MEME Suite focuses on sequence data as the centerpoint of analysis, but in the
+case of ChIP-seq, data are often stored as genomic coordinates with additional
+metadata columns, such as `GRanges` objects in R/Bioconductor.
 
@@
 
-## Core Utilities
-
-The core functionality of `memes` revolves around the interface to MEME Suite
-tools. Each supported MEME Suite tool is assigned a `run` function
-(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
-accept sequence and motif inputs as needed and any valid commandline flags as
-additional function parameters.
-
-**Describe each tool briefly?**
- - This could also be relegated to a table
-
-**table of comparison to commandline software?**
-  - input type:
-  - output type:
-
-Tools requiring sequences as input can use a path to a .fasta format
-file, a `Biostrings::XStringSet`, or a `Biostrings::BStringSetList` as input.
-These data structures support all biological sequence types: DNA,
-RNA, and amino-acid. For the remainder of the text, we will 
-**use chip-seq data analysis as a framework to highlight memes features**
-\strikethrough{focus our examples on DNA sequences.}
-
-Some `memes` tools require motifs as input. In these instances, `memes` will
-always accept a path to a .meme format file, or a list of motifs in
-`universalmotif` format \cite{universalmotif}. The `universalmotif` package
-provides simple import and motif conversion functions for all common motif data
-structures and file types, ensuring cross-compatability with all commonly used
-motif data formats. When `memes` functions return motifs, they are returned to
-the user in `universalmotif` format. This ensures that all `memes` motif outputs
-can be used as inputs to other `memes` functions that accept motif inputs.
 Furthermore, `universalmotif` conversion tools allow `memes` detected motifs to
@@
 and `motifStack`, providing the user with maximum flexibility in their analysis.
-
-When analyzing ChIP-seq data in R, transcription factor binding locations (ChIP
-Peaks) are commonly represented as `GenomicRanges` objects, which contain
-genomic positions of each peak, along with an arbitrary number of metadata
-columns describing features of each peak.
-
-A key difference between `memes` and the MEME Suite tools is that `memes`
-provides tools allowing the peak-level metadata to frame subsequent analysis.
-For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
-entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
-object containing the DNA sequence of each feature in the input ranges. If
-ranges are split into groups by a metadata value, `get_sequence()` will return a
-list of sequences for each group. When a sequence list is used as input into
-`memes` core functions, the function will run once for each group, enabling a
-native "split-apply-combine" workflow without any additional user overhead
-\cite{splitapplycombine, wickham}. Together, these features empower users to
-use range-based metadata as the centerpoint of their analyses.
 

### 2020-09-20 manuscript.Rmd
Removed old draft text, added more sections.

@@
 features for motif analysis such as *de-novo* motif discovery, enrichment
-testing,comparison, and matching. Here we present `memes`, a motif analysis
+testing, comparison, and matching. Here we present `memes`, a motif analysis
 package which provides a seamless interface to MEME Suite tools within R.
@@
   1) must do significant preprocessing to get data in correct input format
+    - written to analyze 1 group of sequences, not amenable to multigroup designs
+    - no common patterns for generating inputs from ChIP-seq whose data are primarily range based
   2) Easily parseable output data often doesn't contain all information
@@
   3) limited data visualization capabilities
+  4) difficult to integrate data with additional sources
 
@@
 to other MEME Suite tools, allowing simple construction of motif analysis
-pipelines without additional data processing steps. Additionally, R/Bioconductor data
-structures provide a full-featured representaiton of MEME Suite output data, providing
-users quick access to all relevant data structures with simple syntax.
+pipelines without additional data processing steps. Additionally, R/Bioconductor
+data structures provide a full-featured representaiton of MEME Suite output
+data, providing users quick access to all relevant data structures with simple
+syntax.
 
@@
 
-### Datatypes
+### Data Structures
 
@@
 and return R objects as output. Sequence inputs are stored in `BioStrings`
-format, an R/Bioconductor package for storing biopolymer sequence data. Motif
-inputs are passed as `universalmotif` objects, another R/Bioconductor package
-for representing motif matrices and their associated metadata. `memes` will also
-always accept paths to files on disk, such as fasta files for sequence inputs,
-and meme format files for motif inputs.
+format, an R/Bioconductor package for storing biopolymer sequence data
+\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
+R/Bioconductor package for representing motif matrices and their associated
+metadata \cite{universalmotif}. `memes` will also always accept paths to files
+on disk, such as fasta files for sequence inputs, and meme format files for
+motif inputs, reducing the need to read large files into memory to run MEME
+Suite tools.
 
-The MEME Suite tools do not return consistent data structures across tools. For
-example, the tool TomTom, which compares query motifs to a database of known
-motifs, returns tab-separated results which do not contain the frequency matrix
-representations of the matched motifs. Instead, users must parse these matrices
-back out from the input databases, creating additional friction for creating
-novel data visualizations and exploratory analyses. Often, these data are
-contained in XML formatted outputs, which although they hold all relevant
-information, are difficult to parse. To our knowledge, no publicly available
-tools exist to parse MEME Suite XML formatted data, therefore users must write
-custom code to extract these data. `memes` exports data import functions for
-each supported MEME Suite tool, which import these data as R data.frames. These
-functions work internally to create the seamless file I/O when calling MEME
-Suite tools, but the import functions can be called directly by users to import
-previously generated data for use in R.
+Although MEME Suite tools accept consistent data type inputs, they do not return
+consistent data structures across tools. For example, the tool TomTom, which
+compares query motifs to a database of known motifs, returns tab-separated
+results which do not contain the frequency matrix representations of the matched
+motifs. Instead, users must write custom code to parse these matrices back out
+from the input databases, creating additional friction for creating novel data
+visualizations and exploratory analyses. Often, output data are also returned in
+XML format, which although they hold all relevant information, are difficult to
+parse. To our knowledge, no publicly available tools exist to parse MEME Suite
+XML formatted data, therefore users must write custom code to extract these
+data. `memes` exports data import functions for each supported MEME Suite tool,
+which import these data as R data.frames. These functions also work internally to
+create the seamless file I/O when calling MEME Suite tools, but the import
+functions can be called directly by users to import previously generated data
+for use in R.
+
+ - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
+ - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
+ - **Need to revise above statement, but note that most tools don't have support**
 
@@
 describe properties of motifs. For example, MEME and DREME return de-novo
-discovered motifs from query regions along with statistical information about
+discovered motifs from query sequences along with statistical information about
 their enrichment. The `universalmotif` data structure is a powerful
-R/Bioconductor representation for motif matrices and their associated metadata. 
+R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
 However, `universalmotif` objects are not ideal for performing bulk manipulation
@@
 nested elements. Many MEME Suite analysis pipelines involve identifying motifs
-which are further analyzed by multiple tools, creating additional metadata. Thus
+which are further analyzed by multiple tools, creating additional metadata which
+cannot be easily tracked in bulk using `universalmotif` format. Thus
 an ideal representation for these data is one which can hold an unlimited amount
-of metadata, and contain complex heirarchies. `memes` therefore extends the
-`universalmotif` format by providing data.frames holding motif metadata which
-retain links to their `universalmotif` representation. These universalmotif
+of metadata, contain complex heirarchies, and be easily manipulated using
+standard analysis tools. `memes` therefore extends the `universalmotif` format
+by providing data.frames holding motif metadata which retain links to their
+`universalmotif` representation. These universalmotif
 data.frames combine the advantages of the `universalmotif` framework for motif
-manipulation with the extensibility and flexibility of R data.frames, and are
-the basis for a majority of `memes` data outputs. These structures are also
-valid input types to `memes` functions, and when used as such, output data are
-appended as new columns to the input data, ensuring data provenance.
+manipulation with the extensibility and flexibility of R data.frames. These
+`universalmotif` data.frames form the basis for a majority of `memes` data
+outputs. These structures are also valid input types to `memes` functions, and
+when used as such, output data are appended as new columns to the input data,
+ensuring data provenance.
 
@@
 
-`universalmotif` data.frames can additionally be used to examine or manipulate
-the properties of any motif database. Using `universalmotif` functions to import
-motif data, `as_universalmotif_dataframe()` converts any `universalmotif` object
+`universalmotif` data.frames are not limited to storing motifs from MEME Suite
+data, and can additionally be used to examine or manipulate the properties of
+any motif list or database. Using `universalmotif` functions to import motif data,
+the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
 into a `universalmotif` data.frame which can be manipulated using base R or
@@
 calling `update_motifs()` to retain the data.frame structure, or when calling
-`as_universalmotif()` to convert the data.frame into `universalmotif` format. In
+`as_universalmotif()` to convert the data.frame back into `universalmotif` format. In
 this way, `memes` enables large-scale motif database manipulation & analysis
@@
 
-### Genomic range-based data as entrypoint
+### Support for genomic range-based data
 
-**CONTINUE**
+# RETURN
+Motif analysis is important for assays like ChIP-seq which store data as genomic
+coordinates, not sequence. This is in conflict with MEME Suite which is designed
+to study sequences. Therefore, users must write custom code to extract DNA
+sequence of their ranges of interest. Although tools like `bedtools getfasta`
+can simplify the process of extracting DNA sequence from genomic coordinates,
+some MEME tools require fasta headers to be specially formatted to extract full
+functionality, requiring users to write additional custom code to properly
+format input data.
 
-MEME Suite focuses on sequence data as the centerpoint of analysis, but in the
-case of ChIP-seq, data are often stored as genomic coordinates with additional
-metadata columns, such as `GRanges` objects in R/Bioconductor.
+`memes` provides the `get_sequence` function which automates the extraction of
+DNA sequence from genomic coordinates while simultaneously producing MEME Suite
+formatted fasta headers. `get_sequence` accepts genomic-range based inputs in
+`GenomicRanges` format, the *de-facto* standard for genomic coordinate
+representation in R \cite{genomicranges}. Other common genomic coordinate
+representations, such as bed format, are easily imported as `GenomicRanges`
+objects into R using preexisting import functions, meaning users must not write
+any custom import functions to work with range-based data using `memes`. 
+Sequences are returned in `Biostrings` format and can therefore be used as
+inputs to all `memes` commands or as inputs to other R/Bioconductor functions
+for sequence analysis.
 
-A key difference between `memes` and the MEME Suite tools is that `memes`
-provides tools allowing the peak-level metadata to frame subsequent analysis.
-For DNA ranges (such as in ChIP-seq), the `get_sequence()` utility acts as the
-entrypoint to analysis with `memes`. `get_sequence()` returns a `Biostrings`
-object containing the DNA sequence of each feature in the input ranges. If
-ranges are split into groups by a metadata value, `get_sequence()` will return a
-list of sequences for each group. When a sequence list is used as input into
-`memes` core functions, the function will run once for each group, enabling a
-native "split-apply-combine" workflow without any additional user overhead
-\cite{splitapplycombine, wickham}. Together, these features empower users to
-use range-based metadata as the centerpoint of their analyses.
+During analysis of multiomic data, users often uncover different functional
+categories of genomic regions through integration of multiple datasets. In these
+instances, motif analysis acts as a powerful tool to examine whether differences
+in motif content is associated with membership of distinct categories. In some
+cases, performing differential motif analysis (using one category as the "test"
+set and another as the "control") can uncover biologically relevant motifs which
+are enriched in one category relative to another. Although the MEME Suite allows
+differential enrichment testing, it does not inheritly provide a mechanism for
+analyzing groups of sequences in parallel, or performing motif analysis with any
+understanding of data categories. The `memes` framework enables "data-aware"
+motif analysis workflows by allowing named lists of `Biostrings` objects as
+input to each function. If using `GenomicRanges`, users can split peak data on a
+metadata column using the base R `split` function, and use this result as input
+to `get_sequence`, which will produce a list of `BioStrings` objects where each
+entry is named after each data category from the split column. When a list is
+used as input to a `memes` function, it runs the corresponding MEME Suite tool
+for each object in the list. Users can also pass the name(s) of a category to
+the `control` argument to automatically enable differential analysis of all
+other list members against the control category sequences.  Therefore, `memes`
+enables complex data-aware differential motif analysis workflows using simple
+syntax to extend the capabilities of the MEME Suite.
 
-### Old Design & Implementation
-The key underlying design principle of `memes` was to create tools that easily
-integrate into the existing suite of tools available in `Bioconductor` and the
-broader `R` package landscape. In this way, `memes` tools can be used to extend
-existing analysis pipelines, or to create novel analyses.
+### Table of supported MEME Suite Tools
+ - comparison to other tools?
+ - Maybe only do this if asked
 
-MEME Suite tools are run on the commandline and use files stored on-disk as
-input while returning a series of output files which together compose the full
-results of the tool. While all tools produce machine-readable data formats, few
-options exist for parsing the full complement of these data into human readable
-format for use outside of the MEME Suite environment.
-
-MEME Suite are run on the commandline and take 
-as input fasta files of sequences
-which will be searched for motifs, and if necessary, .meme format motif files
-containing motifs to use in the search (REF: table of inputs).
-
-The MEME Suite is a series of commandline tools which provide a variety of
-features for motif analysis such as *de-novo* motif discovery, motif enrichment
-testing, motif comparison, and motif matching, among others \cite{meme-suite}.
-MEME Suite tools take as input fasta files of sequences which will be searched
-for motifs, and if necessary, .meme format motif files containing motifs to use
-in the search.
-
-At its most basic form, `memes` acts as a wrapper of these functions: users can
-point `memes` tools to files on disk, `memes` will construct the commandline
-call, output files will be saved on disk and imported into R for further analysis.
-However, `memes` is more than just a simple wrapper of the MEME Suite, it is an
-analysis framework allowing seamless integration of MEME Suite tools by
-providing an abstraction layer which handles file input/output (I/O) to allow
-inputs and outputs in the form of in-memory R objects.
-
-When running a MEME Suite tool using `memes`, inputs are allowed in multiple
-formats providing the user additional flexibility when performing motif analysis
-(see "Core Utilities" for details). Of note, most outputs of `memes` tools can
-be used as inputs to other `memes` tools, allowing easy construction of
-pipelines (for example, discovering *de-novo* motifs with DREME, then finding
-the genomic positions of those motifs in target regions using FIMO).
- - unlike current MEME tools (ex tomtom -> FIMO is hard to do)
-
-By shifting the focus from managing files on disk which may change infrequently to R data
-objects, this framework reinforces the requirement of writing reproducible
-analysis code to generate reproducible results.
-
-Whenever possible, `memes` uses standard R data types for better integration
-with common analysis tools, like `dplyr`, `GenomicRanges`, or base R. In this
-way, users don't need to learn new ways of working with data, and outputs can
-be integrated with other tools with limited effort.
-
-Furthermore, `universalmotif` conversion tools allow `memes` detected motifs to
-be used in many other popular Bioconductor motif packages, such as `TFBSTools`
-and `motifStack`, providing the user with maximum flexibility in their analysis.
-
-### Discriminative analysis
-A powerful feature of several MEME Suite Tools is the ability to perform
-discriminative analysis, where motifs are identified or tested for enrichment
-in one group of sequences relative to another. In many instances, running tools
-in discriminative mode on a set of control sequences outperforms simply using
-shuffled input sequences as the control (back this up, change wording? maybe:
-using biologically relevant control sequences can improve quality of results?).
-Therefore, enabling complex discriminative analysis designs was central to the
-design of `memes`.
-
-When using list input to `memes` tools, names of list object members can be
-passed to the `control` argument. This results in running the tool in
-discriminative mode using those sequences as the control sequences for all other
-list members. By passing multiple names to `control`, the sequences are combined
-together to create a merged background set. In this way, performing data-driven
-discriminative analysis is just as simple as one using shuffled sequences as
-background.
-
-```{r, eval=F}
-############
-# Replace this for a real set of peaks 
-sequences_list <- list("Static" = "seq",
-                       "Increasing" = "seq",
-                       "Decreasing" = "seq")
-############
-
-# run Dreme on Static, Increasing, Decreasing sites using shuffled input as background
-runDreme(sequences_list, "shuffle")
-
-# run Dreme on Increasing, Decreasing sites using Static sites as background
-runDreme(sequences_list, "Static")
-
-# run Dreme on Decreasing sites using a combined set of Static + Increasing sites as background
-runDreme(sequences_list, c("Static", "Increasing"))
-```
-
-## Metadata manipulation
-`memes` is built on top of the `universalmotif` framework, which provides
-methods for importing motifs from various file formats into a unified R data
-representation complete with metadata information \cite{universalmotif}.
-However, a major limitation of the `universalmotif` framework is that 
-manipulating or filtering multiple `universalmotif` entries requires potentially
-unfamiliar overhead for less experienced users, acting as a barrier to
-performing potentially useful operations.
-
-The `as_universalmotif_dataframe()` function from `memes` converts
-`universalmotif` objects into a more familiar `data.frame` format which can be
-manipulated using `tidyverse` or base R tools. These data.frames also contain a
-special `motif` column which stores the `universalmotif` object of each entry.
-Users can update columns in the data.frame which will be reflected in the
-corresponding metadata entries of the `motif` by running `update_motifs()` after
-manipulating the columns. These data.frames can be converted back into
-`universalmotif` format using `as_universalmotif`, which also updates the
-metadata using column values. The results from running the core MEME Suite
-functions are stored in this format whenever appropriate, creating a consistent
-framework for users to explore and manipulate motifs and their associated data values.
-
-**This section is maybe too listy and could be fleshed out more / integrated above better**
-Examples of common manipulations to motif databases which are aided by this
-format include bulk renaming of entries, identification and removal of redundant
-entries, and exploratory analysis of motif databases.
-
-```{r}
-# Example of renaming motifs w/ universalmotif syntax
-library(universalmotif)
-m1 <- create_motif()
-m2 <- create_motif()
-my_motifs <- list(m1,m2)
-
-my_names <- c("motif one", "motif two")
-
-for (i in 1:length(my_motifs)) {
-  my_motifs[[i]]@name <- my_names[i]
-}
-
-# Example of renaming motifs w/ memes syntax
-my_motifs <- list(m1,m2)
-my_motifs %<>% 
-  as_universalmotif_dataframe() %>% 
-  dplyr::mutate(name = my_names) %>% 
-  as_universalmotif()
-```
-
-This data format also allows users to integrate external data sources into their
-analyses, for example, to filter a motif database to include only motifs for
-genes which are expressed in their tissue of interest. Using motif databases
-filtered in this way has been demonstrated in multiple contexts to provide
-higher quality results when searching for novel regulators based on motif
-content \cite{hector franco, eric olson, john stamm?, shep & greenleaf, klu worm
-paper}.
-
-## Data Visualization (section unfinished)
-
-The MEME Suite provides a limited set of data visualizations which have limited
+## Data visualization
+The MEME Suite provides a small set of data visualizations which have limited
 customizability. `memes` leverages the advantages of the R graphics environment
@@
 
-  - `view_tomtom_hits()` allows examination of tomtom results, allowing the user
-  to evaluate the quality of the assigned match.
+ - `view_motifs()` plots motif as stack (if I decide to add this feature... need to think about this)
+ - `ame_plot_heatmap` plot clustered heatmap of AME results 
+   - `ame_compare_heatmap_methods` compare nonlinearity of data visualization in heatmap form
+ - `view_tomtom_hits` visualize and compare TomTom matches to assess match quality
+   - `force_best_match` can be used to update "best" match demonstrating how data visualization can be useful for data curation
 
-```{r}
-# Example comparison of view_tomtom_hits results
-plot_topHits <- example_tomtom %>% 
-  dplyr::mutate(name = "Example Motif (Has Matches)") %>% 
-  update_motifs() %>% 
-  view_tomtom_hits(2)
+The AME tool searches for enrichment of motifs within groups of sequences
+relative to a control set. The meaningful result from this tool is the
+statistical parameter (for example, a p-value) associated with the significance
+of motif enrichment. However, AME does not provide a mechanism for visualizing
+these results. Additionally, AME is often used 
+## RETURN
 
-plot_noMatch <- create_motif() %>% 
-  as_universalmotif_dataframe() %>% 
-  dplyr::mutate(name = "Example Motif (No Matches)") %>% 
-  update_motifs() %>% 
-  {
-    x <- .
-    x$tomtom <- list(NULL)
-    return(x)
-  } %>% 
-  view_tomtom_hits()
-
-cowplot::plot_grid(plotlist = list(plot_topHits[[1]], plot_noMatch[[1]]), ncol = 2, labels = "AUTO")
-```
-
-  - `ame_plot_heatmap()` allows comparison of AME results between multiple
-  groups. Heirarchical ordering of hits shows commonalities and differences in enrichment between groups.
-```{r}
-# AME Heatmap example plots
-example_ame$Decreasing %>% 
-  ame_plot_heatmap()
-
-example_ame %>% 
-  dplyr::bind_rows(.id = "id") %>% 
-  ame_plot_heatmap(group = id)
-```
-
-# Conclusion
- - `memes` is cool
- - use `memes`
- - it's on github (provide link)
- - manual is here:
-
-# Notes
-
-**Example of commandline vs `memes`?**
-For example, to run DREME on the commandline on "input.fasta", while setting the random seed to 100:
-```{bash, eval=F}
-# Shell command setting random seed to 100
-dreme -s 100 input.fasta
-```
-
-The `memes` implementation of the above command would be written as follows:
-```{r, eval=F}
-# memes command setting random seed to 100
-# user explicitly states they want to shuffle input sequences as control
-runDreme("input.fasta", control = "shuffle", s = 100)
-```
-
-### Motif Comparison (TomTom)
- - takes runDreme/meme output as input -> pipes
- - enables using best matched motif as input to other tools (unlike TomTom on commandline)
-
-### Motif Matching (FIMO)
- - returns locations as GRanges
- - integrates well with `plyranges` to facilitate range-based join operations to integrate peak-level metadata with motif information, vice-versa
-
-\begin{figure}
-\includegraphics[width=\textwidth]{figures/memes_model.png}
-\caption{A diagram of memes tools & workflows?
-flowchart of each step, getsequence-> run-> import -> visualize -> goto run
-
-Diagram how meme-suite tools work vs memes?
-}
-\label{fig:model} 
-\end{figure}
+## Containerized analysis maximizes availability and facilitates reproducibility
+`memes` relies on a locally installed version of the MEME Suite which is
+accessed by the users R process. Although R is available on Windows, Mac, and
+Linux operating systems, the MEME Suite is incompatible with Windows, limiting
+its adoption by Windows users. Additional barriers also exist to installing a
+local copy of the MEME Suite on compatible systems. The MEME Suite relies on
+several system-level dependencies whose installation can be difficult for novice
+users. Finally, some tools in the MEME Suite use python to generate shuffled
+control sequences for analysis, which presents a reproducibility issue as the
+random number generation algorithm changed from python2.7 to python3. The MEME
+Suite will build and install on both python2.7 and python3 systems, therefore
+without careful consideration of this behavior, the same code run on two systems
+may not produce identical results, even if using the same major version of the
+MEME Suite. In order to democratize access to the MEME Suite on unsupported
+operating systems, and to facilitate reproducibile motif analysis using `memes`, 
+we have also developed docker and singularity (**NOTE:** singularity is broken
+for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
+preinstalled version of the MEME Suite along with an R/Bioconductor analysis
+environment including the most recent version of `memes` and its dependencies.
 
@@
  - Additional meme suite tools
+ - more data visualizations
  - better support for PSP generation
  - Better support for amino-acid sequence usage (please contact us with feature requests)
+
+# Acknowledgements
+ - Sean Davis automated docker builds
+ - Megan Justice, feedback, feature requests

### 2020-09-20 manuscript.Rmd
add availability section

@@
 
-# Future developments:
+# Results
+
+# Availability and Future developments:
+ - packge source code: `github.com/snystrom/memes`
+ - documentation website: `snystrom.github.io/memes`
+ - docker container: `https://hub.docker.com/r/snystrom/memes_docker`
+ - singularity container: `???`
+
+### Future Features:
  - Additional meme suite tools

### 2020-09-20 R/get_raw_data.R
Add drake plan & raw data for Results section.

@@ -1,0 +1,29 @@
+# Downloads data from GEO if not in data directory
+
+rna_link <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE97956&format=file&file=GSE97956%5FgeneFPKMs%5FL3%5F24hr%5F44hr%2Exlsx"
+chip_link <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE141738&format=file&file=GSE141738%5Funion%5Fchip%5Fpeaks%5Fannotated%2Ecsv%2Egz"
+
+targets <- list("data/rna_fpkm.xlsx" = rna_link,
+                "data/e93_chip_peaks.csv.gz" = chip_link
+               )
+
+download_geo <- function(url, destfile){
+  curl::curl_download(url, destfile)
+  return(destfile)
+}
+
+download_missing_data <- function(link, path){
+  if (!file.exists(path)) {
+    download_geo(link, path)
+  } 
+}
+
+purrr::iwalk(targets, download_missing_data)
+
+
+## Copy flyFactorSurvey .meme format file
+flyFactorDB <- "data/flyFactorSurveyCleaned.meme" 
+if (!file.exists(flyFactorDB)) {
+  pkg_flyFactor <- system.file("extdata/flyFactorSurvey_cleaned.meme", package = "memes")
+  file.copy(from = ffs, to = flyFactorDB)
+}

### 2020-09-20 R/plan.R

@@ -1,0 +1,34 @@
+
+my_plan <- drake_plan(
+  
+  chip_path = file_in("data/e93_chip_peaks.csv.gz"),
+  rna_path = file_in("data/rna_fpkm.xlsx"),
+  
+  full_rnaseq = readxl::read_excel(rna_path) %>%
+    tidyr::pivot_longer(-matches("gene_symbol"), names_to = "time", values_to = "fpkm") %>%
+    dplyr::filter(time != "44hr") %>%
+    dplyr::mutate(time = dplyr::case_when(time == "L3" ~ "Early",
+                                          time == "24hr" ~ "Late")) %>%
+    dplyr::rename(symbol = gene_symbol) %>%
+    dplyr::group_by(symbol) %>%
+    dplyr::filter(max(fpkm) != 0),
+  
+  chip_peaks = readr::read_csv(chip_path, col_types = c("chr" = "c",
+                                                        "start" = "d",
+                                                        "end" = "d",
+                                                        "strand" = "c",
+                                                        "id" = "c",
+                                                        "summit_position" = "d",
+                                                        "peak_binding_description" = "c",
+                                                        "e93_dependent" = "c",
+                                                        "e93_sensitive" = "c",
+                                                        "e93_sensitive_behavior" = "c",
+                                                        "wildtype_3lw_24apf_behavior" = "c")),
+  chip_summits = chip_peaks %>% 
+    GenomicRanges::GRanges() %>% 
+    plyranges::anchor_start() %>% 
+    plyranges::mutate(width = 1) %>% 
+    plyranges::shift_right(.$summit_position) %>% 
+    select(id, peak_binding_description, e93_sensitive_behavior)
+)
+  

### 2020-09-20 _drake.R

@@ -1,0 +1,6 @@
+library(drake)
+library(magrittr)
+source("R/get_raw_data.R")
+source("R/plan.R")
+
+drake_config(my_plan, verbose = 1L)

### 2020-09-21 R/plan.R
update plan & raw data

@@ -13,6 +13,14 @@ my_plan <- drake_plan(
     dplyr::group_by(symbol) %>%
     dplyr::filter(max(fpkm) != 0),
   
+  filtered_rnaseq = full_rnaseq %>% 
+    dplyr::group_by(symbol) %>% 
+    dplyr::filter(max(fpkm) > 5) %>% 
+    dplyr::select(symbol) %>% 
+    unique(),
+  
+  write_parsed_rna = readr::write_csv(filtered_rnaseq, path = file_out("data/wing_expressed_genes.csv")),
+  
   chip_peaks = readr::read_csv(chip_path, col_types = c("chr" = "c",
                                                         "start" = "d",
                                                         "end" = "d",
@@ -25,10 +33,25 @@ my_plan <- drake_plan(
                                                         "e93_sensitive_behavior" = "c",
                                                         "wildtype_3lw_24apf_behavior" = "c")),
   chip_summits = chip_peaks %>% 
+    dplyr::rename("e93_chromatin_response" = "e93_sensitive_behavior") %>% 
+    dplyr::filter(!is.na(e93_chromatin_response)) %>% 
     GenomicRanges::GRanges() %>% 
     plyranges::anchor_start() %>% 
     plyranges::mutate(width = 1) %>% 
     plyranges::shift_right(.$summit_position) %>% 
-    select(id, peak_binding_description, e93_sensitive_behavior)
+    plyranges::select(id, e93_chromatin_response),
+  
+  chip_results = chip_summits %>% 
+    plyranges::anchor_center() %>% 
+    plyranges::mutate(width = 100)
+  
+  #genome = BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3,
+  #
+  #all_seq = chip_results %>% 
+  #  memes::get_sequence(genome),
+  #
+  #all_dreme_res_full = memes::runDreme(all_seq, control = "shuffle", nmotifs = 3)
+  
 )
   
+

### 2020-09-21 manuscript.Rmd
starting Results section

@@
 ```{r setup}
-library(dremeR)
+library(memes)
 library(magrittr)
@@
 
+Here we briefly highlight each of `memes` current features for analyzing
+ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
+tool, and a complete example using `memes` to analyze ChIP-seq data can be found
+in the `memes` vignette and the package website
+([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
+example, we reanalyze recent work examining the causal relationship between the
+binding of the transcription factor E93 to changes in chromatin accessibility
+during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
+utilize ChIP-seq peak calls for E93 which have been annotated according to the
+change in chromatin accessibility observed before & after E93 binding. These
+data are an ideomatic example of range-based genomic data (E93 ChIP peaks)
+containing additional groupings (the chromatin accessibility response following
+DNA binding) whose membership may be influenced by differential motif content.
+We show how `memes` syntax allows complex analysis designs, how `memes`
+utilities enable deep interrogation of results, and how `memes` flexible data
+structures empower users to integrate the `memes` workflow with tools offered by
+other R/Bioconductor packages.
+
+### RETURN HERE
+### NOTE TO SELF:
+ - This is too much of a tutorial & vignette.
+ - move this text to the vignette since there isn't any.
+ - Instead, highlight some of the interface useage like so:
+
+To run dreme, where `peaks` are chip peaks below with a metadata column `e93_respnose`:
+```{r}
+seq <- peaks %>% split(mcols(.)$response) %>% get_sequence
+runDreme(seq, control = "static")
+```
+
+----------------
+```{r, include=FALSE}
+drake::loadd(chip_results)
+```
+
+We begin with a `GenomicRanges` object of E93 ChIP-peak summits +/- 100 bp with
+a metadata column indicating the change in chromatin accessibility associated
+with E93 binding at that region. 
+```{r}
+chip_results %>% head
+```
+
+### Denovo motif discovery using DREME
+
+One approach to identify *de-novo* motifs within ChIP peaks is to use DREME,
+which performs discriminative motif detection in an input set of sequences
+relative to a control set \cite{dreme}. To search for *de-novo* motifs within
+E93 bindings sites using DREME, we first use the `get_sequence()` function which
+returns the DNA sequence within genomic coordinates.
+```{r}
+genome <- BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3
+
+all_seq <- chip_results %>% 
+  get_sequence(genome)
+```
+
+The output is stored as a `DNAStringSet`, shown below.
+```{r}
+head(all_seq, 3)
+```
+
+The `runDreme` command is used to begin a motif search using DREME. Although it
+is possible to use any number of input sequences, runtimes quickly become
+infeasible as sequences exceed 100000 total bp. This can be overcome in a
+variety of ways: users can rank peaks by ChIP-seq signal, or some other
+quantitative value, and use a subset selected by rank, or in the absence of any
+quantitative measure, users can randomly sample their sequences to use as input.
+For simplicity in this example, we subset our data to 1000 randomly selected
+peaks. Setting `control = "shuffle"` uses DREME's built-in dinucleotide
+shuffling algorithm to produce a control set of seqeunces.
+
+```{r}
+set.seed(123) # setting a random seed ensures sampling produces a reproducible subsample
+sequence_subset <- sample(all_seq, 1000)
+all_dreme_res <- runDreme(sequence_subset, control = "shuffle", nmotifs = 3)
+```
+
+```{r}
+all_dreme_res
+```
+
+### Preparing a motif database for searching
+
+Transcription factors often bind with sequence specificity to regulate activity
+of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
+DNA binding motifs for transcription factors can be a key step in identifying
+regulators of genomic elements.
+
+The TomTom tool is used to compare unknown motifs against a list of known motifs
+to identify matches. By passing the results from DREME into TomTom and searching
+within a database of *Drosophila* transcription factor motifs, we can identify
+candidate transcription factors which may also bind at E93 ChIP peaks. A major
+limitation of this approach is that transcription factors containing similar
+families of DNA binding domain often possess highly similar motifs, making it
+difficult to identify the "true" binding factor of a discovered motif.
+Additionally, when searching for matches against a motif database, TomTom must
+account for multiple testing, therefore using a larger than necessary motif
+database can produce a large multiple testing penalty, limiting sensititity of
+detection. One way to overcome these limitations is to limit the transcription
+factor motif database to include only motifs for transcription factors expressed
+in the sample of interest. Accounting for transcription factor expression during
+motif analysis has been demonstrated to  increase the probability of identifying
+biologically relevant transcription factor candidates \cite{hector franco's STAR paper, others?}.
+
+The `memes` universalmotif data.frame structure can be used to integrate
+expression data with a motif database. First, we import a pre-filtered list of
+genes expressed in a timecourse of **Drosophila** wing development.
+
+```{r}
+rna_counts <- read.csv("data/wing_expressed_genes.csv", stringsAsFactors = FALSE)
+```
+```{r, echo = FALSE}
+head(rna_counts)
+```
+
+Next we import a database of known *Drosophila* transcription factor motifs
+generated from the Fly Factor Survey in `universalmotif` format, and convert to a data.frame format using `as_universalmotif_dataframe()` \cite{fly factor}.
+```{r}
+flyFactor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
+  as_universalmotif_dataframe()
+```
+```{r, echo = FALSE}
+head(flyFactor, 5)
+```
+
+```{r}
+tomtom_results <- runTomTom(all_dreme_res, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
+```
+Here we visualize the top 3 matches from TomTom for each *de-novo* motif using
+`view_tomtom_hits()`. Results are plotted in descending order according to match
+score. The results of this function can be integrated into additional R graphics
+libraries, such as `patchwork` or `cowplot` to generate publication-quality
+figures \cite{patchwork, cowplot}.
+```{r, fig.height=5, fig.width=12}
+tomtom_results %>% 
+  view_tomtom_hits(3) %>% 
+  patchwork::wrap_plots(., nrow = 1, tag_level = "new") +
+    patchwork::plot_annotation(tag_levels = "A")
+```
+## Replace above chunks with sections below 
+(rstudio crashed without saving)
+--------------
+Next we import a database of known *Drosophila* transcription factor motifs
+generated from the Fly Factor Survey in `universalmotif` format, and convert to a data.frame format using `as_universalmotif_dataframe()` \cite{fly factor}.
+```{r}
+fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
+				  as_universalmotif_dataframe()
+	```
+	```{r, echo = FALSE}
+	head(fly_factor, 5)
+	```
+
+The motif database can be subsequently filtered to include only those entries
+which match an expressed gene. After filtering, the database is converted back
+to `universalmotif` format using `as_universalmotif`.
+```{r}
+# in this database, altname stores the gene symbol
+fly_factor_expressed <- fly_factor %>% 
+				  dplyr::filter(altname %in% expressed_genes$symbol) %>% 
+					  as_universalmotif()
+		```
+
+Passing the results object from *de-novo* discovery with `runDreme` and the
+filtered motif database to `runTomTom`
+```{r}
+tomtom_results <- runTomTom(all_dreme_res, database = fly_factor_expressed, dist = "ed")
+tomtom_results <- runTomTom(all_dreme_res, database = fly_factor, dist = "ed")
+```
+
+----------------
+
+ - prepare database
+  - filter FPKM counts, note this needs to be chosen empirically and is not always a good idea,
+  - we do this for demonstration purposes only
+ - Denovo discovery in sensitivity categories (static as background)
+  - TOMTOM matching, view_match_motifs(3)
+    - try patchwork to combine them into figure
+ - AME Across binding groups
+  - reduce redundant hits
+  - plot heatmap
+ - FIMO within sensitivity
+  - rederive PWM for E93 within sites
+   - note differences in specific bases
+
 # Availability and Future developments:

### 2020-09-21 runDocker.sh
added runDocker.sh to launch the dev environment

@@ -1,0 +1,1 @@
+sudo docker run -p 8787:8787 -e PASSWORD=$1 -v /mnt/c/Users/Spencer/pkg_dev/memes_paper:/mnt/memes_paper snystrom/memes_docker:devel

### 2020-09-22 R/plan.R
Update plan.

@@ -13,6 +13,7 @@ my_plan <- drake_plan(
     dplyr::group_by(symbol) %>%
     dplyr::filter(max(fpkm) != 0),
   
+  
   filtered_rnaseq = full_rnaseq %>% 
     dplyr::group_by(symbol) %>% 
     dplyr::filter(max(fpkm) > 5) %>% 
@@ -20,6 +21,12 @@ my_plan <- drake_plan(
     unique(),
   
   write_parsed_rna = readr::write_csv(filtered_rnaseq, path = file_out("data/wing_expressed_genes.csv")),
+  
+  max_expression_rnaseq = full_rnaseq %>% 
+    dplyr::group_by(symbol) %>% 
+    dplyr::filter(fpkm == max(fpkm)) %>% 
+    dplyr::select(-time) %>% 
+    dplyr::ungroup(),
   
   chip_peaks = readr::read_csv(chip_path, col_types = c("chr" = "c",
                                                         "start" = "d",
@@ -41,14 +48,20 @@ my_plan <- drake_plan(
     plyranges::shift_right(.$summit_position) %>% 
     plyranges::select(id, e93_chromatin_response),
   
+  # ChIP summits are extended to 100bp regions for analysis
   chip_results = chip_summits %>% 
     plyranges::anchor_center() %>% 
-    plyranges::mutate(width = 100)
+    plyranges::mutate(width = 100),
   
-  #genome = BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3,
+  
+  genome = BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3,
+  
+  sequences_by_response = chip_results %>% 
+    split(mcols(.)$e93_chromatin_response) %>% 
+    memes::get_sequence(genome),
+  
+  dreme_vs_static = runDreme(sequences_by_response, control = "Static"),
   #
-  #all_seq = chip_results %>% 
-  #  memes::get_sequence(genome),
   #
   #all_dreme_res_full = memes::runDreme(all_seq, control = "shuffle", nmotifs = 3)
   

### 2020-09-22 _drake.R

@@ -1,5 +1,7 @@
 library(drake)
 library(magrittr)
+library(GenomicRanges)
+library(memes)
 source("R/get_raw_data.R")
 source("R/plan.R")
 

### 2020-09-22 manuscript.Rmd
Almost final draft

--- 2020-09-20 manuscript.Rmd
@@
 - name: Daniel J. McKay
-output: word_document
+output: 
+ - word_document:
+   fig.caption: yes
 ---
 
-```{r setup}
-library(dremeR)
-library(magrittr)
+```{r setup, echo = FALSE}
+# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
+suppressPackageStartupMessages(library(memes))
+suppressPackageStartupMessages(library(magrittr))
+suppressPackageStartupMessages(library(GenomicRanges))
+suppressPackageStartupMessages(library(ggplot2))
+drake::r_make()
 ```
@@
 an ideal representation for these data is one which can hold an unlimited amount
-of metadata, contain complex heirarchies, and be easily manipulated using
+of metadata, contain complex hierarchies, and be easily manipulated using
 standard analysis tools. `memes` therefore extends the `universalmotif` format
@@
 are enriched in one category relative to another. Although the MEME Suite allows
-differential enrichment testing, it does not inheritly provide a mechanism for
+differential enrichment testing, it does not inherently provide a mechanism for
 analyzing groups of sequences in parallel, or performing motif analysis with any
@@
 MEME Suite. In order to democratize access to the MEME Suite on unsupported
-operating systems, and to facilitate reproducibile motif analysis using `memes`, 
+operating systems, and to facilitate reproducible motif analysis using `memes`, 
 we have also developed docker and singularity (**NOTE:** singularity is broken
@@
 environment including the most recent version of `memes` and its dependencies.
+As new container versions are released, they are version tagged and stored to
+ensure reproducibility while allowing updates to the container.
 
-# Future developments:
- - Additional meme suite tools
- - more data visualizations
- - better support for PSP generation
- - Better support for amino-acid sequence usage (please contact us with feature requests)
+# Results
+
+Here we briefly highlight each of `memes` current features for analyzing
+ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
+tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
+in the `memes` vignette and the package website
+([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
+example, we reanalyze recent work examining the causal relationship between the
+binding of the transcription factor E93 to changes in chromatin accessibility
+during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
+utilize ChIP-seq peak calls for E93 which have been annotated according to the
+change in chromatin accessibility observed before & after E93 binding. These
+data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
+containing additional groupings (the chromatin accessibility response following
+DNA binding) whose membership may be influenced by differential motif content.
+We show how `memes` syntax allows complex analysis designs, how `memes`
+utilities enable deep interrogation of results, and how `memes` flexible data
+structures empower users to integrate the `memes` workflow with tools offered by
+other R/Bioconductor packages.
+
+```{r, echo = F}
+drake::loadd(chip_results)
+```
+
+The aforementioned ChIP-seq peaks are stored as a GRanges object with a metadata
+column (`e93_chromatin_response`) indicating whether chromatin accessibility
+tends to increase ("Increasing"), decrease ("Decreasing"), or remain unchanged
+("Static") following E93 binding.
+```{r}
+head(chip_results, 3)
+```
+
+Using the `get_sequence` function, GRanges objects are converted into
+DNAStringSet outputs.
+```{r}
+dm.genome <- BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3
+
+all_sequences <- chip_results %>% 
+  get_sequence(dm.genome)
+```
+
+```{r}
+head(all_sequences)
+```
+In order to perform analysis within different groups of peaks, the GRanges object can be `split` on a metadata column before input to `get_sequence`.
+```{r}
+sequences_by_response <- chip_results %>% 
+  split(mcols(.)$e93_chromatin_response) %>% 
+  get_sequence(dm.genome)
+
+```
+
+This produces in a BStringSetList where list members contain a DNAStringSet for each group of sequences.
+```{r}
+head(sequences_by_response)
+```
+
+## Denovo motif analysis
+The DREME tool can be used to discover short, novel motifs present in a set of
+input sequences relative to a control set. The `runDreme` command is the `memes`
+interface to the DREME tool. `runDreme` syntax enables users to produce complex
+discriminative analysis designs using intuitive syntax. Examples of possible
+designs and their syntax are compared below.
+
+```{r, eval=F}
+# use all sequences vs shuffled background
+runDreme(all_sequences, control = "shuffle")
+
+# for each response category, discover motifs against a shuffled background set
+# Produces:
+#  - Increasing vs Shuffled Increasing
+#  - Decreasing vs Shuffled Decreasing
+#  - Static vs Shuffled Static
+runDreme(sequences_by_response, control = "shuffle")
+
+# Search for motifs enriched in the "Increasing" peaks relative to "Decreasing" peaks
+# Produces:
+#  - Increasing vs Decreasing
+runDreme(sequences_by_response$Increasing, control = sequences_by_response$Decreasing)
+
+# Use the "Static" response category as the control set to discover motifs 
+# enriched in each remaining category relative to static sites,
+# Produces:
+#  - Increasing vs Static
+#  - Decreasing vs Static
+runDreme(sequences_by_response, control = "Static")
+
+# Combine the "Static" and "Increasing" sequences and use as a background set to
+# discover motifs enriched in the remaining categories relative
+# Produces:
+#  - Decreasing vs Static+Increasing
+runDreme(sequences_by_response, control = c("Static", "Increasing"))
+```
+
+In order to discover motifs associated with dynamic changes to chromatin
+accessibility, we use static sites as the control set to discover motifs
+enriched in both increasing and decreasing sites.
+
+```{r, eval=F}
+dreme_vs_static <- runDreme(sequences_by_response, control = "Static")
+```
+
+```{r, echo=FALSE}
+# I secretly cache this to decrease build-time of the manuscript
+drake::loadd(dreme_vs_static)
+```
+
+```{r, fig.height=4, fig.width=7, echo = FALSE}
+dreme_fig_top <- purrr::map(dreme_vs_static, "motif") %>% 
+  purrr::map(universalmotif::view_motifs) %>% 
+  purrr::imap(~{
+    .x +
+      ggtitle(paste(.y, "Motifs")) +
+      theme(plot.title = element_text(hjust = 0.5, size = 14))
+  }) %>% 
+  patchwork::wrap_plots(ncol = 2) 
+```
+```{r, echo=FALSE}
+
+# Require that the minimum overlap for similarity metric is equal to the
+# smallest possible overlap if the shortest motif was completely contained
+# within the largest motif.
+get_min_overlap <- function(df){
+  min(df$length) / max(df$length)
+}
+
+get_motif_correlation <- function(df, method = "PCC"){
+  df$motif %>% 
+    universalmotif::compare_motifs(min.overlap = get_min_overlap(df), method = method)
+}
+
+plot_correlation_heatmap <- function(mat, name){
+
+  min_color <- round(min(mat) - 0.1, digits = 1)
+  colors <- circlize::colorRamp2(c(min_color, 1), c("white", "dodgerblue3"))
+
+  rownames(mat) <- NULL
+  ComplexHeatmap::Heatmap(mat,
+                          col = colors, 
+                          name = "Pearson Correlation", 
+                          column_names_rot = 45,
+                          column_title = paste(name, "Motif Similarity")
+                          )
+
+}
+```
+
+```{r, fig.height=5, fig.width=10, echo = FALSE}
+dreme_fig_bottom <- dreme_vs_static %>% 
+  purrr::map(get_motif_correlation) %>% 
+  purrr::imap(plot_correlation_heatmap) %>% 
+  # capture heatmaps as grobs
+  purrr::map(~{grid::grid.grabExpr(draw(.x))}) %>% 
+  patchwork::wrap_plots(nrow = 1) 
+```
+
+These data are readily visualized using the `universalmotif::view_motifs`
+function. Visualization of the results reveals two distinct motifs associated
+with decreasing and increasing sites \ref{fig:dec_v_inc_dreme}. In this
+analysis, the *de-novo* discovered motifs within each category appear visually
+similar to each other, however, the MEME Suite does not provide a mechanism to
+compare groups of motifs based on all pairwise similarity metrics. We utilized
+the `universalmotif::compare_motifs` function to compute Pearson correlation
+coefficients for each set of motifs to quantitatively assess motif similarity
+\ref{fig:dec_v_inc_dreme}.
+
+```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites.
+C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
+patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
+  patchwork::plot_annotation(tag_levels = "A")
+```
+Transcription factors often bind with sequence specificity to regulate activity
+of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
+DNA binding motifs for transcription factors can be a key step in identifying
+regulators of genomic elements. The TomTom tool is used to compare unknown
+motifs against a list of known motifs to identify matches, `memes` provides the
+`runTomTom` function as an interface to this tool. By passing the results from
+`runDreme` into `runTomTom` and searching within a database of *Drosophila* transcription
+factor motifs, we can identify candidate transcription factors which may bind
+the motifs associated with increasing and decreasing chromatin accessibility.
+
+```{r}
+dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
+```
+Using this approach, the results can be visualized using 
+the `view_tomtom_hits()` function to visually inspect the matches
+assigned by TomTom, providing a simple way to assess the quality of the match. 
+A representative plot of these data is shown in \ref{fig:tomtom_example}, while
+the full results can be viewed in \ref{tab:tomtom_table}. We
+uncover that the motifs associated with decreasing sites are highly similar to
+the E93 motif. Conversely, motifs found in increasing sites match the
+transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
+data support the hypothesis that E93 binding to its motif may play a larger role
+in chromatin closing activity, while binding of the transcription factor br is
+associated with chromatin opening.
+```{r tomtom_example, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
+plot_top_hit <- function(x){
+  x %>% 
+    view_tomtom_hits(1)
+}
+
+top_hit_grid <- function(x, title){
+  x %>% 
+    patchwork::wrap_plots(nrow = 1)
+}
+
+# Plotting every hit:
+# decided to just use a representative sample insted
+#dreme_vs_static_tomtom %>% 
+#  purrr::imap(~{plot_top_hit(.x) %>% top_hit_grid(., .y)}) %>% 
+#  patchwork::wrap_plots(ncol = 1)
+
+dreme_vs_static_tomtom %>% 
+  purrr::imap(~{
+    .x %>% 
+      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
+      update_motifs() %>% 
+      plot_top_hit() %>% 
+      .[[2]]
+    }) %>% 
+  patchwork::wrap_plots() +
+    patchwork::plot_annotation(tag_levels = "A")
+```
+```{r}
+library(gt)
+
+dreme_vs_static_tomtom %>% 
+  dplyr::bind_rows(.id = "category") %>% 
+  dplyr::mutate(category = paste(category, "Motifs")) %>% 
+  dplyr::arrange(category, rank) %>% 
+  dplyr::select(category, name, best_match_altname) %>% 
+  dplyr::rename("Best Match TF" = best_match_altname) %>% 
+  gt(rowname_col = "name",
+     groupname_col = "category",
+     lab = "tab:tomtom_table")
+
+```
+
+## Motif enrichment testing using runAme
+
+Discovery and matching of *de-novo* motifs is only one way to find candidate
+transcription factors within target sites. Indeed, in many instances, requiring
+that a motif is recovered *de-novo* is not ideal, as these approaches are
+"discovery based" as opposed to hypothesis driven. Another approach, implemented
+by the AME tool, is to search for known motif instances in target sequences and
+test for their overrepresentation relative to a background set of sequences
+\cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
+accepts a set of sequences as input an control sets, and will peform enrichment
+testing using a provided motif database for occurrances of each provided motif.
+
+A major limitation of this approach is that transcription factors containing
+similar families of DNA binding domain often possess highly similar motifs,
+making it difficult to identify the "true" binding factor associated with an
+overrepresented motif. Additionally, when searching for matches against a motif
+database, AME must account for multiple testing, therefore using a larger than
+necessary motif database can produce a large multiple testing penalty, limiting
+sensitivity of detection. One way to overcome these limitations is to limit the
+transcription factor motif database to include only motifs for transcription
+factors expressed in the sample of interest. Accounting for transcription factor
+expression during motif analysis has been demonstrated to increase the
+probability of identifying biologically relevant transcription factor candidates
+\cite{hector franco's STAR paper, others?}.
+
+The `memes` universalmotif data.frame structure can be used to integrate
+expression data with a motif database to remove entries for transcription
+factors which are not expressed. To do so, we import a *Drosophila*
+transcription factor motif database generated by the Fly Factor Survey and
+convert to universalmotif data.frame format \cite{fly factor}. In this database,
+the `altname` column stores the gene symbol.
+
+```{r}
+fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
+  as_universalmotif_dataframe()
+```
+```{r, echo = FALSE}
+head(fly_factor, 3)
+```
+Next, we import a pre-filtered list of genes expressed in a timecourse of
+**Drosophila** wing development.
+```{r}
+wing_expressed_genes <- read.csv("data/wing_expressed_genes.csv")
+```
+
+```{r, echo = FALSE}
+n_fly_factor_expressed <- fly_factor %>% 
+  dplyr::filter((altname %in% wing_expressed_genes$symbol)) %>% 
+  nrow()
+n_fly_factor_unexpressed <- fly_factor %>% 
+  dplyr::filter(!(altname %in% wing_expressed_genes$symbol)) %>% 
+  nrow()
+```
+
+Finally, we subset the motif database to only expressed genes using `dplyr`
+data.frame subsetting syntax (note that base R subsetting functions operate
+equally well on these data structures), then convert the data.frame back into
+`universalmotif` format using `as_universalmotif`. This filtering step removes
+`r (n_fly_factor_unexpressed / nrow(fly_factor)) %>% {. * 100} %>% round()`% of
+entries from the original database, greatly reducing the multiple-testing
+correction.
+
+```{r}
+fly_factor_expressed <- fly_factor %>% 
+  dplyr::filter(altname %in% wing_expressed_genes$symbol) %>% 
+  as_universalmotif()
+```
+
+`runAme` syntax is identical to `runDreme` in that discriminative designs can be
+constructed by calling list entries by name. Data can be visualized using
+`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
+Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
+sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
+Additionally, AME detects several other transcription factor motifs which
+distinguish decreasing and increasing sites, providing additional clues to
+potential factors which also bind E93 ChIP peaks to affect chromatin
+accessibility.
+```{r}
+ame_vs_static <- runAme(sequences_by_response, control = "Static", database = fly_factor_expressed)
+```
+
+```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
+ame_vs_static %>% 
+  dplyr::bind_rows(.id = "category") %>% 
+  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
+  dplyr::group_by(category, motif_alt_id) %>% 
+  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
+  ame_plot_heatmap(group = category, id = motif_alt_id) +
+    labs(x = "Transcription Factor",
+         y = "ChIP Peak Type")
+```
+### Motif matching using runFimo
+
+A striking result from these analysis is that the E93 motif is so strongly
+enriched within E93 ChIP peaks which decrease in accessibility. Significance of
+motif enrichment can be driven by several factors, such as quality of the query
+motif relative to the canonical motif or differences in motif number in one
+group relative to other groups. These questions can be explored directly by
+identifying motif occurrances in target regions and examining their properties.
+FIMO allows users to match motifs in input sequences while returning information
+about the quality of each match in the form of a quantitative score \cite{fimo}.
+
+In order to examine the properties of the E93 motif between different ChIP
+peaks, we scan all E93 ChIP peaks for matches to the Fly Factor Survey E93 motif
+using `runFimo`. Results are returned as a GRanges object containing the
+positions of each motif. Setting `skip_matched_sequence` to `FALSE` adds the
+sequence of the detected motif to the output.
+
+```{r}
+e93_fimo <- runFimo(all_sequences, fly_factor_expressed$Eip93F_SANGER_10, skip_matched_sequence = FALSE, thresh = 1e-3)
+```
+
+Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.
+```{r}
+e93_fimo %<>% 
+  plyranges::join_overlap_intersect(chip_results)
+```
+
+```{r, echo=F}
+e93_stack <- e93_fimo %>% 
+  data.frame %>%  
+  split(.$e93_chromatin_response) %>% 
+  purrr::imap(~{
+    universalmotif::create_motif(.x$matched_sequence, name = paste(.y, "E93"))
+  }) %>% 
+  rev %>% 
+  {
+    m <- .
+    e93_flyfactor <- fly_factor_expressed$Eip93F_SANGER_10
+    e93_flyfactor@name <- "E93 FlyFactorSurvey"
+
+    m$e93 <- e93_flyfactor
+    return(m)
+  } %>% 
+  rev %>% 
+  universalmotif::view_motifs()
+```
+```{r, echo=F}
+n_motifs <- chip_results %>% 
+  plyranges::mutate(has_match = ifelse(. %over% e93_fimo, "Match", "No Match")) %>% 
+  data.frame %>% 
+  ggplot(aes(e93_chromatin_response)) +
+    geom_bar(aes(fill = forcats::fct_rev(has_match)), position = "fill") +
+     scale_fill_manual(values = c("Match" = "firebrick",
+                                  "No Match" = "Black")) +
+     labs(fill = "E93 Motif Match",
+          y = "Fraction of Sites",
+          x = "Response to E93 binding") 
+```
+```{r, fig.height=5, fig.width=3, echo=F}
+motif_quality <- e93_fimo %>% 
+  data.frame %>% 
+  ggplot(aes(e93_chromatin_response, score)) +
+    geom_boxplot(aes(fill = e93_chromatin_response), notch = TRUE) +
+    theme_bw() +
+    labs(y = "FIMO Score",
+         x = "Chromatin Response in E93 ChIP Peak") +
+    theme(legend.position = "none")
+```
+
+Using this approach we can deeply examine the properties of the E93 motif within
+each chromatin response category. First, by counting the number of E93 motifs
+within each category, we demonstrate that Decreasing sites are more likely than
+increasing or static sites to contain an E93 motif \ref{fig:fimo}A. We extend
+these observations by rederiving position-weight matrices from sequences
+matching the E93 motif within each category, allowing visual inspection of motif
+quality across groups \ref{fig:fimo}B. Notably, differences in quality at base
+positions 8-12 appear to distinguish increasing from decreasing motifs, where
+decreasing motifs are more likely to have strong A bases at positions 8 and 9,
+while E93 motifs from increasing sites are more likely to have a T base pair at
+position 12 \ref{fig:fimo}B. Examination of bulk FIMO scores (where higher
+scores represent motifs more similar to the reference) also reveals differences
+in E93 motif quality between groups, in particular, E93 motifs from Decreasing
+sites have higher scores \ref{fig:fimo}C. Together, these data demonstrate that
+a key distinguishing factor between whether a sites will decrease or increase in
+chromatin accessibility following E93 binding is the number and quality of E93
+motifs at that site.
+
+In summary, `memes` establishes a powerful motif analysis environment by
+leveraging the speed and utility of the MEME Suite set of tools in conjunction
+with the flexible and extensive R/Bioconductor package landscape.
+```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A.
+Stacked barplot showing fraction of each chromatin response category containing E93 motif matches.
+B.
+PWMs generated from E93 motif sequences detected in each chromatin response category.
+C.
+Boxplot of FIMO Score for each E93 motif within each chromatin response category.
+Outliers are plotted as distinct points."}
+#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
+patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
+                      e93_stack, motif_quality, 
+                      ncol = 2, widths = c(0.6, 0.4),
+                      guides = 'collect') +
+    patchwork::plot_annotation(tag_levels = "A")
+```
+
+# Availability and Future developments:
+
+The `memes` package source code and installation instructions are available on
+github: [github.com/snystrom/memes](github.com/snystrom/memes). Documentation is
+stored in the package vignettes, and also available at the package website:
+[snystrom.github.io/memes](snystrom.github.io/memes). The `memes_docker`
+containers are available from DockerHub
+[https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
+and their source code is hosted at github:
+[https://github.com/snystrom/memes_docker](https://github.com/snystrom/memes_docker).
+
+This manuscript was automatically generated using Rmarkdown within the
+`memes_docker` container (TODO: ADD VERSION). Its source code, raw data, and
+instructions to reproduce all analysis can be found at
+[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/).
+
+In the future we hope to add additional data visualizations for examining motif
+positioning within features. We will continue to add support for additional MEME Suite
+tools in future versions of the package. Finally, we hope to improve the `memes`
+tooling for analyzing amino-acid motifs, which although fully supported by our
+current framework, may require extra tools which we have not considered.
 
 # Acknowledgements
- - Sean Davis automated docker builds
- - Megan Justice, feedback, feature requests
+We would like to acknowledge Megan Justice, and members of the McKay lab for
+helpful comments and feedback.

### 2020-09-22 manuscript.Rmd
Draft complete.

@@
 ---
-title: '{memes}: an R interface to the MEME Suite'
+title: 'Memes: an R interface to the MEME Suite'
 author:
-- ORCID: null
+- ORCID: 0000-0003-1000-1579
   name: Spencer L. Nystrom
@@
 output: 
- - word_document:
-   fig.caption: yes
+ - word_document
 ---
 
-```{r setup, echo = FALSE}
+```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
 # Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
@@
 suppressPackageStartupMessages(library(ggplot2))
+suppressPackageStartupMessages(library(ComplexHeatmap))
 drake::r_make()
@@
 
-The MEME Suite family of tools provides a
-comprehensive set of utilities to interrogate motif content across a broad range
-of biological contexts. However, several factors limit the full potential of
-these tools for use in data analysis. 
-  1) must do significant preprocessing to get data in correct input format
-    - written to analyze 1 group of sequences, not amenable to multigroup designs
-    - no common patterns for generating inputs from ChIP-seq whose data are primarily range based
-  2) Easily parseable output data often doesn't contain all information
-    - ex tomtom alternative hits in .tsv output only present in IUPAC format, not matrix
-      - matrix is present in XML file, but no tools exist to parse this
-  3) limited data visualization capabilities
-  4) difficult to integrate data with additional sources
+The MEME Suite family of tools provides a comprehensive set of utilities to
+interrogate motif content across a broad range of biological contexts. However,
+several factors limit the full potential of these tools for use in data
+analysis. MEME Suite tools require carefully formatted inputs to achieve full
+functionality, yet few tools exist to simplify the process of data formatting,
+requiring instead that users write custom code to prepare their data.
+Furthermore, the output data from each MEME Suite tool often have complex
+structures which must be parsed to extract the full suite of information, again
+requiring users to write custom code for this task. Finally, although the
+data-generation capabilities of the MEME Suite are excellent, the tools lack
+powerful ways to visualize the results. These factors act as barriers to
+adoption and preclude deep analysis of the data.
 
@@
 pipelines without additional data processing steps. Additionally, R/Bioconductor
-data structures provide a full-featured representaiton of MEME Suite output
+data structures provide a full-featured representation of MEME Suite output
 data, providing users quick access to all relevant data structures with simple
@@
 accept sequence and motif inputs as needed and any valid commandline flags as
-additional function parameters.
-
-**Describe each tool briefly?**
- - This could also be relegated to a table
-
-**table of comparison to commandline software?**
-  - input type:
-  - output type:
+additional function parameters. A full list of currently supported MEME Suite
+tools and their function is found in \ref{tab:tool_function}.
 
@@
  - **Need to revise above statement, but note that most tools don't have support**
+ - ? To our knowledge, no publicly available tools exist to parse XML formatted data from all MEME Suite tools ...
 
@@
 
-# RETURN
 Motif analysis is important for assays like ChIP-seq which store data as genomic
@@
 
- - `view_motifs()` plots motif as stack (if I decide to add this feature... need to think about this)
- - `ame_plot_heatmap` plot clustered heatmap of AME results 
-   - `ame_compare_heatmap_methods` compare nonlinearity of data visualization in heatmap form
- - `view_tomtom_hits` visualize and compare TomTom matches to assess match quality
-   - `force_best_match` can be used to update "best" match demonstrating how data visualization can be useful for data curation
+The TomTom tool allows users to compare unknown motifs to a set of known motifs
+to identify a best match. Visual inspection of motif comparison data is a key
+step in assessing the quality and correctness of a match. The `view_tomtom_hits`
+function allows users to compare query motifs with the list of potential matches
+as assigned by TomTom. Often, matching tools are useful for reducing the
+search-space for potential candidates, but users are free to leverage their
+understanding of the biology to override these matches. The `force_best_match`
+function allows users to quickly reassign the TomTom best match to a
+lower-ranked, but still statistically significant match to ease downstream data
+processing.
 
@@
 of motif enrichment. However, AME does not provide a mechanism for visualizing
-these results. Additionally, AME is often used 
-## RETURN
+these results. Additionally, AME is often used to examine motif content of
+multiple groups of sequences, therefore a plot which can represent statistical
+significance and group membership of a motif match could be a useful way of
+visualizing the data. `memes` visualized AME results with the `ame_plot_heatmap`
+function, which returns a `ggplot2` formatted heatmap comparing motif
+significance within one or multiple groups. Several options exist to customize
+the heatmap values to help capture different aspects of the output data. The
+`ame_compare_heatmap_methods()` function enables users to compare the
+distribution of values between samples in order to select a threshold which
+accurately captures the dynamic range of their results.
 
@@
 associated with chromatin opening.
-```{r tomtom_example, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
+```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
 plot_top_hit <- function(x){
@@
 ```
-```{r}
+```{r, echo=F}
 library(gt)
@@
   gt(rowname_col = "name",
-     groupname_col = "category",
-     lab = "tab:tomtom_table")
+     groupname_col = "category")
+     # this feature is in development branch??
+     # search SO: gt table reference
+     #lab = "tab:tomtom_table")
 

### 2020-10-19 manuscript.Rmd
add some edits from DJM

@@
 
-Biopolymers, such as DNA and protein, perform varying functions based on their
+Biopolymers, such as DNA and protein molecules, perform varying functions based on their
 primary sequence. Short, repeated sequences, or "motifs" represent functional
-units within biopolymers which can act as interaction surfaces, create
+units within biopolymers that can act as interaction surfaces, create
 structure, or have enzymatic activity. Identification of shared motifs across
-multiple sequences can provide evidence for shared function, thus the ability to
+multiple sequences can provide evidence for shared function. Thus, the ability to
 identify, classify, and compare motifs represents a key step in the analysis of
-biological sequences. Although tools implementing these features exist within
-Bioconductor, they often rely on proprietary data types with poor
-interoperability with other tools. Furthermore, several of these packages will
-be deprecated in the next Bioconductor release, leaving a gap in functionality
-with no clear alternatives.
+biological sequences. **TODO: Transition sentence RE: R/motifs** Although tools
+implementing these features exist within Bioconductor, they often rely on
+proprietary data types with poor interoperability with other tools. **TODO: remove this sentence?** Furthermore,
+several of these packages will be deprecated in the next Bioconductor release,
+leaving a gap in functionality with no clear alternatives.
 
-The MEME Suite is a series of commandline tools which provide a variety of
+The MEME Suite is a series of commandline tools that provide a variety of
 features for motif analysis including *de-novo* motif discovery, enrichment
 testing, comparison, and matching. Here we present `memes`, a motif analysis
-package which provides a seamless interface to MEME Suite tools within R.
+package that provides a seamless interface to MEME Suite tools within R.
 
@@
 to integrate external data to aid analysis, such as filtering a set of
-transcription factor motifs based on gene expression. `memes` also provides
+transcription factor motifs based on gene expression. Lastly, `memes` also provides
 simple utilities for data visualization.
@@
 iteration during analysis. Here we present several examples of how `memes`
-allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
+allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
 interoperability between MEME Suite tools and the broader R landscape.
 
-# Abstract Draft
+# Abstract Draft (OLD: delete??)
 Biopolymers, such as DNA and protein, perform varying functions based on their
@@
 
-Biopolymers, such as DNA and protein, perform varying functions based on their
-primary sequence. Short, repeated sequences, or "motifs" represent functional
-units within biopolymers which can act as interaction surfaces, create
-structure, or have enzymatic activity. Identification of shared motifs across
-multiple sequences can provide evidence for shared function, thus the ability to
-identify, classify, and compare motifs represents a key step in the analysis of
-biological sequences. Although tools that implement these features have
-historically existed in the Bioconductor landscape, they often rely on
-proprietary data types with poor interoperability with other tools. Furthermore,
-several of these packages will be deprecated in the next Bioconductor release,
-leaving a gap in functionality with no clear alternatives.
+Biopolymers, such as DNA and protein molecules, perform varying functions based
+on the composition of their primary sequence. Short, repeated sequences, or
+"motifs" represent functional units within biopolymers which can act as
+interaction surfaces, create structure, or have enzymatic activity.
+Identification of shared motifs across multiple sequences can provide evidence
+for shared function, thus the ability to identify, classify, and compare motifs
+represents a key step in the analysis of biological sequences. Although tools
+that implement these features have historically existed in the Bioconductor
+landscape, they often rely on proprietary data types with poor interoperability
+with other tools. Furthermore, several of these packages will be deprecated in
+the next Bioconductor release, leaving a gap in functionality with no clear
+alternatives.
 
-The MEME Suite is a series of commandline and web-based tools which provide a variety
+The MEME Suite is a series of commandline and web-based tools that provide a variety
 of features for motif analysis such as *de-novo* motif discovery, enrichment
 testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
-accepted set of tools with over 22,000 citations to date, and over 30,000 unique
+utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
 users of the webserver implementation annually \cite{meme-usage-report-page,

### 2020-10-19 manuscript.Rmd
Merge changes w/ dan edits

@@
 
-# Bioconductor Abstract (300 words)
+# Abstract
 
@@
 
-# Abstract Draft (OLD: delete??)
+# Introduction
+
 Biopolymers, such as DNA and protein, perform varying functions based on their
@@
 leaving a gap in functionality with no clear alternatives.
-
-The MEME Suite is a series of commandline tools which provide a variety of
-features for motif analysis such as *de-novo* motif discovery, enrichment
-testing, comparison, and matching. Here we present `memes`, a motif analysis
-package which provides a seamless interface to MEME Suite tools within R.
-
-`memes` uses base R and Bioconductor data types for better integration with
-common analysis tools like the `tidyverse` and other Bioconductor packages.
-Unlike the commandline implementation, `memes` outputs also function as inputs
-to other MEME Suite tools, allowing simple construction of motif analysis
-pipelines.
-
-In addition to wrapping MEME Suite utilities, `memes` extends the functionality
-of `universalmotif` to provide base R and `tidyverse` compatible syntax for
-manipulating and analyzing motif metadata. These features allow motif databases
-to integrate external data to aid analysis, such as filtering a set of
-transcription factor motifs based on gene expression in the tissue
-of interest. `memes` also provides simple utilities for data visualization where
-appropriate.
-
-`memes` is designed for maximum flexibility and ease of use to allow users to
-iterate rapidly during analysis. Here we present several examples of how `memes`
-allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
-interoperability between MEME Suite tools and the broader R package landscape.
-
-# Introduction
-
-The MEME Suite is a series of commandline tools which provide a variety of
-features for motif analysis such as *de-novo* motif discovery, motif enrichment
-testing, motif comparison, and motif matching, among others \cite{meme-suite}.
-MEME Suite tools take as input fasta files of sequences which will be searched
-for motifs, and if necessary, .meme format motif files containing motifs to use
-in the search.
-
-Data from MEME Suite tools can be underutilized because HTML interface?
- - HTML is great human-readable reporting interface
- - `memes` allows customized reporting.
-
-Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memes` can fill the gap and add additional value.
-
-Biopolymers, such as DNA and protein molecules, perform varying functions based
-on the composition of their primary sequence. Short, repeated sequences, or
-"motifs" represent functional units within biopolymers which can act as
-interaction surfaces, create structure, or have enzymatic activity.
-Identification of shared motifs across multiple sequences can provide evidence
-for shared function, thus the ability to identify, classify, and compare motifs
-represents a key step in the analysis of biological sequences. Although tools
-that implement these features have historically existed in the Bioconductor
-landscape, they often rely on proprietary data types with poor interoperability
-with other tools. Furthermore, several of these packages will be deprecated in
-the next Bioconductor release, leaving a gap in functionality with no clear
-alternatives.
 

### 2020-10-22 manuscript.Rmd
More edits.

@@
 structure, or have enzymatic activity. Identification of shared motifs across
-multiple sequences can provide evidence for shared function, thus the ability to
+multiple sequences can provide evidence for shared function, for example,
+identification of kinase substrates based on similarities in phosphorylation
+site sequence, or characterizing DNA elements based on shared transcription
+factor binding sequences (TODO: ref kinase motif:
+https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to
 identify, classify, and compare motifs represents a key step in the analysis of
-biological sequences. Although tools that implement these features have
-historically existed in the Bioconductor landscape, they often rely on
-proprietary data types with poor interoperability with other tools. Furthermore,
-several of these packages will be deprecated in the next Bioconductor release,
+biological sequences. **TODO: R/Bioconductor transition sentence**. Although
+tools that implement these features have historically existed in the
+Bioconductor landscape, they often rely on proprietary data types with poor
+interoperability with other tools, creating barriers to comprehensive motif
+analysis within R. 
+**TODO: fix this section, how to transition?**
+Furthermore, popular motif analysis packages, MotIV, and motifRG 
+will be deprecated in the next Bioconductor release,
 leaving a gap in functionality with no clear alternatives.
@@
 users of the webserver implementation annually \cite{meme-usage-report-page,
-google-scholar-meme}.
+google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
+set of utilities to interrogate motif content across a broad range of biological
+contexts. However, several factors limit the full potential of these tools for
+use in data analysis. MEME Suite tools require carefully formatted inputs to
+achieve full functionality, yet few tools exist to simplify the process of data
+formatting, requiring instead that users write custom code to prepare their
+data, or prepare the inputs by hand, both of which have the potential to be
+error prone without rigorous testing. Furthermore, the output data from each
+MEME Suite tool often have complex structures that must be parsed to extract the
+full suite of information, again requiring users to write custom code for this
+task. Finally, although the data-generation capabilities of the MEME Suite are
+excellent, the tools lack powerful ways to visualize the results. Collectively,
+these factors act as barriers to adoption and preclude deep analysis of the
+data, as well as communication of results to the scientific community.
 
-The MEME Suite family of tools provides a comprehensive set of utilities to
-interrogate motif content across a broad range of biological contexts. However,
-several factors limit the full potential of these tools for use in data
-analysis. MEME Suite tools require carefully formatted inputs to achieve full
-functionality, yet few tools exist to simplify the process of data formatting,
-requiring instead that users write custom code to prepare their data.
-Furthermore, the output data from each MEME Suite tool often have complex
-structures which must be parsed to extract the full suite of information, again
-requiring users to write custom code for this task. Finally, although the
-data-generation capabilities of the MEME Suite are excellent, the tools lack
-powerful ways to visualize the results. These factors act as barriers to
-adoption and preclude deep analysis of the data.
-
-Here we present `memes`, a motif analysis package which provides a seamless
+Here we present `memes`, a motif analysis package that provides a seamless
 interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
@@
 data, providing users quick access to all relevant data structures with simple
-syntax.
+syntax. **TODO: Depending on length restrictions, end here**
 
@@
 of `universalmotif` to provide base R and `tidyverse` compatible syntax for
-manipulating and analyzing motif metadata. These features allow motif databases
-to integrate external data to aid analysis, such as filtering a set of
-transcription factor motifs based on gene expression in the tissue of interest.
+manipulating and analyzing motif metadata. These features allow users to
+integrate external data with motif databases to aid analysis, such as filtering
+a set of transcription factor motifs based on gene expression in the tissue of
+interest, then using that filtered set to search for motifs associated with
+expressed factors.
 
@@
 
-# Results 
 ## Design & Implementation
@@
 R/Bioconductor package for representing motif matrices and their associated
-metadata \cite{universalmotif}. `memes` will also always accept paths to files
+metadata \cite{universalmotif}. `memes` will also accept paths to files
 on disk, such as fasta files for sequence inputs, and meme format files for
-motif inputs, reducing the need to read large files into memory to run MEME
-Suite tools.
+motif inputs, reducing the need to read large files into memory.
+
+**Restructure below:**
+ - MEME tools return HTML reports for each tool which display data in user-friendly way, but poor for data processing.
+However, some of the simple raw data (like tsv data) don't contain the full set of statistics reported to the user.
+These data are contained in XML files, which are difficult to parse.
 
 Although MEME Suite tools accept consistent data type inputs, they do not return
-consistent data structures across tools. For example, the tool TomTom, which
+consistent data structures across tools. For example, the TomTom tool, which
 compares query motifs to a database of known motifs, returns tab-separated
-results which do not contain the frequency matrix representations of the matched
+results that do not contain the frequency matrix representations of the matched
 motifs. Instead, users must write custom code to parse these matrices back out
@@
 
-Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
-describe properties of motifs. For example, MEME and DREME return de-novo
+Many of the MEME Suite tools return files which
+contianing descriptive properties of motifs. For example, MEME and DREME return de-novo
 discovered motifs from query sequences along with statistical information about
@@
 
-Motif analysis is important for assays like ChIP-seq which store data as genomic
-coordinates, not sequence. This is in conflict with MEME Suite which is designed
-to study sequences. Therefore, users must write custom code to extract DNA
-sequence of their ranges of interest. Although tools like `bedtools getfasta`
-can simplify the process of extracting DNA sequence from genomic coordinates,
-some MEME tools require fasta headers to be specially formatted to extract full
-functionality, requiring users to write additional custom code to properly
-format input data.
+Motif analysis is often employed in ChIP-seq analysis, in which data is stored as genomic
+coordinates rather than sequence. However, MEME Suite tools are designed to work
+with sequences. While existing tools such as bedtools can extract DNA sequence
+from gneomic coordinates, some MEME tools require fasta headers to be
+specifically formatted. As a result, users must write custom code to extracd DNA
+sequence for their genomic ranges of interest.
 
-`memes` provides the `get_sequence` function which automates the extraction of
+The `memes` function `get_sequence` automates extraction of
 DNA sequence from genomic coordinates while simultaneously producing MEME Suite
@@
 representations, such as bed format, are easily imported as `GenomicRanges`
-objects into R using preexisting import functions, meaning users must not write
-any custom import functions to work with range-based data using `memes`. 
-Sequences are returned in `Biostrings` format and can therefore be used as
-inputs to all `memes` commands or as inputs to other R/Bioconductor functions
-for sequence analysis.
+objects into R using preexisting import functions, meaning `memes` users do not
+have to write any custom import functions to work with range-based data using
+`memes`.  Sequences are returned in `Biostrings` format and can therefore be
+used as inputs to all `memes` commands or as inputs to other R/Bioconductor
+functions for sequence analysis.
 
-During analysis of multiomic data, users often uncover different functional
-categories of genomic regions through integration of multiple datasets. In these
-instances, motif analysis acts as a powerful tool to examine whether differences
-in motif content is associated with membership of distinct categories. In some
-cases, performing differential motif analysis (using one category as the "test"
-set and another as the "control") can uncover biologically relevant motifs which
-are enriched in one category relative to another. Although the MEME Suite allows
-differential enrichment testing, it does not inherently provide a mechanism for
-analyzing groups of sequences in parallel, or performing motif analysis with any
-understanding of data categories. The `memes` framework enables "data-aware"
-motif analysis workflows by allowing named lists of `Biostrings` objects as
-input to each function. If using `GenomicRanges`, users can split peak data on a
-metadata column using the base R `split` function, and use this result as input
-to `get_sequence`, which will produce a list of `BioStrings` objects where each
-entry is named after each data category from the split column. When a list is
-used as input to a `memes` function, it runs the corresponding MEME Suite tool
-for each object in the list. Users can also pass the name(s) of a category to
-the `control` argument to automatically enable differential analysis of all
-other list members against the control category sequences.  Therefore, `memes`
-enables complex data-aware differential motif analysis workflows using simple
-syntax to extend the capabilities of the MEME Suite.
+During analysis of multiomic data, users can identify different functional
+categories of genomic regions through integration with orthogonal datasets. **TODO
+For example, .**
 
-### Table of supported MEME Suite Tools
- - comparison to other tools?
- - Maybe only do this if asked
+In these instances, motif analysis acts as a powerful tool to
+examine whether differences in motif content is associated with membership of
+distinct categories. In some cases, performing differential motif analysis
+(using one category as the "test" set and another as the "control") can uncover
+biologically relevant motifs which are enriched in one category relative to
+another. Although the MEME Suite allows differential enrichment testing, it does
+not inherently provide a mechanism for analyzing groups of sequences in
+parallel, or performing motif analysis with any understanding of data
+categories. The `memes` framework enables "data-aware" motif analysis workflows
+by allowing named lists of `Biostrings` objects as input to each function. If
+using `GenomicRanges`, users can split peak data on a metadata column using the
+base R `split` function, and then use this result as input to `get_sequence`, which
+will produce a list of `BioStrings` objects where each entry is named after the
+data categories from the split column. When a list is used as input to a `memes`
+function, it runs the corresponding MEME Suite tool for each object in the list.
+Users can also pass the name(s) of a category to the `control` argument to
+enable differential analysis of the remaining list members against the
+control category sequences. Therefore, `memes` enables complex data-aware
+differential motif analysis workflows using simple syntax to extend the
+capabilities of the MEME Suite.
 
 ## Data visualization
-The MEME Suite provides a small set of data visualizations which have limited
+The MEME Suite provides a small set of data visualizations that have limited
 customizability. `memes` leverages the advantages of the R graphics environment
-to provide novel data visualizations which are infinitely customizable by the user.
+to provide a wide range of data visualization options that are highly
+customizable. We describe two scenarios below.
 
 The TomTom tool allows users to compare unknown motifs to a set of known motifs
-to identify a best match. Visual inspection of motif comparison data is a key
-step in assessing the quality and correctness of a match. The `view_tomtom_hits`
+to identify the best match. Visual inspection of motif comparison data is a key
+step in assessing the quality and accuracy of a match. The `view_tomtom_hits`
 function allows users to compare query motifs with the list of potential matches
-as assigned by TomTom. Often, matching tools are useful for reducing the
-search-space for potential candidates, but users are free to leverage their
-understanding of the biology to override these matches. The `force_best_match`
-function allows users to quickly reassign the TomTom best match to a
-lower-ranked, but still statistically significant match to ease downstream data
-processing.
+as assigned by TomTom. The `force_best_match` function allows users to
+reassign the TomTom best match to a lower-ranked (but still statistically
+significant) match in order to highlight motifs with greater biological
+relevance (e.g. to skip over a transcription factor that is not expressed in the
+experimental sample, or when two motifs are matched equally well).
 
-The AME tool searches for enrichment of motifs within groups of sequences
-relative to a control set. The meaningful result from this tool is the
+The AME tool searches for enrichment of motifs within a set of experimental
+sequences relative to a control set. The meaningful result from this tool is the
 statistical parameter (for example, a p-value) associated with the significance
 of motif enrichment. However, AME does not provide a mechanism for visualizing
-these results. Additionally, AME is often used to examine motif content of
-multiple groups of sequences, therefore a plot which can represent statistical
-significance and group membership of a motif match could be a useful way of
-visualizing the data. `memes` visualized AME results with the `ame_plot_heatmap`
-function, which returns a `ggplot2` formatted heatmap comparing motif
-significance within one or multiple groups. Several options exist to customize
-the heatmap values to help capture different aspects of the output data. The
-`ame_compare_heatmap_methods()` function enables users to compare the
-distribution of values between samples in order to select a threshold which
-accurately captures the dynamic range of their results.
+these results.
+
+The `ame_plot_heatmap` function in `memes` returns a `ggplot2` formatted heatmap of
+statistical significance of motif enrichment. If AME is used to examine motif
+content of multiple groups of sequences, the `ame_plot_heatmap` function can also
+return a plot comparing motif significance within multiple groups. Several
+options exist to customize the heatmap values in order to capture different aspects
+of the output data. The `ame_compare_heatmap_methods()` function enables users
+to compare the distribution of values between samples in order to select a
+threshold that accurately captures the dynamic range of their results.
 
@@
 `memes` relies on a locally installed version of the MEME Suite which is
-accessed by the users R process. Although R is available on Windows, Mac, and
+accessed by the user's R process. Although R is available on Windows, Mac, and
 Linux operating systems, the MEME Suite is incompatible with Windows, limiting
 its adoption by Windows users. Additional barriers also exist to installing a
-local copy of the MEME Suite on compatible systems. The MEME Suite relies on
-several system-level dependencies whose installation can be difficult for novice
-users. Finally, some tools in the MEME Suite use python to generate shuffled
-control sequences for analysis, which presents a reproducibility issue as the
-random number generation algorithm changed from python2.7 to python3. The MEME
-Suite will build and install on both python2.7 and python3 systems, therefore
-without careful consideration of this behavior, the same code run on two systems
-may not produce identical results, even if using the same major version of the
-MEME Suite. In order to democratize access to the MEME Suite on unsupported
-operating systems, and to facilitate reproducible motif analysis using `memes`, 
-we have also developed docker and singularity (**NOTE:** singularity is broken
-for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
-preinstalled version of the MEME Suite along with an R/Bioconductor analysis
-environment including the most recent version of `memes` and its dependencies.
-As new container versions are released, they are version tagged and stored to
-ensure reproducibility while allowing updates to the container.
+local copy of the MEME Suite on compatible systems, for example, the MEME Suite
+relies on several system-level dependencies whose installation can be difficult
+for novice users. Finally, some tools in the MEME Suite use python to generate
+shuffled control sequences for analysis, which presents a reproducibility issue
+as the random number generation algorithm changed between python2.7 and python3. The
+MEME Suite will build and install on both python2.7 and python3 systems,
+therefore without careful consideration of this behavior, the same code run on
+two systems may not produce identical results, even if using the same major
+version of the MEME Suite. In order to increase access to the MEME Suite on
+unsupported operating systems, and to facilitate reproducible motif analysis
+using `memes`, we have also developed docker and singularity (**TODO: NOTE:**
+singularity is broken for `rocker/rstudio`, so maybe omit this until fixed)
+containers containing a preinstalled version of the MEME Suite along with an
+R/Bioconductor analysis environment including the most recent version of `memes`
+and its dependencies.  As new container versions are released, they are
+version-tagged and stored to ensure reproducibility while allowing updates to
+the container.
 
@@
 tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
-in the `memes` vignette and the package website
+in the `memes` vignettes and the package website
 ([snystrom.github.io/memes](snystrom.github.io/memes)). In the following

### 2020-10-27 manuscript.Rmd
reorganizing the results.

@@
 
-The core functionality of `memes` revolves around the interface to MEME Suite
-tools. Each supported MEME Suite tool is assigned a `run` function
-(`runDreme()`, `runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
-accept sequence and motif inputs as needed and any valid commandline flags as
-additional function parameters. A full list of currently supported MEME Suite
-tools and their function is found in \ref{tab:tool_function}.
-
-### Data Structures
-
 MEME Suite tools are run on the commandline and use files stored on-disk as
 input while returning a series of output files containing varying data types. 
-`memes` reduces each MEME Suite tool to an R function which internally handles
-file I/O such that all MEME Suite calling functions accept R objects as input,
-and return R objects as output. Sequence inputs are stored in `BioStrings`
-format, an R/Bioconductor package for storing biopolymer sequence data
-\cite{BioStrings}. Motif inputs are passed as `universalmotif` objects, another
-R/Bioconductor package for representing motif matrices and their associated
-metadata \cite{universalmotif}. `memes` will also accept paths to files
+As a wrapper of MEME tools, `memes` functions similarly by assigning each
+supported MEME Suite tool to a `run` function (`runDreme()`,
+`runMeme()`, `runAme()`, `runFimo()`, `runTomTom()`), which
+internally writes input objects to files on disk, runs the tool, then imports
+the data as R objects. These functions accept sequence and motif inputs as
+required by the tool. Sequence inputs are accepted in `BioStrings` format, an
+R/Bioconductor package for storing biopolymer sequence data \cite{BioStrings}.
+Motif inputs are passed as `universalmotif` objects, another R/Bioconductor
+package for representing motif matrices and their associated metadata
+\cite{universalmotif}. `memes` `run` functions will also accept paths to files
 on disk, such as fasta files for sequence inputs, and meme format files for
-motif inputs, reducing the need to read large files into memory.
+motif inputs, reducing the need to read large files into memory. Finally, each
+`run` function contains optional function parameters mirroring the commandline
+arguments provided by each MEME Suite tool. In this way, `memes` provides a
+feature-complete interface to the supported MEME Suite tools.
 
-**Restructure below:**
- - MEME tools return HTML reports for each tool which display data in user-friendly way, but poor for data processing.
-However, some of the simple raw data (like tsv data) don't contain the full set of statistics reported to the user.
-These data are contained in XML files, which are difficult to parse.
+### Output Data Structures
 
-Although MEME Suite tools accept consistent data type inputs, they do not return
-consistent data structures across tools. For example, the TomTom tool, which
-compares query motifs to a database of known motifs, returns tab-separated
-results that do not contain the frequency matrix representations of the matched
-motifs. Instead, users must write custom code to parse these matrices back out
-from the input databases, creating additional friction for creating novel data
-visualizations and exploratory analyses. Often, output data are also returned in
-XML format, which although they hold all relevant information, are difficult to
-parse. To our knowledge, no publicly available tools exist to parse MEME Suite
-XML formatted data, therefore users must write custom code to extract these
-data. `memes` exports data import functions for each supported MEME Suite tool,
-which import these data as R data.frames. These functions also work internally to
-create the seamless file I/O when calling MEME Suite tools, but the import
-functions can be called directly by users to import previously generated data
-for use in R.
+MEME tools return HTML reports for each tool that display data in a
+user-friendly way, however, these files are not ideal for downstream processing.
+Depending on the tool, data are also returned in tab-separated, or XML format,
+which are more amenable to computational processing. However, in the case that
+tab-separated data are returned, results are often incomplete. For example, the
+TomTom tool, which compares query motifs to a database of known motifs, returns
+tab-separated results that do not contain the frequency matrix representations
+of the matched motifs. Instead, users must write custom code to parse these
+matrices back out from the input databases, creating additional barriers to
+analysis. In the case that data are returned in XML format, these files contain
+all relevant data, however are difficult to parse, again requiring users to
+write custom code to extract the necessary data. Finally, the data types
+contained in MEME Suite outputs are complex and multidimensional, and thus
+require special data structures to properly organize the data in memory.
 
- - **NOTE: Biopython has a MEME/MAST parser but it works on tsv https://biopython.org/docs/1.74/api/Bio.motifs.meme.html**
- - **Actually, they merged XML import in March: https://github.com/biopython/biopython/pull/2354**
- - **Need to revise above statement, but note that most tools don't have support**
- - ? To our knowledge, no publicly available tools exist to parse XML formatted data from all MEME Suite tools ...
+`memes` provides data import functions for each supported MEME Suite tool, which
+import these data as modified R data.frames (described in detail below). These
+functions can be called directly by users to import data previously generated by
+the commandline or webserver versions of the MEME Suite for use in R. These
+import functions also underlie the import step internal to each of the `run`
+functions.
 
-Many of the MEME Suite tools return files which
-contianing descriptive properties of motifs. For example, MEME and DREME return de-novo
-discovered motifs from query sequences along with statistical information about
-their enrichment. The `universalmotif` data structure is a powerful
-R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
-However, `universalmotif` objects are not ideal for performing bulk manipulation
-or queries of metadata, nor do they support complex data structures such as
-nested elements. Many MEME Suite analysis pipelines involve identifying motifs
-which are further analyzed by multiple tools, creating additional metadata which
-cannot be easily tracked in bulk using `universalmotif` format. Thus
-an ideal representation for these data is one which can hold an unlimited amount
-of metadata, contain complex hierarchies, and be easily manipulated using
-standard analysis tools. `memes` therefore extends the `universalmotif` format
-by providing data.frames holding motif metadata which retain links to their
-`universalmotif` representation. These universalmotif
-data.frames combine the advantages of the `universalmotif` framework for motif
-manipulation with the extensibility and flexibility of R data.frames. These
-`universalmotif` data.frames form the basis for a majority of `memes` data
-outputs. These structures are also valid input types to `memes` functions, and
-when used as such, output data are appended as new columns to the input data,
-ensuring data provenance.
+### structured data.frames hold multidimensional output data
+
+ - as mentioned previously, the multidimensional naature of meme results is difficult to represent
+ - Output data usually contain motifs & descriptive statistics about those motifs
+ - motif info can be stored in universalmotif format, however, not enough space for data
+ - data.frames do good job of storing data (incl nested data), but cannot store motifs
+ - we solve this problem by storing the unviversalmotif inside the data.frame (which is valid R)
+ - we add nested list columns as needed to link multidimensional data
+ - advantage of this approach over an S4 is all normal tools work on it (tidyverse), joins, etc.
+
+Many of the MEME Suite tools return files contianing descriptive
+properties of motifs. For example, MEME and DREME return de-novo discovered
+motifs from query sequences along with statistical information about their
+enrichment. The `universalmotif` data structure is a powerful R/Bioconductor
+representation for motif matrices and their associated metadata
+\cite{universalmotif}.  However, `universalmotif` objects are not ideal for
+performing bulk manipulation or queries of metadata, nor do they support complex
+data structures such as nested elements. Many MEME Suite analysis pipelines
+involve identifying motifs which are further analyzed by multiple tools,
+creating additional metadata which cannot be easily tracked in bulk using
+`universalmotif` format. Thus an ideal representation for these data is one
+which can hold an unlimited amount of metadata, contain complex hierarchies, and
+be easily manipulated using standard analysis tools. `memes` therefore extends
+the `universalmotif` format by providing data.frames holding motif metadata
+which retain links to their `universalmotif` representation. These
+universalmotif data.frames combine the advantages of the `universalmotif`
+framework for motif manipulation with the extensibility and flexibility of R
+data.frames. These `universalmotif` data.frames form the basis for a majority of
+`memes` data outputs. These structures are also valid input types to `memes`
+functions, and when used as such, output data are appended as new columns to the
+input data, ensuring data provenance.
 
@@
 
-Motif analysis is often employed in ChIP-seq analysis, in which data is stored as genomic
-coordinates rather than sequence. However, MEME Suite tools are designed to work
-with sequences. While existing tools such as bedtools can extract DNA sequence
-from gneomic coordinates, some MEME tools require fasta headers to be
-specifically formatted. As a result, users must write custom code to extracd DNA
-sequence for their genomic ranges of interest.
+Motif analysis is often employed in ChIP-seq analysis, in which data is stored
+as genomic coordinates rather than sequence. However, MEME Suite tools are
+designed to work with sequences. While existing tools such as bedtools can
+extract DNA sequence from gneomic coordinates, some MEME tools require fasta
+headers to be specifically formatted. As a result, users must write custom code
+to extracd DNA sequence for their genomic ranges of interest.
 

### 2020-10-27 manuscript.Rmd
typo fix

@@
 
+# CONTINUE
  - as mentioned previously, the multidimensional naature of meme results is difficult to represent
@@
 
-Many of the MEME Suite tools return files contianing descriptive
+Many of the MEME Suite tools return files containing descriptive
 properties of motifs. For example, MEME and DREME return de-novo discovered

### 2020-10-30 manuscript.Rmd
Finish second draft edits.

--- 2020-10-27 manuscript.Rmd
@@
 factor binding sequences (TODO: ref kinase motif:
-https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to
-identify, classify, and compare motifs represents a key step in the analysis of
-biological sequences. **TODO: R/Bioconductor transition sentence**. Although
-tools that implement these features have historically existed in the
-Bioconductor landscape, they often rely on proprietary data types with poor
-interoperability with other tools, creating barriers to comprehensive motif
-analysis within R. 
-**TODO: fix this section, how to transition?**
-Furthermore, popular motif analysis packages, MotIV, and motifRG 
-will be deprecated in the next Bioconductor release,
-leaving a gap in functionality with no clear alternatives.
+https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
+classify, and compare motifs represents a key step in the analysis of biological
+sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
+implement these features exist in Bioconductor, they often rely on
+proprietary data types or have poor interoperability with other tools, creating
+barriers to comprehensive motif analysis within R. The recent addition of
+`universalmotif` to the Bioconductor landscape partially addresss these issues
+by standardizing motif formats and providing a powerful set of tools for
+examining the properties of motifs, however the Bioconductor motif analysis
+landscape remains incomplete \cite{universalmotif}.
 
@@
 interest, then using that filtered set to search for motifs associated with
-expressed factors.
+expressed factors. By maintaining compatability with `universalmotif`, `memes`
+outputs can be readily used in a diverse set of tools beyond the `memes`
+framework.
 
@@
 
- - as mentioned previously, the multidimensional naature of meme results is difficult to represent
- - Output data usually contain motifs & descriptive statistics about those motifs
- - motif info can be stored in universalmotif format, however, not enough space for data
- - data.frames do good job of storing data (incl nested data), but cannot store motifs
- - we solve this problem by storing the unviversalmotif inside the data.frame (which is valid R)
- - we add nested list columns as needed to link multidimensional data
- - advantage of this approach over an S4 is all normal tools work on it (tidyverse), joins, etc.
+Data returned from MEME Suite tools are complex and multidimensional, making
+them difficult to represent in a simple data structure. For example, MEME and
+DREME return de-novo discovered motifs from query sequences along with
+statistical information about their enrichment. In this instance, a position
+frequency matrix (PFM) of the discovered motif can be represented as a matrix,
+while the properties of that matrix (e.g. the name of the motif, the E-value
+from the enrichment test, background nucleotide frequency, etc.) must be encoded
+outside of the matrix, while maintaining their relationship to the
+corresponding PFM. However, other MEME Suite tools, such as TOMTOM, which
+compares one motif to a series of several motifs to identify possible matches,
+produce an additional layer of complexity such that input matrices (which
+contain metadata as previously described) can have a one-to-many relationship
+with other motif matrices (again with their own metadata). Thus an ideal
+representation for these data is one which can hold an unlimited number of motif
+matrices and their metadata, contain complex hierarchies, and be easily
+manipulated using standard analysis tools.
 
-Many of the MEME Suite tools return files contianing descriptive
-properties of motifs. For example, MEME and DREME return de-novo discovered
-motifs from query sequences along with statistical information about their
-enrichment. The `universalmotif` data structure is a powerful R/Bioconductor
-representation for motif matrices and their associated metadata
-\cite{universalmotif}.  However, `universalmotif` objects are not ideal for
-performing bulk manipulation or queries of metadata, nor do they support complex
-data structures such as nested elements. Many MEME Suite analysis pipelines
-involve identifying motifs which are further analyzed by multiple tools,
-creating additional metadata which cannot be easily tracked in bulk using
-`universalmotif` format. Thus an ideal representation for these data is one
-which can hold an unlimited amount of metadata, contain complex hierarchies, and
-be easily manipulated using standard analysis tools. `memes` therefore extends
-the `universalmotif` format by providing data.frames holding motif metadata
-which retain links to their `universalmotif` representation. These
-universalmotif data.frames combine the advantages of the `universalmotif`
-framework for motif manipulation with the extensibility and flexibility of R
-data.frames. These `universalmotif` data.frames form the basis for a majority of
-`memes` data outputs. These structures are also valid input types to `memes`
+The `universalmotif` data structure is a powerful R/Bioconductor representation
+for motif matrices and their associated metadata \cite{universalmotif}. However,
+`universalmotif` objects are not ideal for performing bulk manipulation or
+queries of metadata, nor do they support complex data structures such as nested
+elements. R data.frames support nesting arbitrary data types within columns, but
+are not sufficient to properly store motif matrices. In order to create an
+extensible, nested data structure for tracking MEME Suite data, `memes`
+leverages the power of R data.frames and the `universalmotif` format by nesting
+`universalmotif` objects within data.frames. These universalmotif data.frames
+store information about motifs along rows, where each column stores metadata
+associated with a motif, including a `motif` column holding a `universalmotif`
+representation of that motif, which can be manipulated by altering columns of
+the data.frame (described in detail below). Adding one-to-many relationships within this
+structure is trivial, as additional unviersalmotif data.frames can be nested
+within eachother. These universalmotif data.frames form the basis for a majority
+of `memes` data outputs. These structures are also valid input types to `memes`
 functions, and when used as such, output data are appended as new columns to the
-input data, ensuring data provenance.
+input data, ensuring data provenance. Finally, because the universalmotif
+data.frame is just a data.frame, it can be manipulated using base R or
+`tidyverse` functions, just like regular R data.frames, lowering the cognitive
+overhead to interfacing with these structures.
 
@@
 
-`universalmotif` data.frames are not limited to storing motifs from MEME Suite
+universalmotif data.frames are not limited to storing motifs from MEME Suite
 data, and can additionally be used to examine or manipulate the properties of
-any motif list or database. Using `universalmotif` functions to import motif data,
+any motif list. Using `universalmotif` functions to import motif data,
 the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
-into a `universalmotif` data.frame which can be manipulated using base R or
-`tidyverse` functions, just like regular R data.frames. Manipulation of linked
-metadata columns will result in updating the linked `universalmotif` object when
-calling `update_motifs()` to retain the data.frame structure, or when calling
-`as_universalmotif()` to convert the data.frame back into `universalmotif` format. In
-this way, `memes` enables large-scale motif database manipulation & analysis
-within the preexisting R framework.
+into a `universalmotif` data.frame.  Manipulation of linked metadata columns
+will result in updating the linked `universalmotif` object when calling
+`update_motifs()` to retain the data.frame structure, or when calling
+`as_universalmotif()` to convert the data.frame back into `universalmotif`
+format. In this way, `memes` enables large-scale motif database manipulation &
+analysis within the preexisting R framework.
+
+The universalmotif data.frames generated by memes are not limited to storing
+motifs from MEME Suite data. They can also be used to manipulate or store other
+descriptive properties of motifs from multiple sources. `universalmotif`
+provides several import functions which can convert motifs from external
+databases queried from the `MotifDb` package, or from files in a variety of
+non-meme file formats \cite{motifdb}. Once imported in `universalmotif` format
+the `memes` function `as_universalmotif_dataframe()` converts any
+`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
+data stored in this format, if changes are made to the linked metadata columns,
+the linked universalmotif object can be updated by calling `update_motifs()` to
+retain the data.frame structure for further analysis, or by calling
+`as_universalmotif()` to convert the data.frame back into universalmotif format,
+discarding columns which cannot be stored within `universalmotif` format.
+Therefore, this data structure empowers users to import and analyze motif lists
+from a diverse set of sources for use with `memes` or simply for manipulating
+motif metadata using common R tools. In this way, memes enables large-scale
+motif database manipulation & analysis within the preexisting R framework.
 
@@
 headers to be specifically formatted. As a result, users must write custom code
-to extracd DNA sequence for their genomic ranges of interest.
+to extract the DNA sequence for their genomic ranges of interest.
 
@@
 
-During analysis of multiomic data, users can identify different functional
-categories of genomic regions through integration with orthogonal datasets. **TODO
-For example, .**
+### Data-aware motif analysis
 
-In these instances, motif analysis acts as a powerful tool to
-examine whether differences in motif content is associated with membership of
-distinct categories. In some cases, performing differential motif analysis
-(using one category as the "test" set and another as the "control") can uncover
-biologically relevant motifs which are enriched in one category relative to
-another. Although the MEME Suite allows differential enrichment testing, it does
+Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
+sequences relative to another set, can uncover biologically relevant motifs
+associated with membership to distinct categories. For example, during analysis
+of multiomic data, users can identify different functional categories of genomic
+regions through integration with orthogonal datasets (e.g.  categorizing
+transcription factor binding sites by the presence or absence of another
+factor). In these instances, motif analysis acts as a powerful tool to examine whether
+differences in motif content is associated with membership of distinct
+categories (e.g. motifs enriched at sites where both factors are bound vs singly
+bound). Although the MEME Suite allows differential enrichment testing, it does
 not inherently provide a mechanism for analyzing groups of sequences in
@@
 function allows users to compare query motifs with the list of potential matches
-as assigned by TomTom. The `force_best_match` function allows users to
-reassign the TomTom best match to a lower-ranked (but still statistically
-significant) match in order to highlight motifs with greater biological
-relevance (e.g. to skip over a transcription factor that is not expressed in the
-experimental sample, or when two motifs are matched equally well).
+as assigned by TomTom, similar to the commandline behavior. The
+`force_best_match` function allows users to reassign the TomTom best match
+to a lower-ranked (but still statistically significant) match in order to
+highlight motifs with greater biological relevance (e.g. to skip over a
+transcription factor that is not expressed in the experimental sample, or when
+two motifs are matched equally well).
 

### 2020-11-05 manuscript.Rmd
more edits

@@
 primary sequence. Short, repeated sequences, or "motifs" represent functional
-units within biopolymers which can act as interaction surfaces, create
+units within biopolymers that can act as interaction surfaces, create
 structure, or have enzymatic activity. Identification of shared motifs across
@@
 by standardizing motif formats and providing a powerful set of tools for
-examining the properties of motifs, however the Bioconductor motif analysis
+examining the properties of motifs. However the Bioconductor motif analysis
 landscape remains incomplete \cite{universalmotif}.
@@
 excellent, the tools lack powerful ways to visualize the results. Collectively,
-these factors act as barriers to adoption and preclude deep analysis of the
-data, as well as communication of results to the scientific community.
+these factors act as barriers to adoption, preclude deeper analysis of the
+data, and limit communication of results to the scientific community.
 
@@
 Providing R representations of MEME Suite data objects enables generation of
-novel data visualizations using the well established tools available in the R
+novel data visualizations using the well-established tools available in the R
 landscape. Therefore, `memes` empowers users to deeply analyze and visualize
@@
 MEME tools return HTML reports for each tool that display data in a
-user-friendly way, however, these files are not ideal for downstream processing.
+user-friendly way; however, these files are not ideal for downstream processing.
 Depending on the tool, data are also returned in tab-separated, or XML format,
@@
 analysis. In the case that data are returned in XML format, these files contain
-all relevant data, however are difficult to parse, again requiring users to
+all relevant data; however, XML files are difficult to parse, again requiring users to
 write custom code to extract the necessary data. Finally, the data types
@@
 with other motif matrices (again with their own metadata). Thus an ideal
-representation for these data is one which can hold an unlimited number of motif
+representation for these data is one that can hold an unlimited number of motif
 matrices and their metadata, contain complex hierarchies, and be easily
@@
 input data, ensuring data provenance. Finally, because the universalmotif
-data.frame is just a data.frame, it can be manipulated using base R or
-`tidyverse` functions, just like regular R data.frames, lowering the cognitive
-overhead to interfacing with these structures.
+data.frame is a standard R data.frame, it can be manipulated using base R or
+`tidyverse` functions, lowering the cognitive overhead to interfacing with these
+structures.
 
@@
 data, and can additionally be used to examine or manipulate the properties of
-any motif list. Using `universalmotif` functions to import motif data,
-the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
-into a `universalmotif` data.frame.  Manipulation of linked metadata columns
-will result in updating the linked `universalmotif` object when calling
+any motif list. Using `universalmotif` functions to import motif data, the
+`memes` function `as_universalmotif_dataframe()` converts any `universalmotif`
+object into a `universalmotif` data.frame.  Manipulation of linked metadata
+columns will result in updating the linked `universalmotif` object when calling
 `update_motifs()` to retain the data.frame structure, or when calling
@@
 enable differential analysis of the remaining list members against the
-control category sequences. Therefore, `memes` enables complex data-aware
+control category sequences. In this manner, `memes` enables complex data-aware
 differential motif analysis workflows using simple syntax to extend the

### 2020-11-09 manuscript.Rmd
More edits.

@@
   name: Spencer L. Nystrom
-- name: Daniel J. McKay
+- ORCID: 0000-0001-8226-0604
+  name: Daniel J. McKay
 output: 
@@
 
-Biopolymers, such as DNA and protein molecules, perform varying functions based on their
-primary sequence. Short, repeated sequences, or "motifs" represent functional
-units within biopolymers that can act as interaction surfaces, create
-structure, or have enzymatic activity. Identification of shared motifs across
-multiple sequences can provide evidence for shared function. Thus, the ability to
-identify, classify, and compare motifs represents a key step in the analysis of
-biological sequences. **TODO: Transition sentence RE: R/motifs** Although tools
-implementing these features exist within Bioconductor, they often rely on
-proprietary data types with poor interoperability with other tools. **TODO: remove this sentence?** Furthermore,
-several of these packages will be deprecated in the next Bioconductor release,
-leaving a gap in functionality with no clear alternatives.
-
-The MEME Suite is a series of commandline tools that provide a variety of
-features for motif analysis including *de-novo* motif discovery, enrichment
-testing, comparison, and matching. Here we present `memes`, a motif analysis
-package that provides a seamless interface to MEME Suite tools within R.
-
-`memes` uses standard R and Bioconductor data types for better integration with
-common analysis tools like the `tidyverse` and `GRanges`. `memes` outputs also
-function as inputs to other `memes` commands, allowing simple construction of motif
-analysis pipelines.
-
-In addition to wrapping MEME Suite utilities, `memes` extends the functionality
-of `universalmotif` to provide base R and `tidyverse` compatible syntax for
-manipulating and analyzing motif metadata. These features allow motif databases
-to integrate external data to aid analysis, such as filtering a set of
-transcription factor motifs based on gene expression. Lastly, `memes` also provides
-simple utilities for data visualization.
-
-`memes` is designed for maximum flexibility and ease of use to allow rapid
-iteration during analysis. Here we present several examples of how `memes`
-allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
-interoperability between MEME Suite tools and the broader R landscape.
+Identification of biopolymer motifs represents a key step in the analysis of
+biological sequences. The MEME Suite is a widely used toolkit for comprehensive
+analysis  of biopolymer motifs; however, these tools are poorly integrated
+within popular analysis frameworks like the R/Bioconductor project, creating
+barriers to their use. Here we present `memes`, an R package which provides a
+seamless R interface to the MEME Suite. `memes` provides a novel "data aware"
+interface to these tools, enabling rapid and complex discriminative motif analysis workflows.
+In addition to interfacing with popular MEME Suite tools, `memes` leverages
+existing R/Bioconductor data structures to store the complex, multidimensional
+data returned by MEME Suite tools for rapid data access and manipulation.
+Finally, `memes` provides data visualization capabilities to facilitate
+communication of results. `memes` is available at
+[github.com/snystrom/memes](github.com/snystrom/memes).
 
@@
 primary sequence. Short, repeated sequences, or "motifs" represent functional
-units within biopolymers that can act as interaction surfaces, create
-structure, or have enzymatic activity. Identification of shared motifs across
-multiple sequences can provide evidence for shared function, for example,
-identification of kinase substrates based on similarities in phosphorylation
-site sequence, or characterizing DNA elements based on shared transcription
-factor binding sequences (TODO: ref kinase motif:
+units within biopolymers that can act as interaction surfaces, create structure,
+or have enzymatic activity. Identification of shared motifs across multiple
+sequences can provide evidence for shared function, for example, identification
+of kinase substrates based on similarities in phosphorylation site sequence, or
+characterizing DNA elements based on shared transcription factor binding
+sequences (TODO: ref kinase motif:
 https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
 classify, and compare motifs represents a key step in the analysis of biological
-sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
-implement these features exist in Bioconductor, they often rely on
-proprietary data types or have poor interoperability with other tools, creating
-barriers to comprehensive motif analysis within R. The recent addition of
-`universalmotif` to the Bioconductor landscape partially addresss these issues
-by standardizing motif formats and providing a powerful set of tools for
-examining the properties of motifs. However the Bioconductor motif analysis
-landscape remains incomplete \cite{universalmotif}.
+sequences.
 
-The MEME Suite is a series of commandline and web-based tools that provide a variety
-of features for motif analysis such as *de-novo* motif discovery, enrichment
-testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
-utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
-users of the webserver implementation annually \cite{meme-usage-report-page,
-google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
-set of utilities to interrogate motif content across a broad range of biological
-contexts. However, several factors limit the full potential of these tools for
-use in data analysis. MEME Suite tools require carefully formatted inputs to
-achieve full functionality, yet few tools exist to simplify the process of data
-formatting, requiring instead that users write custom code to prepare their
-data, or prepare the inputs by hand, both of which have the potential to be
-error prone without rigorous testing. Furthermore, the output data from each
-MEME Suite tool often have complex structures that must be parsed to extract the
-full suite of information, again requiring users to write custom code for this
-task. Finally, although the data-generation capabilities of the MEME Suite are
-excellent, the tools lack powerful ways to visualize the results. Collectively,
-these factors act as barriers to adoption, preclude deeper analysis of the
-data, and limit communication of results to the scientific community.
+The MEME Suite is a widely utilized set of tools to interrogate motif content
+across a broad range of biological contexts \cite{meme-suite}. With over 22,000
+citations to date, and greater than 30,000 unique users of the webserver
+implementation annually, the MEME Suite has emerged as a standard tool in the
+field \cite{meme-usage-report-page, google-scholar-meme}. However, several
+factors limit the full potential of these tools for use in data analysis. MEME
+Suite tools require carefully formatted inputs to achieve full functionality,
+yet few tools exist to simplify the process of data formatting, requiring
+instead that users write custom code to prepare their data, or prepare the
+inputs by hand, both of which have the potential to be error prone without
+rigorous testing.  Furthermore, the output data from each MEME Suite tool often
+have complex structures that must be parsed to extract the full suite of
+information, again requiring users to write custom code for this task. Finally,
+although the data-generation capabilities of the MEME Suite are excellent, the
+tools lack powerful ways to visualize the results. Collectively, these factors
+act as barriers to adoption, preclude deeper analysis of the data, and limit
+communication of results to the scientific community.
 
@@
 data.frame is a standard R data.frame, it can be manipulated using base R or
-`tidyverse` functions, lowering the cognitive overhead to interfacing with these
+`tidyverse` functions, making it easier to interface with these
 structures.
@@
 **Figure?: showing how the data.frame connects to universalmotif**
-
-universalmotif data.frames are not limited to storing motifs from MEME Suite
-data, and can additionally be used to examine or manipulate the properties of
-any motif list. Using `universalmotif` functions to import motif data, the
-`memes` function `as_universalmotif_dataframe()` converts any `universalmotif`
-object into a `universalmotif` data.frame.  Manipulation of linked metadata
-columns will result in updating the linked `universalmotif` object when calling
-`update_motifs()` to retain the data.frame structure, or when calling
-`as_universalmotif()` to convert the data.frame back into `universalmotif`
-format. In this way, `memes` enables large-scale motif database manipulation &
-analysis within the preexisting R framework.
 
@@
 databases queried from the `MotifDb` package, or from files in a variety of
-non-meme file formats \cite{motifdb}. Once imported in `universalmotif` format
+non-meme file formats \cite{motifdb, universalmotif}. Once imported in `universalmotif` format
 the `memes` function `as_universalmotif_dataframe()` converts any
@@
 `as_universalmotif()` to convert the data.frame back into universalmotif format,
-discarding columns which cannot be stored within `universalmotif` format.
+discarding columns that cannot be stored within `universalmotif` format.
 Therefore, this data structure empowers users to import and analyze motif lists
-from a diverse set of sources for use with `memes` or simply for manipulating
+from a diverse set of sources for use with `memes` or for simple manipulation of
 motif metadata using common R tools. In this way, memes enables large-scale
@@
 transcription factor binding sites by the presence or absence of another
-factor). In these instances, motif analysis acts as a powerful tool to examine whether
-differences in motif content is associated with membership of distinct
-categories (e.g. motifs enriched at sites where both factors are bound vs singly
-bound). Although the MEME Suite allows differential enrichment testing, it does
+factor). Although the MEME Suite allows differential enrichment testing, it does
 not inherently provide a mechanism for analyzing groups of sequences in
-parallel, or performing motif analysis with any understanding of data
+parallel, or for performing motif analysis with any understanding of data
 categories. The `memes` framework enables "data-aware" motif analysis workflows
@@
 ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
-tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
+tool, and a worked example using `memes` to analyze ChIP-seq data can be found
 in the `memes` vignettes and the package website
@@
 change in chromatin accessibility observed before & after E93 binding. These
-data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
+data are an emblematic example of range-based genomic data (E93 ChIP peaks)
 containing additional groupings (the chromatin accessibility response following
@@
 ```
+
 Transcription factors often bind with sequence specificity to regulate activity
-of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
-DNA binding motifs for transcription factors can be a key step in identifying
-regulators of genomic elements. The TomTom tool is used to compare unknown
-motifs against a list of known motifs to identify matches, `memes` provides the
+of nearby genes. Therefore, comparison of *de-novo* motifs with known DNA
+binding motifs for transcription factors can be a key step in identifying
+transcriptional regulators. The TomTom tool is used to compare unknown motifs
+against a list of known motifs to identify matches, `memes` provides the
 `runTomTom` function as an interface to this tool. By passing the results from
-`runDreme` into `runTomTom` and searching within a database of *Drosophila* transcription
-factor motifs, we can identify candidate transcription factors which may bind
-the motifs associated with increasing and decreasing chromatin accessibility.
+`runDreme` into `runTomTom` and searching within a database of *Drosophila*
+transcription factor motifs, we can identify candidate transcription factors
+that may bind the motifs associated with increasing and decreasing chromatin
+accessibility.
 
@@
 ```
-Using this approach, the results can be visualized using 
-the `view_tomtom_hits()` function to visually inspect the matches
-assigned by TomTom, providing a simple way to assess the quality of the match. 
-A representative plot of these data is shown in \ref{fig:tomtom_example}, while
-the full results can be viewed in \ref{tab:tomtom_table}. We
-uncover that the motifs associated with decreasing sites are highly similar to
-the E93 motif. Conversely, motifs found in increasing sites match the
-transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
-data support the hypothesis that E93 binding to its motif may play a larger role
-in chromatin closing activity, while binding of the transcription factor br is
-associated with chromatin opening.
+
+Using this approach, the results can be visualized using  the
+`view_tomtom_hits()` function to visually inspect the matches assigned by
+TomTom, providing a simple way to assess the quality of the match.  A
+representative plot of these data is shown in \ref{fig:tomtom_example}, while
+the full results can be viewed in \ref{tab:tomtom_table}. This analysis revealed
+that the motifs associated with decreasing sites are highly similar to the E93
+motif. Conversely, motifs found in increasing sites match the transcription
+factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These data support
+the hypothesis that E93 binding to its motif may play a larger role in chromatin
+closing activity, while binding of the transcription factor br is associated
+with chromatin opening.
+
 ```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
@@
 that a motif is recovered *de-novo* is not ideal, as these approaches are
-"discovery based" as opposed to hypothesis driven. Another approach, implemented
+"discovery based". Another approach, implemented
 by the AME tool, is to search for known motif instances in target sequences and
@@
 \cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
-accepts a set of sequences as input an control sets, and will peform enrichment
+accepts a set of sequences as input and control sets, and will peform enrichment
 testing using a provided motif database for occurrances of each provided motif.
@@
 expression data with a motif database to remove entries for transcription
-factors which are not expressed. To do so, we import a *Drosophila*
+factors that are not expressed. To do so, we import a *Drosophila*
 transcription factor motif database generated by the Fly Factor Survey and
@@
 ```
+
 ```{r, echo = FALSE}
@@
 ```
+
 Next, we import a pre-filtered list of genes expressed in a timecourse of
-**Drosophila** wing development.
+*Drosophila* wing development.
+
 ```{r}
@@
 sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
-Additionally, AME detects several other transcription factor motifs which
+Additionally, AME detects several other transcription factor motifs that
 distinguish decreasing and increasing sites, providing additional clues to
-potential factors which also bind E93 ChIP peaks to affect chromatin
-accessibility.
+potential factors that also bind with E93 affect chromatin accessibility.
+
 ```{r}
@@
 ```
+
 ### Motif matching using runFimo
 
-A striking result from these analysis is that the E93 motif is so strongly
-enriched within E93 ChIP peaks which decrease in accessibility. Significance of
+A striking result from these analyses is that the E93 motif is so strongly
+enriched within E93 ChIP peaks that decrease in accessibility. Significance of
 motif enrichment can be driven by several factors, such as quality of the query
@@
 Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.
+
 ```{r}
@@
 ```
+
 ```{r, echo=F}
@@
 ```
+
 ```{r, fig.height=5, fig.width=3, echo=F}
@@
 with the flexible and extensive R/Bioconductor package landscape.
+
 ```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A.

### 2020-11-09 manuscript.Rmd
Finish DJM edits.

@@
 of multiomic data, users can identify different functional categories of genomic
-regions through integration with orthogonal datasets (e.g. categorizing
+regions through integration with orthogonal datasets, such as categorizing
 transcription factor binding sites by the presence or absence of another
-factor). Although the MEME Suite allows differential enrichment testing, it does
+factor. Although the MEME Suite allows differential enrichment testing, it does
 not inherently provide a mechanism for analyzing groups of sequences in
@@
 tooling for analyzing amino-acid motifs, which although fully supported by our
-current framework, may require extra tools which we have not considered.
+current framework, may require extra tools that we have not considered.
 

### 2020-11-11 manuscript.Rmd
Fix figure callouts.

@@
 output: 
- - word_document
+  bookdown::word_document2:
+    fig_caption: yes
 ---
 
-```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
+```{r setup, echo = FALSE, warning=FALSE, message=FALSE, include=FALSE}
 # Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
@@
 function. Visualization of the results reveals two distinct motifs associated
-with decreasing and increasing sites \ref{fig:dec_v_inc_dreme}. In this
+with decreasing and increasing sites (**Figure \@ref(fig:dec-v-inc-dreme)**). In this
 analysis, the *de-novo* discovered motifs within each category appear visually
@@
 coefficients for each set of motifs to quantitatively assess motif similarity
-\ref{fig:dec_v_inc_dreme}.
+(**Figure \@ref(fig:dec-v-inc-dreme)**).
 
-```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites.
+```{r dec-v-inc-dreme, fig.height=10, fig.width=10, echo=F, fig.cap="PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites.
 C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
@@
 TomTom, providing a simple way to assess the quality of the match.  A
-representative plot of these data is shown in \ref{fig:tomtom_example}, while
-the full results can be viewed in \ref{tab:tomtom_table}. This analysis revealed
+representative plot of these data is shown in **Figure \@ref(fig:tomtom-example)**, while
+the full results can be viewed in **Table \@ref(tab:tomtom-table)**. This analysis revealed
 that the motifs associated with decreasing sites are highly similar to the E93
 motif. Conversely, motifs found in increasing sites match the transcription
-factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These data support
+factors br, l(3)neo38, gl, and lola (**Table \@ref(tab:tomtom-table)**). These data support
 the hypothesis that E93 binding to its motif may play a larger role in chromatin
@@
 
-```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
+```{r tomtom-example, echo = F, fig.height=3, fig.width=7.5, fig.cap="Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
 plot_top_hit <- function(x){
@@
   gt(rowname_col = "name",
-     groupname_col = "category")
+     groupname_col = "category") %>% 
+  tab_header(title = "Best match results from TomTom")
      # this feature is in development branch??
-     # search SO: gt table reference
-     #lab = "tab:tomtom_table")
+     # search: R markdown gt table reference
+     # https://community.rstudio.com/t/how-to-set-a-reference-in-rmarkdown-to-a-gt-table/23781/2
+     #lab = "tab:tomtom-table")
 
@@
 Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
-sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
+sites, supporting the *de-novo* discovery results (**Figure \@ref(fig:ame-heatmap)**).
 Additionally, AME detects several other transcription factor motifs that
@@
 
-```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
+```{r ame-heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
 ame_vs_static %>% 
@@
 within each category, we demonstrate that Decreasing sites are more likely than
-increasing or static sites to contain an E93 motif \ref{fig:fimo}A. We extend
+increasing or static sites to contain an E93 motif (**Figure \@ref(fig:fimo)A**). We extend
 these observations by rederiving position-weight matrices from sequences
 matching the E93 motif within each category, allowing visual inspection of motif
-quality across groups \ref{fig:fimo}B. Notably, differences in quality at base
+quality across groups (**Figure \@ref(fig:fimo)B**). Notably, differences in quality at base
 positions 8-12 appear to distinguish increasing from decreasing motifs, where
@@
 while E93 motifs from increasing sites are more likely to have a T base pair at
-position 12 \ref{fig:fimo}B. Examination of bulk FIMO scores (where higher
+position 12 (**Figure \@ref(fig:fimo)B**). Examination of bulk FIMO scores (where higher
 scores represent motifs more similar to the reference) also reveals differences
 in E93 motif quality between groups, in particular, E93 motifs from Decreasing
-sites have higher scores \ref{fig:fimo}C. Together, these data demonstrate that
+sites have higher scores (**Figure \@ref(fig:fimo)C**). Together, these data demonstrate that
 a key distinguishing factor between whether a sites will decrease or increase in
@@
 
-```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A.
+```{r fimo, echo = F, fig.width=7.5, fig.height=9, fig.cap="A.
 Stacked barplot showing fraction of each chromatin response category containing E93 motif matches.
@@
 This manuscript was automatically generated using Rmarkdown within the
-`memes_docker` container (TODO: ADD VERSION). Its source code, raw data, and
+`memes_docker` container (**TODO: ADD VERSION**). Its source code, raw data, and
 instructions to reproduce all analysis can be found at

### 2020-11-11 R/plan.R

@@ -20,7 +20,7 @@ my_plan <- drake_plan(
     dplyr::select(symbol) %>% 
     unique(),
   
-  write_parsed_rna = readr::write_csv(filtered_rnaseq, path = file_out("data/wing_expressed_genes.csv")),
+  write_parsed_rna = readr::write_csv(filtered_rnaseq, file = file_out("data/wing_expressed_genes.csv")),
   
   max_expression_rnaseq = full_rnaseq %>% 
     dplyr::group_by(symbol) %>% 

### 2021-03-26 manuscript.Rmd
Grammar edits

@@
 distinguish decreasing and increasing sites, providing additional clues to
-potential factors that also bind with E93 affect chromatin accessibility.
+potential factors that also bind with E93 to affect chromatin accessibility.
 
@@
 sites have higher scores (**Figure \@ref(fig:fimo)C**). Together, these data demonstrate that
-a key distinguishing factor between whether a sites will decrease or increase in
+a key distinguishing factor between whether a site will decrease or increase in
 chromatin accessibility following E93 binding is the number and quality of E93
@@
 
-# Availability and Future developments:
+# Availability and Future Developments:
 

### 2021-03-26 manuscript.Rmd
Cleanup function references in text

@@
 
-The `memes` function `get_sequence` automates extraction of
+The `memes` function `get_sequence()` automates extraction of
 DNA sequence from genomic coordinates while simultaneously producing MEME Suite
@@
 
-The `ame_plot_heatmap` function in `memes` returns a `ggplot2` formatted heatmap of
+The `ame_plot_heatmap()` function in `memes` returns a `ggplot2` formatted heatmap of
 statistical significance of motif enrichment. If AME is used to examine motif
-content of multiple groups of sequences, the `ame_plot_heatmap` function can also
+content of multiple groups of sequences, the `ame_plot_heatmap()` function can also
 return a plot comparing motif significance within multiple groups. Several
@@
 during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
-utilize ChIP-seq peak calls for E93 which have been annotated according to the
-change in chromatin accessibility observed before & after E93 binding. These
+utilize ChIP-seq peak calls for E93 that have been annotated according to the
+change in chromatin accessibility observed before and after E93 binding. These
 data are an emblematic example of range-based genomic data (E93 ChIP peaks)
@@
 
-Using the `get_sequence` function, GRanges objects are converted into
+Using the `get_sequence()` function, GRanges objects are converted into
 DNAStringSet outputs.
@@
 ```
-In order to perform analysis within different groups of peaks, the GRanges object can be `split` on a metadata column before input to `get_sequence`.
+In order to perform analysis within different groups of peaks, the GRanges object can be `split()` on a metadata column before input to `get_sequence()`.
 ```{r}
@@
 The DREME tool can be used to discover short, novel motifs present in a set of
-input sequences relative to a control set. The `runDreme` command is the `memes`
-interface to the DREME tool. `runDreme` syntax enables users to produce complex
+input sequences relative to a control set. The `runDreme()` command is the `memes`
+interface to the DREME tool. `runDreme()` syntax enables users to produce complex
 discriminative analysis designs using intuitive syntax. Examples of possible
@@
 
-These data are readily visualized using the `universalmotif::view_motifs`
+These data are readily visualized using the `universalmotif::view_motifs()`
 function. Visualization of the results reveals two distinct motifs associated
@@
 compare groups of motifs based on all pairwise similarity metrics. We utilized
-the `universalmotif::compare_motifs` function to compute Pearson correlation
+the `universalmotif::compare_motifs()` function to compute Pearson correlation
 coefficients for each set of motifs to quantitatively assess motif similarity
@@
 against a list of known motifs to identify matches, `memes` provides the
-`runTomTom` function as an interface to this tool. By passing the results from
-`runDreme` into `runTomTom` and searching within a database of *Drosophila*
+`runTomTom()` function as an interface to this tool. By passing the results from
+`runDreme()` into `runTomTom()` and searching within a database of *Drosophila*
 transcription factor motifs, we can identify candidate transcription factors
@@
 
-`runAme` syntax is identical to `runDreme` in that discriminative designs can be
+`runAme()` syntax is identical to `runDreme()` in that discriminative designs can be
 constructed by calling list entries by name. Data can be visualized using
@@
 peaks, we scan all E93 ChIP peaks for matches to the Fly Factor Survey E93 motif
-using `runFimo`. Results are returned as a GRanges object containing the
+using `runFimo()`. Results are returned as a GRanges object containing the
 positions of each motif. Setting `skip_matched_sequence` to `FALSE` adds the

### 2021-03-26 manuscript.Rmd
add notes

@@
 syntax. **TODO: Depending on length restrictions, end here**
+
+**TODO: just end above since universalmotif_df exists now**
 
@@
 
-The `universalmotif` data structure is a powerful R/Bioconductor representation
-for motif matrices and their associated metadata \cite{universalmotif}. However,
-`universalmotif` objects are not ideal for performing bulk manipulation or
-queries of metadata, nor do they support complex data structures such as nested
-elements. R data.frames support nesting arbitrary data types within columns, but
-are not sufficient to properly store motif matrices. In order to create an
-extensible, nested data structure for tracking MEME Suite data, `memes`
-leverages the power of R data.frames and the `universalmotif` format by nesting
-`universalmotif` objects within data.frames. These universalmotif data.frames
-store information about motifs along rows, where each column stores metadata
-associated with a motif, including a `motif` column holding a `universalmotif`
-representation of that motif, which can be manipulated by altering columns of
-the data.frame (described in detail below). Adding one-to-many relationships within this
-structure is trivial, as additional unviersalmotif data.frames can be nested
-within eachother. These universalmotif data.frames form the basis for a majority
-of `memes` data outputs. These structures are also valid input types to `memes`
-functions, and when used as such, output data are appended as new columns to the
-input data, ensuring data provenance. Finally, because the universalmotif
-data.frame is a standard R data.frame, it can be manipulated using base R or
-`tidyverse` functions, making it easier to interface with these
-structures.
+The `universalmotif_df` data structure is a powerful R/Bioconductor
+representation for motif matrices and their associated metadata
+\cite{universalmotif}. `universalmotif_df` format is an alternative
+representation of `universalmotif` objects where motifs are stored along rows of
+a data.frame, and columns store metadata associated with each motif. Adding
+one-to-many relationships within this structure is trivial, as additional
+`unviersalmotif_df`s can be nested within eachother. These `universalmotif_df`s
+form the basis for a majority of `memes` data outputs. These
+structures are also valid input types to `memes` functions, and when used as
+such, output data are appended as new columns to the input data, ensuring data
+provenance.
 
 **Figure?: showing how the data.frame connects to universalmotif**
+
+## TODO:
+ - `remove_duplicate_motifs`
+ -
 
@@
 DNA sequence from genomic coordinates while simultaneously producing MEME Suite
-formatted fasta headers. `get_sequence` accepts genomic-range based inputs in
+formatted fasta headers. `get_sequence()` accepts genomic-range based inputs in
 `GenomicRanges` format, the *de-facto* standard for genomic coordinate
@@
 using `GenomicRanges`, users can split peak data on a metadata column using the
-base R `split` function, and then use this result as input to `get_sequence`, which
+base R `split()` function, and then use this result as input to `get_sequence()`, which
 will produce a list of `BioStrings` objects where each entry is named after the
@@
 to identify the best match. Visual inspection of motif comparison data is a key
-step in assessing the quality and accuracy of a match. The `view_tomtom_hits`
+step in assessing the quality and accuracy of a match. The `view_tomtom_hits()`
 function allows users to compare query motifs with the list of potential matches
 as assigned by TomTom, similar to the commandline behavior. The
-`force_best_match` function allows users to reassign the TomTom best match
+`force_best_match()` function allows users to reassign the TomTom best match
 to a lower-ranked (but still statistically significant) match in order to

### 2021-03-26 manuscript.Rmd
remove stuff about df

@@
 data, providing users quick access to all relevant data structures with simple
-syntax. **TODO: Depending on length restrictions, end here**
-
-**TODO: just end above since universalmotif_df exists now**
-
-In addition to wrapping MEME Suite utilities, `memes` extends the functionality
-of `universalmotif` to provide base R and `tidyverse` compatible syntax for
-manipulating and analyzing motif metadata. These features allow users to
-integrate external data with motif databases to aid analysis, such as filtering
-a set of transcription factor motifs based on gene expression in the tissue of
-interest, then using that filtered set to search for motifs associated with
-expressed factors. By maintaining compatability with `universalmotif`, `memes`
-outputs can be readily used in a diverse set of tools beyond the `memes`
-framework.
-
-`memes` also provides simple utilities for data visualization where appropriate.
-Providing R representations of MEME Suite data objects enables generation of
-novel data visualizations using the well-established tools available in the R
-landscape. Therefore, `memes` empowers users to deeply analyze and visualize
-motif analysis data.
+syntax.
 
@@
 `unviersalmotif_df`s can be nested within eachother. These `universalmotif_df`s
-form the basis for a majority of `memes` data outputs. These
-structures are also valid input types to `memes` functions, and when used as
-such, output data are appended as new columns to the input data, ensuring data
-provenance.
-
-**Figure?: showing how the data.frame connects to universalmotif**
-
-## TODO:
- - `remove_duplicate_motifs`
- -
-
-The universalmotif data.frames generated by memes are not limited to storing
-motifs from MEME Suite data. They can also be used to manipulate or store other
-descriptive properties of motifs from multiple sources. `universalmotif`
-provides several import functions which can convert motifs from external
-databases queried from the `MotifDb` package, or from files in a variety of
-non-meme file formats \cite{motifdb, universalmotif}. Once imported in `universalmotif` format
-the `memes` function `as_universalmotif_dataframe()` converts any
-`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
-data stored in this format, if changes are made to the linked metadata columns,
-the linked universalmotif object can be updated by calling `update_motifs()` to
-retain the data.frame structure for further analysis, or by calling
-`as_universalmotif()` to convert the data.frame back into universalmotif format,
-discarding columns that cannot be stored within `universalmotif` format.
-Therefore, this data structure empowers users to import and analyze motif lists
-from a diverse set of sources for use with `memes` or for simple manipulation of
-motif metadata using common R tools. In this way, memes enables large-scale
-motif database manipulation & analysis within the preexisting R framework.
+form the basis for a majority of `memes` data outputs. These structures are also
+valid input types to `memes` functions, and when used as such, output data are
+appended as new columns to the input data, ensuring data provenance. Finally,
+because `universalmotif_df`s are extensions of base R data.frames, they can be
+manipulated using base R and `tidyverse` workflows.  Therefore, `memes` data
+integrate seamlessly with common R workflows.
 

### 2021-03-29 manuscript.Rmd
Grammar fixes, some notes

@@
 suppressPackageStartupMessages(library(ComplexHeatmap))
+# TODO:
+# use Rstudio pkg repo?
+# Need to install: ComplexHeatmap, drake, readxl, gt, bookdown
 drake::r_make()
@@
 DREME return de-novo discovered motifs from query sequences along with
-statistical information about their enrichment. In this instance, a position
+statistical information about their enrichment \cite{meme, dreme}. In this instance, a position
 frequency matrix (PFM) of the discovered motif can be represented as a matrix,
@@
 contain metadata as previously described) can have a one-to-many relationship
-with other motif matrices (again with their own metadata). Thus an ideal
+with other motif matrices (again with their own metadata) \cite{tomtom}. Thus an ideal
 representation for these data is one that can hold an unlimited number of motif
@@
 
-Motif analysis is often employed in ChIP-seq analysis, in which data is stored
+Motif analysis is often employed in ChIP-seq analysis, in which data are stored
 as genomic coordinates rather than sequence. However, MEME Suite tools are
 designed to work with sequences. While existing tools such as bedtools can
-extract DNA sequence from gneomic coordinates, some MEME tools require fasta
+extract DNA sequence from genomic coordinates, some MEME tools require fasta
 headers to be specifically formatted. As a result, users must write custom code
@@
 
-Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
-sequences relative to another set, can uncover biologically relevant motifs
-associated with membership to distinct categories. For example, during analysis
-of multiomic data, users can identify different functional categories of genomic
-regions through integration with orthogonal datasets, such as categorizing
-transcription factor binding sites by the presence or absence of another
-factor. Although the MEME Suite allows differential enrichment testing, it does
-not inherently provide a mechanism for analyzing groups of sequences in
-parallel, or for performing motif analysis with any understanding of data
-categories. The `memes` framework enables "data-aware" motif analysis workflows
-by allowing named lists of `Biostrings` objects as input to each function. If
-using `GenomicRanges`, users can split peak data on a metadata column using the
-base R `split()` function, and then use this result as input to `get_sequence()`, which
-will produce a list of `BioStrings` objects where each entry is named after the
-data categories from the split column. When a list is used as input to a `memes`
-function, it runs the corresponding MEME Suite tool for each object in the list.
-Users can also pass the name(s) of a category to the `control` argument to
-enable differential analysis of the remaining list members against the
-control category sequences. In this manner, `memes` enables complex data-aware
-differential motif analysis workflows using simple syntax to extend the
-capabilities of the MEME Suite.
+Discriminative (or "differential") motif analysis, in which motifs are
+discovered in one set of sequences relative to another set, can uncover
+biologically relevant motifs associated with membership to distinct categories.
+For example, during analysis of multiomic data, users can identify different
+functional categories of genomic regions through integration with orthogonal
+datasets, such as categorizing transcription factor binding sites by the
+presence or absence of another factor \cite{???}. Although the MEME Suite allows
+differential enrichment testing, it does not inherently provide a mechanism for
+analyzing groups of sequences in parallel, or for performing motif analysis with
+an understanding of data categories. The `memes` framework enables "data-aware"
+motif analysis workflows by allowing named lists of `Biostrings` objects as
+input to each function. If using `GenomicRanges`, users can split peak data on a
+metadata column using the base R `split()` function, and then use this result as
+input to `get_sequence()`, which will produce a list of `BioStrings` objects
+where each entry is named after the data categories from the split column. When
+a list is used as input to a `memes` function, it runs the corresponding MEME
+Suite tool for each object in the list. Users can also pass the name(s) of a
+category to the `control` argument to enable differential analysis of the
+remaining list members against the control category sequences. In this manner,
+`memes` enables complex data-aware differential motif analysis workflows using
+simple syntax to extend the capabilities of the MEME Suite.
 
@@
 ```{r, eval=F}
-# use all sequences vs shuffled background
+# Use all sequences vs shuffled background
+# Produces:
+#  - All Sequences vs Shuffled All Sequences
 runDreme(all_sequences, control = "shuffle")
 
-# for each response category, discover motifs against a shuffled background set
+# For each response category, discover motifs against a shuffled background set
 # Produces:
@@
 # Use the "Static" response category as the control set to discover motifs 
-# enriched in each remaining category relative to static sites,
+# enriched in each remaining category relative to static sites
 # Produces:
@@
 # Combine the "Static" and "Increasing" sequences and use as a background set to
-# discover motifs enriched in the remaining categories relative
+# discover motifs enriched in the remaining categories 
 # Produces:
@@
 that a motif is recovered *de-novo* is not ideal, as these approaches are
-"discovery based". Another approach, implemented
+less sensitive than targeted searches. Another approach, implemented
 by the AME tool, is to search for known motif instances in target sequences and
@@
 fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
-  as_universalmotif_dataframe()
+  {
+    # TODO: why aren't names added by default on import anymore?
+    names(.) <- purrr::map_chr(., ~{.x['name']})
+    .
+  } %>% 
+  to_df() %>% 
+  # TODO: print method doesn't play nice w/ Rmd
+  data.frame
 ```
@@
 equally well on these data structures), then convert the data.frame back into
-`universalmotif` format using `as_universalmotif`. This filtering step removes
+`universalmotif` format using `to_list()`. This filtering step removes
 `r (n_fly_factor_unexpressed / nrow(fly_factor)) %>% {. * 100} %>% round()`% of
@@
   dplyr::filter(altname %in% wing_expressed_genes$symbol) %>%
-  as_universalmotif()
+  to_list()
 ```
@@
     m$e93 <- e93_flyfactor
-    return(m)
+    m
   } %>% 
@@
 [snystrom.github.io/memes](snystrom.github.io/memes). The `memes_docker`
-containers are available from DockerHub
+
 [https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),

### 2021-03-30 manuscript.Rmd
Update manuscript to have correct citaions, etc. Added a template docx.

@@
     fig_caption: yes
+    reference_docx: template.docx
+
+bibliography: references.bib
+csl: vancouver-brackets.csl
 ---
@@
 suppressPackageStartupMessages(library(ComplexHeatmap))
-# TODO:
-# use Rstudio pkg repo?
+suppressPackageStartupMessages(library(universalmotif))
+# TODO add auto install chunk:
 # Need to install: ComplexHeatmap, drake, readxl, gt, bookdown
+# version pin these?
 drake::r_make()
@@
 
-# Abstract
+# Abstract {-}
 
@@
 
-# Introduction
+# Introduction {-}
 
@@
 characterizing DNA elements based on shared transcription factor binding
-sequences (TODO: ref kinase motif:
-https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
+sequences [@Kemp1990]. Thus the ability to identify,
 classify, and compare motifs represents a key step in the analysis of biological
@@
 The MEME Suite is a widely utilized set of tools to interrogate motif content
-across a broad range of biological contexts \cite{meme-suite}. With over 22,000
+across a broad range of biological contexts [@Bailey2009]. With over 22,000
 citations to date, and greater than 30,000 unique users of the webserver
 implementation annually, the MEME Suite has emerged as a standard tool in the
-field \cite{meme-usage-report-page, google-scholar-meme}. However, several
+field [@meme_usage; @meme_citations]. However, several
 factors limit the full potential of these tools for use in data analysis. MEME
@@
 
-## Design & Implementation
+## Design & Implementation {-}
 
-### Core Utilities
+### Core Utilities {-}
 
@@
 required by the tool. Sequence inputs are accepted in `BioStrings` format, an
-R/Bioconductor package for storing biopolymer sequence data \cite{BioStrings}.
+R/Bioconductor package for storing biopolymer sequence data [@Biostrings].
 Motif inputs are passed as `universalmotif` objects, another R/Bioconductor
 package for representing motif matrices and their associated metadata
-\cite{universalmotif}. `memes` `run` functions will also accept paths to files
+[@universalmotif]. `memes` `run` functions will also accept paths to files
 on disk, such as fasta files for sequence inputs, and meme format files for
@@
 
-### Output Data Structures
+### Output Data Structures {-}
 
@@
 
-### structured data.frames hold multidimensional output data
+### structured data.frames hold multidimensional output data {-}
 
@@
 DREME return de-novo discovered motifs from query sequences along with
-statistical information about their enrichment \cite{meme, dreme}. In this instance, a position
+statistical information about their enrichment [@Bailey1994; @Bailey2011]. In this instance, a position
 frequency matrix (PFM) of the discovered motif can be represented as a matrix,
@@
 contain metadata as previously described) can have a one-to-many relationship
-with other motif matrices (again with their own metadata) \cite{tomtom}. Thus an ideal
+with other motif matrices (again with their own metadata) [@Gupta2007]. Thus an ideal
 representation for these data is one that can hold an unlimited number of motif
@@
 representation for motif matrices and their associated metadata
-\cite{universalmotif}. `universalmotif_df` format is an alternative
+[@universalmotif]. `universalmotif_df` format is an alternative
 representation of `universalmotif` objects where motifs are stored along rows of
@@
 
-### Support for genomic range-based data
+### Support for genomic range-based data {-}
 
@@
 `GenomicRanges` format, the *de-facto* standard for genomic coordinate
-representation in R \cite{genomicranges}. Other common genomic coordinate
+representation in R [@GRanges]. Other common genomic coordinate
 representations, such as bed format, are easily imported as `GenomicRanges`
@@
 
-### Data-aware motif analysis
+### Data-aware motif analysis {-}
 
@@
 datasets, such as categorizing transcription factor binding sites by the
-presence or absence of another factor \cite{???}. Although the MEME Suite allows
+presence or absence of another factor [@Nystrom2020]. Although the MEME Suite allows
 differential enrichment testing, it does not inherently provide a mechanism for
@@
 
-## Data visualization
+## Data visualization {-}
 The MEME Suite provides a small set of data visualizations that have limited
@@
 to identify the best match. Visual inspection of motif comparison data is a key
-step in assessing the quality and accuracy of a match. The `view_tomtom_hits()`
+step in assessing the quality and accuracy of a match [@Gupta2007]. The `view_tomtom_hits()`
 function allows users to compare query motifs with the list of potential matches
@@
 The AME tool searches for enrichment of motifs within a set of experimental
-sequences relative to a control set. The meaningful result from this tool is the
+sequences relative to a control set [@Buske2010]. The meaningful result from this tool is the
 statistical parameter (for example, a p-value) associated with the significance
@@
 
-## Containerized analysis maximizes availability and facilitates reproducibility
+## Containerized analysis maximizes availability and facilitates reproducibility {-}
 `memes` relies on a locally installed version of the MEME Suite which is
@@
 unsupported operating systems, and to facilitate reproducible motif analysis
-using `memes`, we have also developed docker and singularity (**TODO: NOTE:**
-singularity is broken for `rocker/rstudio`, so maybe omit this until fixed)
-containers containing a preinstalled version of the MEME Suite along with an
-R/Bioconductor analysis environment including the most recent version of `memes`
-and its dependencies. As new container versions are released, they are
-version-tagged and stored to ensure reproducibility while allowing updates to
-the container.
+using `memes`, we have also developed a docker container with a
+preinstalled version of the MEME Suite along with an R/Bioconductor analysis
+environment including the most recent version of `memes` and its dependencies.
+As new container versions are released, they are version-tagged and stored to
+ensure reproducibility while allowing updates to the container.
 
-# Results
+# Results {-}
 
@@
 binding of the transcription factor E93 to changes in chromatin accessibility
-during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
+during *Drosophila* wing development [@Nystrom2020]. Here, we
 utilize ChIP-seq peak calls for E93 that have been annotated according to the
@@
 
-## Denovo motif analysis
+## Denovo motif analysis {-}
 The DREME tool can be used to discover short, novel motifs present in a set of
@@
                           col = colors, 
-                          name = "Pearson Correlation", 
+                          name = "Pearson's\nR", 
                           column_names_rot = 45,
-                          column_title = paste(name, "Motif Similarity")
+                          column_title = paste(name, "Motif Similarity"), raster_magick_filter = 
                           )
@@
   # capture heatmaps as grobs
-  purrr::map(~{grid::grid.grabExpr(draw(.x))}) %>% 
+  # add some right padding
+  # https://jokergoo.github.io/ComplexHeatmap-reference/book/a-list-of-heatmaps.html#manually-increase-space-around-the-plot
+  purrr::map(~{grid::grid.grabExpr(draw(.x, padding = unit(c(0,5,0,10), "mm")))}) %>% 
   patchwork::wrap_plots(nrow = 1) 
@@
 
-```{r}
+```{r, message = FALSE}
 dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
@@
 # Plotting every hit:
-# decided to just use a representative sample insted
+# decided to just use a representative sample instead
 #dreme_vs_static_tomtom %>% 
@@
   gt(rowname_col = "name",
-     groupname_col = "category") %>% 
-  tab_header(title = "Best match results from TomTom")
+     groupname_col = "category")
+  # Caption: Table 1: Summary of all TomTom matches for denovo motifs discovered in E93 responsive ChIP peaks.
+
+  #tab_header(title = "Best match results from TomTom")
      # this feature is in development branch??
+     # remotes::install_github("rstudio/gt", ref="eff3be7384365a44459691e49b9b740420cd0851")
      # search: R markdown gt table reference
@@
 
-## Motif enrichment testing using runAme
+## Motif enrichment testing using runAme {-}
 
@@
 test for their overrepresentation relative to a background set of sequences
-\cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
-accepts a set of sequences as input and control sets, and will peform enrichment
+[@Buske2010]. The `runAme()` function is the `memes` interface to the AME tool. It
+accepts a set of sequences as input and control sets, and will perform enrichment
 testing using a provided motif database for occurrances of each provided motif.
@@
 probability of identifying biologically relevant transcription factor candidates
-\cite{hector franco's STAR paper, others?}.
+[@Nystrom2020; @Malladi2020].
 
-The `memes` universalmotif data.frame structure can be used to integrate
+The `universalmotif_df` structure can be used to integrate
 expression data with a motif database to remove entries for transcription
@@
 transcription factor motif database generated by the Fly Factor Survey and
-convert to universalmotif data.frame format \cite{fly factor}. In this database,
+convert to `universalmotif_df` format [@Zhu2011]. In this database,
 the `altname` column stores the gene symbol.
@@
 fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
-  {
-    # TODO: why aren't names added by default on import anymore?
-    names(.) <- purrr::map_chr(., ~{.x['name']})
-    .
-  } %>% 
-  to_df() %>% 
-  # TODO: print method doesn't play nice w/ Rmd
-  data.frame
+  # Add motif names to the list entry
+  setNames(., purrr::map_chr(., ~{.x['name']})) %>% 
+  to_df()
 ```
@@
 ```{r, echo = FALSE}
-head(fly_factor, 3)
+# Convert to tibble so printing looks better
+# TODO: use pillar to make a custom print for final publication?
+head(tibble::as_tibble(fly_factor) %>% dplyr::select(!tidyselect::contains("motif"), motif), 3)
 ```
@@
 
-### Motif matching using runFimo
+### Motif matching using runFimo {-}
 
@@
 FIMO allows users to match motifs in input sequences while returning information
-about the quality of each match in the form of a quantitative score \cite{fimo}.
+about the quality of each match in the form of a quantitative score [@Grant2011].
 
@@
 
-Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.
+Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect [@plyranges].
 
@@
 
-# Availability and Future Developments:
+# Availability and Future Developments {-}
 
@@
 stored in the package vignettes, and also available at the package website:
-[snystrom.github.io/memes](snystrom.github.io/memes). The `memes_docker`
-
-[https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
-and their source code is hosted at github:
+[snystrom.github.io/memes](snystrom.github.io/memes).
+The `memes_docker` container  is available on dockerhub: [https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
+and the container source code is hosted at github:
 [https://github.com/snystrom/memes_docker](https://github.com/snystrom/memes_docker).
@@
 This manuscript was automatically generated using Rmarkdown within the
-`memes_docker` container (**TODO: ADD VERSION**). Its source code, raw data, and
+`memes_docker` container. Its source code, raw data, and
 instructions to reproduce all analysis can be found at
@@
 
-# Acknowledgements
-We would like to acknowledge Megan Justice, and members of the McKay lab for
-helpful comments and feedback.
+# Acknowledgements {-}
+We would like to acknowledge Megan Justice, Garrett Graham, and members of the
+McKay lab for helpful comments and feedback. Benjamin Jean-Marie Tremblay added
+additional features to universalmotif which greatly benefitted the development
+of memes.
+
+# References {-}
+
+<div id="refs"></div>

### 2021-04-06 manuscript.Rmd
Final edits from DJM

@@
 units within biopolymers that can act as interaction surfaces, create structure,
-or have enzymatic activity. Identification of shared motifs across multiple
-sequences can provide evidence for shared function, for example, identification
+or contribute to enzymatic activity. Identification of similar motifs across multiple
+sequences can provide evidence for shared function, such as identification
 of kinase substrates based on similarities in phosphorylation site sequence, or
@@
 iterate rapidly during analysis. Here we present several examples of how `memes`
-allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
+allows novel analyses of transcription factor binding profile (ChIP-seq) and
+open chromatin profile (FAIRE-seq) data by facilitating seamless
 interoperability between MEME Suite tools and the broader R package landscape.
@@
 
-The `ame_plot_heatmap()` function in `memes` returns a `ggplot2` formatted heatmap of
+The `plot_ame_heatmap()` function in `memes` returns a `ggplot2` formatted heatmap of
 statistical significance of motif enrichment. If AME is used to examine motif
-content of multiple groups of sequences, the `ame_plot_heatmap()` function can also
+content of multiple groups of sequences, the `plot_ame_heatmap()` function can also
 return a plot comparing motif significance within multiple groups. Several
@@
 In order to discover motifs associated with dynamic changes to chromatin
-accessibility, we use static sites as the control set to discover motifs
-enriched in both increasing and decreasing sites.
+accessibility, we use open chromatin sites that do not change in accessibility
+(i.e. "static sites") as the control set to discover motifs enriched in either
+increasing or decreasing sites.
 
@@
 analysis, the *de-novo* discovered motifs within each category appear visually
-similar to each other, however, the MEME Suite does not provide a mechanism to
+similar to each other; however, the MEME Suite does not provide a mechanism to
 compare groups of motifs based on all pairwise similarity metrics. We utilized
@@
 motif. Conversely, motifs found in increasing sites match the transcription
-factors br, l(3)neo38, gl, and lola (**Table \@ref(tab:tomtom-table)**). These data support
+factors Br, L(3)neo38, Gl, and Lola (**Table \@ref(tab:tomtom-table)**). These data support
 the hypothesis that E93 binding to its motif may play a larger role in chromatin
@@
 accepts a set of sequences as input and control sets, and will perform enrichment
-testing using a provided motif database for occurrances of each provided motif.
+testing using a provided motif database for occurrences of each provided motif.
 
@@
 constructed by calling list entries by name. Data can be visualized using
-`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
-Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
+`plot_ame_heatmap()`, revealing that the E93 motif is strongly enriched at
+Decreasing sites, while l(3)neo38, lola, and br motifs are enriched in Increasing
 sites, supporting the *de-novo* discovery results (**Figure \@ref(fig:ame-heatmap)**).
@@
   dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
-  ame_plot_heatmap(group = category, id = motif_alt_id) +
+  plot_ame_heatmap(group = category, id = motif_alt_id) +
     labs(x = "Transcription Factor",
@@
 group relative to other groups. These questions can be explored directly by
-identifying motif occurrances in target regions and examining their properties.
+identifying motif occurrences in target regions and examining their properties.
 FIMO allows users to match motifs in input sequences while returning information

### 2021-04-06 manuscript.Rmd
hard code table reference. Add note indicating import functions are made by me.

@@
 
-`memes` provides data import functions for each supported MEME Suite tool, which
-import these data as modified R data.frames (described in detail below). These
-functions can be called directly by users to import data previously generated by
-the commandline or webserver versions of the MEME Suite for use in R. These
-import functions also underlie the import step internal to each of the `run`
-functions.
+`memes` provides custom-built data import functions for each supported MEME
+Suite tool, which import these data as modified R data.frames (described in
+detail below). These functions can be called directly by users to import data
+previously generated by the commandline or webserver versions of the MEME Suite
+for use in R. These import functions also underlie the import step internal to
+each of the `run` functions, ensuring consistent performance.
 
-### structured data.frames hold multidimensional output data {-}
+### Structured data.frames hold multidimensional output data {-}
 
@@
 representative plot of these data is shown in **Figure \@ref(fig:tomtom-example)**, while
-the full results can be viewed in **Table \@ref(tab:tomtom-table)**. This analysis revealed
+the full results can be viewed in **Table 1**. This analysis revealed
 that the motifs associated with decreasing sites are highly similar to the E93
 motif. Conversely, motifs found in increasing sites match the transcription
-factors Br, L(3)neo38, Gl, and Lola (**Table \@ref(tab:tomtom-table)**). These data support
+factors Br, L(3)neo38, Gl, and Lola (**Table 1**). These data support
 the hypothesis that E93 binding to its motif may play a larger role in chromatin

### 2021-04-07 manuscript.Rmd
add install chunk for uninstalled packages

@@
 # Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
+
+# Install pkgs not preinstalled in container
+need_install <- c("ComplexHeatmap", "drake", "readxl", "gt", "bookdown")
+
+lapply(need_install, function(x){
+  if (!require(x, character.only = TRUE)) {
+    BiocManager::install(x, dependencies = TRUE)
+  }
+})
+
+# Load pkgs
 suppressPackageStartupMessages(library(memes))
@@
 suppressPackageStartupMessages(library(universalmotif))
-# TODO add auto install chunk:
-# Need to install: ComplexHeatmap, drake, readxl, gt, bookdown
-# version pin these?
+
 drake::r_make()
@@
 using `runFimo()`. Results are returned as a GRanges object containing the
-positions of each motif. Setting `skip_matched_sequence` to `FALSE` adds the
-sequence of the detected motif to the output.
+positions of each motif.
 
 ```{r}
-e93_fimo <- runFimo(all_sequences, fly_factor_expressed$Eip93F_SANGER_10, skip_matched_sequence = FALSE, thresh = 1e-3)
+e93_fimo <- runFimo(all_sequences, fly_factor_expressed$Eip93F_SANGER_10, thresh = 1e-3)
 ```
@@
 instructions to reproduce all analysis can be found at
-[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/).
+[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/). Data used in this manuscript were from: [GSE141738](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141738).
 

### 2021-04-08 manuscript.Rmd
Add link to GEO for data
Update manuscript to reflect acceptance to bioconductor

@@
 seamless R interface to the MEME Suite. `memes` provides a novel "data aware"
-interface to these tools, enabling rapid and complex discriminative motif analysis workflows.
-In addition to interfacing with popular MEME Suite tools, `memes` leverages
-existing R/Bioconductor data structures to store the complex, multidimensional
-data returned by MEME Suite tools for rapid data access and manipulation.
-Finally, `memes` provides data visualization capabilities to facilitate
-communication of results. `memes` is available at
+interface to these tools, enabling rapid and complex discriminative motif
+analysis workflows.  In addition to interfacing with popular MEME Suite tools,
+`memes` leverages existing R/Bioconductor data structures to store the complex,
+multidimensional data returned by MEME Suite tools for rapid data access and
+manipulation.  Finally, `memes` provides data visualization capabilities to
+facilitate communication of results. `memes` is available as a Bioconductor
+package at
+[https://bioconductor.org/packages/memes](https://bioconductor.org/packages/memes),
+and the source code can be found at
 [github.com/snystrom/memes](github.com/snystrom/memes).
@@
 The MEME Suite is a widely utilized set of tools to interrogate motif content
-across a broad range of biological contexts [@Bailey2009]. With over 22,000
+across a broad range of biological contexts [@Bailey2009]. With over 25,000
 citations to date, and greater than 30,000 unique users of the webserver
@@
 
-The `memes` package source code and installation instructions are available on
-github: [github.com/snystrom/memes](github.com/snystrom/memes). Documentation is
-stored in the package vignettes, and also available at the package website:
-[snystrom.github.io/memes](snystrom.github.io/memes).
-The `memes_docker` container  is available on dockerhub: [https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
+`memes` is part of Bioconductor. Installation instructions can be found at
+[https://bioconductor.org/packages/memes](https://bioconductor.org/packages/memes).
+The `memes` package source code is available on github:
+[github.com/snystrom/memes](github.com/snystrom/memes). Documentation is stored
+in the package vignettes, and also available at the package website:
+[snystrom.github.io/memes](snystrom.github.io/memes). The `memes_docker`
+container  is available on dockerhub:
+[https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
 and the container source code is hosted at github:
@@
 instructions to reproduce all analysis can be found at
-[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/). Data used in this manuscript were from: [GSE141738](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141738).
+[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/). 
+Data used in this manuscript can be found on GEO at the following accession number:
+[GSE141738](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141738).
 

### 2021-07-13 manuscript.Rmd
add new refs, edit "complex" out to respond to reviewer comment.

@@
     reference_docx: template.docx
+    keep_tex: true
 
@@
 ```{r setup, echo = FALSE, warning=FALSE, message=FALSE, include=FALSE}
+# Don't remove tex intermediate files
+options(tinytex.clean = FALSE)
 # Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
@@
 Here we present `memes`, a motif analysis package that provides a seamless
-interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
+interface to a selection of popular MEME Suite tools within R. Six MEME Suite tools are contained within the `memes` wrapper (Table 1). `memes` uses base R and Bioconductor
 data types for data input and output, facilitating better integration with
-common analysis tools like the `tidyverse` and other Bioconductor packages.
+common analysis tools like the `tidyverse` and other Bioconductor packages [@rcore; @bioconductor; @tidyverse].
 Unlike the commandline implementation, `memes` outputs also function as inputs
@@
 write custom code to extract the necessary data. Finally, the data types
-contained in MEME Suite outputs are complex and multidimensional, and thus
+contained in MEME Suite outputs are context-specific and multidimensional, and thus
 require special data structures to properly organize the data in memory.
@@
 
-Data returned from MEME Suite tools are complex and multidimensional, making
+Data returned from MEME Suite tools are often multidimensional, making
 them difficult to represent in a simple data structure. For example, MEME and
@@
 compares one motif to a series of several motifs to identify possible matches,
-produce an additional layer of complexity such that input matrices (which
+produce an additional layer of information such that input matrices (which
 contain metadata as previously described) can have a one-to-many relationship
@@
 representation for these data is one that can hold an unlimited number of motif
-matrices and their metadata, contain complex hierarchies, and be easily
+matrices and their metadata, retain heirarchical relationships, and be easily
 manipulated using standard analysis tools.
@@
 remaining list members against the control category sequences. In this manner,
-`memes` enables complex data-aware differential motif analysis workflows using
+`memes` enables data-aware differential motif analysis workflows using
 simple syntax to extend the capabilities of the MEME Suite.
@@
 DNA binding) whose membership may be influenced by differential motif content.
-We show how `memes` syntax allows complex analysis designs, how `memes`
+We show how `memes` syntax allows sophisticated analysis designs, how `memes`
 utilities enable deep interrogation of results, and how `memes` flexible data

### 2021-07-13 manuscript.Rmd
fix title & table callouts

@@
 ---
-title: 'Memes: an R interface to the MEME Suite'
+title: 'Memes: a motif analysis environment in R using tools from the MEME Suite'
 author:
@@
 Here we present `memes`, a motif analysis package that provides a seamless
-interface to a selection of popular MEME Suite tools within R. Six MEME Suite tools are contained within the `memes` wrapper (Table 1). `memes` uses base R and Bioconductor
+interface to a selection of popular MEME Suite tools within R. Six MEME Suite tools are contained within the `memes` wrapper (**Table 1**). `memes` uses base R and Bioconductor
 data types for data input and output, facilitating better integration with
@@
 representative plot of these data is shown in **Figure \@ref(fig:tomtom-example)**, while
-the full results can be viewed in **Table 1**. This analysis revealed
+the full results can be viewed in **Table 2**. This analysis revealed
 that the motifs associated with decreasing sites are highly similar to the E93
 motif. Conversely, motifs found in increasing sites match the transcription
-factors Br, L(3)neo38, Gl, and Lola (**Table 1**). These data support
+factors Br, L(3)neo38, Gl, and Lola (**Table 2**). These data support
 the hypothesis that E93 binding to its motif may play a larger role in chromatin
